---
title: "lec15.Rmd"
author: "Adam Kapelner"
date: "April 23, 2020"
output: html_document
---

# Using Probability Estimation to do Classification

First repeat quickly (a) load the adult data (b) do a training / test split and (c) build the logisitc model.

```{r}
pacman::p_load_gh("coatless/ucidata")
data(adult)
adult = na.omit(adult) #kill any observations with missingness

set.seed(1)
train_size = 5000
train_indices = sample(1 : nrow(adult), train_size)
adult_train = adult[train_indices, ]
y_train = adult_train$income
X_train = adult_train
X_train$income = NULL

test_size = 5000
test_indices = sample(setdiff(1 : nrow(adult), train_indices), test_size)
adult_test = adult[test_indices, ]
y_test = adult_test$income
X_test = adult_test
X_test$income = NULL

logistic_mod = glm(income ~ ., adult_train, family = "binomial")
p_hats_train = predict(logistic_mod, adult_train, type = "response")
p_hats_test = predict(logistic_mod, adult_test, type = "response")
```

Let's establish a rule: if the probability estimate is greater than or equal to 50%, let's classify the observation as positive, otherwise 0.

```{r}
y_hats_train = factor(ifelse(p_hats_train >= 0.5, ">50K", "<=50K"))
```

How did this "classifier" do in-sample?

```{r}
mean(y_hats_train != y_train)
table(y_train, y_hats_train)
```

Let's see the same thing oos:

```{r}
y_hats_test = factor(ifelse(p_hats_test >= 0.5, ">50K", "<=50K"))
mean(y_hats_test != y_test)
oos_conf_table = table(y_test, y_hats_test)
oos_conf_table
```

A tad bit worse. Here are estimates of the future performance for each class:

```{r}
n = sum(oos_conf_table)
fp = oos_conf_table[1, 2]
fn = oos_conf_table[2, 1]
tp = oos_conf_table[2, 2]
tn = oos_conf_table[1, 1]
num_pred_pos = sum(oos_conf_table[, 2])
num_pred_neg = sum(oos_conf_table[, 1])
num_pos = sum(oos_conf_table[2, ])
num_neg = sum(oos_conf_table[1, ])
precision = tp / num_pred_pos
cat("precision", round(precision * 100, 2), "%\n")
recall = tp / num_pos
cat("recall", round(recall * 100, 2), "%\n")
false_discovery_rate = 1 - precision
cat("false_discovery_rate", round(false_discovery_rate * 100, 2), "%\n")
false_omission_rate = fn / num_pred_neg
cat("false_omission_rate", round(false_omission_rate * 100, 2), "%\n")
```

This whole classifier hinged on the decision of 50%! What if we change it??

# Asymmetric Cost Classifiers

Let's establish a *new* rule: if the probability estimate is greater than or equal to 90%, let's classify the observation as positive, otherwise 0.

```{r}
y_hats_test = factor(ifelse(p_hats_test >= 0.9, ">50K", "<=50K"))
mean(y_hats_test != y_test)
oos_conf_table = table(y_test, y_hats_test)
oos_conf_table
```

Of course the misclassification error went up! But now look at the confusion table! The second column represents all $\hat{y} = 1$ and there's not too many of them! Why? You've made it *much* harder to classify something as positive. Here's the new additional performance metrics now:

```{r}
n = sum(oos_conf_table)
fp = oos_conf_table[1, 2]
fn = oos_conf_table[2, 1]
tp = oos_conf_table[2, 2]
tn = oos_conf_table[1, 1]
num_pred_pos = sum(oos_conf_table[, 2])
num_pred_neg = sum(oos_conf_table[, 1])
num_pos = sum(oos_conf_table[2, ])
num_neg = sum(oos_conf_table[1, ])
precision = tp / num_pred_pos
cat("precision", round(precision * 100, 2), "%\n")
recall = tp / num_pos
cat("recall", round(recall * 100, 2), "%\n")
false_discovery_rate = 1 - precision
cat("false_discovery_rate", round(false_discovery_rate * 100, 2), "%\n")
false_omission_rate = fn / num_pred_neg
cat("false_omission_rate", round(false_omission_rate * 100, 2), "%\n")
```

We don't make many false discoveries but we make a lot of false omissions! It's a tradeoff...

# Receiver-Operator Curve Plot

The entire classifier is indexed by that indicator function probability threshold which creates the classification decision. Why not see look at the entire range of possible classification models. We do this with a function. We will go through it slowly and explain each piece:

```{r}
#' Computes performance metrics for a binary probabilistic classifer
#'
#' Each row of the result will represent one of the many models and its elements record the performance of that model so we can (1) pick a "best" model at the end and (2) overall understand the performance of the probability estimates a la the Brier scores, etc.
#'
#' @param p_hats  The probability estimates for n predictions
#' @param y_true  The true observed responses
#' @param res     The resolution to use for the grid of threshold values (defaults to 1e-3)
#'
#' @return        The matrix of all performance results
compute_metrics_prob_classifier = function(p_hats, y_true, res = 0.001){
  #we first make the grid of all prob thresholds
  p_thresholds = seq(0 + res, 1 - res, by = res) #values of 0 or 1 are trivial
  
  #now we create a matrix which will house all of our results
  performance_metrics = matrix(NA, nrow = length(p_thresholds), ncol = 12)
  colnames(performance_metrics) = c(
    "p_th",
    "TN",
    "FP",
    "FN",
    "TP",
    "miscl_err",
    "precision",
    "recall",
    "FDR",
    "FPR",
    "FOR",
    "miss_rate"
  )
  
  #now we iterate through each p_th and calculate all metrics about the classifier and save
  n = length(y_true)
  for (i in 1 : length(p_thresholds)){
    p_th = p_thresholds[i]
    y_hats = factor(ifelse(p_hats >= p_th, ">50K", "<=50K"))
    confusion_table = table(
      factor(y_true, levels = c("<=50K", ">50K")),
      factor(y_hats, levels = c("<=50K", ">50K"))
    )
      
    fp = confusion_table[1, 2]
    fn = confusion_table[2, 1]
    tp = confusion_table[2, 2]
    tn = confusion_table[1, 1]
    npp = sum(confusion_table[, 2])
    npn = sum(confusion_table[, 1])
    np = sum(confusion_table[2, ])
    nn = sum(confusion_table[1, ])
  
    performance_metrics[i, ] = c(
      p_th,
      tn,
      fp,
      fn,
      tp,
      (fp + fn) / n,
      tp / npp, #precision
      tp / np,  #recall
      fp / npp, #false discovery rate (FDR)
      fp / nn,  #false positive rate (FPR)
      fn / npn, #false omission rate (FOR)
      fn / np   #miss rate
    )
  }
  
  #finally return the matrix
  performance_metrics
}
```

Now let's generate performance results for the in-sample data:

```{r}
pacman::p_load(data.table, magrittr)
performance_metrics_in_sample = compute_metrics_prob_classifier(p_hats_train, y_train) %>% data.table
performance_metrics_in_sample
```

Now let's plot the ROC curve

```{r}
pacman::p_load(ggplot2)
ggplot(performance_metrics_in_sample) +
  geom_line(aes(x = FPR, y = recall)) +
  geom_abline(intercept = 0, slope = 1, col = "orange") + 
  coord_fixed() + xlim(0, 1) + ylim(0, 1)
```

Now calculate the area under the curve (AUC) which is used to evaluate the probabilistic classifier (just like the Brier score) using a trapezoid area function. 

```{r}
pacman::p_load(pracma)
-trapz(performance_metrics_in_sample$FPR, performance_metrics_in_sample$recall)
```

This is not bad at all!

Note that I should add $<0, 0>$ and $<1, 1>$ as points before this is done but I didn't...

How do we do out of sample?


```{r}
performance_metrics_oos = compute_metrics_prob_classifier(p_hats_test, y_test) %>% data.table
performance_metrics_oos
```

And graph the ROC:


```{r}
#first we do our own melting of two data frames together to make it long format
performance_metrics_in_and_oos = rbind(
    cbind(performance_metrics_in_sample, data.table(sample = "in")),
    cbind(performance_metrics_oos, data.table(sample = "out"))
)
ggplot(performance_metrics_in_and_oos) +
  geom_line(aes(x = FPR, y = recall, col = sample)) +
  geom_abline(intercept = 0, slope = 1, col = "orange") + 
  coord_fixed() + xlim(0, 1) + ylim(0, 1)
```


```{r}
-trapz(performance_metrics_oos$FPR, performance_metrics_oos$recall)
```


Not bad at all - only a tad worse! In the real world it's usually a lot worse. We are lucky we have n = 5,000 in both a train and test set.

# Detection Error Tradeoff curve

```{r}
ggplot(performance_metrics_in_and_oos) +
  geom_line(aes(x = FDR, y = miss_rate, col = sample)) +
  coord_fixed() + xlim(0, 1) + ylim(0, 1)
```

