---
title: "Practice Lecture 1 MATH 342W Queens College"
author: "Professor Adam Kapelner"
date: "February 4, 2021"
---

You can get started with this cheatsheet:
https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf

Acknowledgement: I would like to thank Professor Andreas Buja at Wharton as I cribbed lots of his notes. Here's the sources:
http://stat.wharton.upenn.edu/~buja/STAT-470-503-770/CHAPTERS/
http://stat.wharton.upenn.edu/~buja/STAT-961/

as well as Yonatan Rosenblatt of Ben Gurion University:
http://www.john-ros.com/Rcourse

## Basic R

What is R?

* An open-source, free standard for statistical and numerical programming
* interpreted, scripted and high level (like Python and not like e.g. C++)
* Has a tremendous developer community and industry community
* Very good for "one-shot" scripting - very common in data science
* New stats algorithms and methodologies appear here first!
* I joke that I'm an R developer for a living
* History: came from Bell Labs. C --> S --> R

What is RStudio? An IDE. It is not the only IDE for R, but it is the most popular at the time of this writing. It consists of the file window (what you're reading) and the console window (below) and a few other convenient windows which will be explained later.

File types:

* The standard R script has a .R extension
* This file if you note has an .Rmd extension meeaning "R Markdown". Markdown is a way of typestting simple documents with basic code directives. You can read up on this by yourself. "R Markdown" is a document type that combines R code within markdown documents where the R code is in "chunks".

Let's see some R! Since R is interpreted, you can run commands directly into its command line by using ctrl-enter line-by-line or ctrl-shift-enter for the entire "chunk":

```{r}
5
6
```

This is not the most exciting command. The above is called a "chunk" and it tells the R Markdown compiler that there is R code inside those tick marks.

Here's some R code commented:

```{r}
5 #This will produce the number 5
```

Everything beyond the "#" is a comment as is ignored during interpretation.

Let's start by using R as a calculator

```{r}
5 + 5
5 - 5
5 / 5
5 * 5
5^5 #or 5 ** 5 (not common)
5 / 3
100 * (1 + 0.03)^5
5 %% 3
5 * 5 %% 3
(5 * 5) %% 3 #note order of operations - use parentheses
```

You can have multiple statements per line. The semicolon ends a statement like C++ but it is omitted usually because if the line ends it is implied. 

```{r}
5 + 5; 5 - 5
```

Multiple statements per line is not generally done unless you are making a point.

You can also extend expressions for many lines if you have a dangling operator at the end of the line e.g. this plus sign:

```{r}
5+5+5
```
Avoid this unless the line gets so long it's difficult to read. Note the "+" at the end of the line so the interpreter knows to keep looking to the next line.

Let's take a look at how to represent numbers using scientific notation

```{R}
3,125 #commas are illegal and note that RStudio finds basic errors for us
3.125e3 #scientific notation
3.125E3
```


Every language has a means of invoking functions. Here are some included famous math functions. R uses parentheses to invoke functions. The quantities within the parentheses are known as "parameters" or "arguments" to the function.

```{r}
sqrt(2)
log(2)
log(2, 2)
log(2, base = 2) #This is preferred: specify the argument by name after the first argument.
abs(-1)
floor(1.2)
ceiling(1.2)
round(1.2345)
round(1.2345, 2)
log10(1000)
sin(pi) #pi is a hard-wired numerical constant, the only one in the language (I think)
#note the numerical error
acos(1)
exp(1)
```

We can control the number of digits via

```{r}
options(digits = 3)
3.123456789E-6
options(digits = 7) #the default
3.123456789E-6
```


There are lots of options. Look at the help file for a function by using the question mark:

```{r}
?options
```

Let's talk about basic data types:

```{r}
TRUE #logical
FALSE
T
F
class(TRUE)
1 #numeric, think of double type in C++ or Java
class(1)
1L
class(1L) #what's the L doing? It's forcing a certain type
```

There are more formal ways to "force" AKA "cast":

```{r}
class(as.integer(1))
class(as.numeric(1L)) #what's the L doing? It's forcing a certain type
1.L #what's this? Not an error but a warning
1e3 #scientific notation convenience
1E3
class(1E3)
"A"
class("A")
as.integer(TRUE)
as.numeric(TRUE)
as.integer(FALSE)
as.numeric(FALSE)
as.logical(0)
as.logical(1)
as.logical(0.1) #anything else besides zero...
as.character(1)
as.character(0.1)
as.logical("a") #more on this "character" later
as.logical("FALSE") #more on this "character" later
as.logical("T") #more on this "character" later
```

All of the above are actually 1-dim vectors. We will explore this later.

To create vectors, we use the "c" function which is short for "combine". This function takes any number of arguments

```{r}
c(1, 1, 2, 3, 5, 8)
```

We can also use shorthand "ladder" notation:

```{r}
1 : 3
1 : 100
-3 : 3
3 : -3
-(1 : 5)
```

The "seq" function (short for "sequence") is key for this too:

```{r}
seq(1, 9) #short for 1 : 9
seq(1, 9, by = 2)
seq(1, 9, length.out = 3)
seq(1, 9, length.out = 4)
```

and so is the "rep" function (short for "replicate"):

```{r}
rep(1, 10)
rep(c(1, 2, 3), 10)
rep(seq(1, 2, by = 0.1), 10)
```

Every language has a means to assign a piece of data to variables:

```{r}
x = 1 #I use this... only because I like using one less character and I like my code to read like other popular languages
x
x <- 1 #but this is actually preferred according to the the R spec
1 -> x #even those this is legal, please don't do this!!
x
T = "my string"
x_the_number = 1 #underscored variables preferred
xTheNumber = 1 #camel case is a different style
x.the.number = 1 #some people like this but I don't because it is confusing with other languages e.g. Java
```

And a means to see what variables have been assigned and remove these variables:

```{r}
ls() #who's out there
rm(x)
x
rm(x_the_number, xTheNumber, x.the.number)
rm(list = ls()) #clear the workspace completely!
```

Printing them out is important:

```{r}
print(x) #gives you the output numbered by the number of outputs
cat(x) #just prints and it is preferred
cat(x); cat(x) #it prints without a line break
cat(x, "\n") #you can have many arguments to cat and it will concatenate them
cat("x is", x, "\n"); cat("x is", x, "\n")
```
There is a means to use logical comparisons:

```{r}
x = 1
x == 1
x == 0
x != 1
x != 0
x > 1
x < 1
x >= 1
x <= 1
x = TRUE; y = FALSE
x & y #AND
x | y #OR
```

Previously, we used vector types without even knowing. Let's talk about some simple operations on vectors:

```{r}
x = seq(1, 100) 
length(x) #how long is this vector
head(x) #the first elements
tail(x) #the last elements
head(x, 11) #the first 11 elements
tail(x, 13) #the last 13 elements
x[1] #square brackets are the "getter" - it gets the value at that index. R has one-based indexing while C, C++, Python, Java, have zero-based indexing. Get used to it. SORRY!!!
x[1] = 1984 #square brackets also serve as the "setter" - it sets the value at that index
x[1]
x[6]
x[length(x)] #unfortunately no "-1" last index convenience...
x[c(2, 77)]
x[2 : 77]
x[seq(1, 100, by = 2)] #odd numbers
x[-1] #everything but 1
x[-(1 : 10)] #everything but 1 - 10
x[-seq(2, 100, by = 2)] #another way to see odd numbers
x = c(rep(1, 10), rep(2, 10), rep(3, 10))
unique(x) #only unique elements (in order in which they appear)
x[1 : 10] = 10 : 1 # you can set in vectors as well
x[1 : 10]
```

and comparisons on vectors

```{r}
x = 1 : 5
y = 1 : 5
x == y
all.equal(x, y)
```

Most operations in R are vector-operations and these are preferred because they are optimized:

```{r}
x = 1 : 5
x + 2
x * 2
x^2
```
A note about logical vectors:

```{r}
x = c(TRUE, TRUE)
y = c(TRUE, FALSE)
x = c(FALSE, FALSE)
!x
x & y
x | y
x && y #vector AND - confusing - recommend not to use this
x || y #vector OR - confusing - recommend not to use this
as.numeric(x)
sum(x) #does the numeric conversion for you
prod(x) #does the numeric conversion for you
any(x) #convenient function
all(x)
#sometimes a useful function
xor(TRUE, FALSE)
xor(TRUE, TRUE)
xor(FALSE, FALSE)
```

Sampling is very important:

```{r}
x = 1 : 100
sample(x, 3)
sample(x, 101)
sample(x, 3, replace = TRUE)
sort(sample(x, 101, replace = TRUE))
sample(x) #default is length of the vector i.e. just shuffles
```

There are many "illegal" computation values in R: NaN, Inf, -Inf,

```{r}
1 / 0 #unlike C, Java -- no error... handles this natively
1 / 0 == Inf #this works
0 / 0 #unlike C, Java -- no error... handles this natively
x = 0 / 0
x == NaN #beware!!
is.nan(x)
-1 / 0
1 + 1 / 0
1 + Inf
1 / Inf
1 / (-Inf)
1 / NaN
log(0)
log(-1)
sqrt(-1)
```

There are a couple more "illegalish" values that are non-computational: NA, NULL, empty vectors

```{r}
NA #this is the value of missing
x = NA
x == NA #beware!!
is.na(x)
x = NULL #special reserved primitive for data that is "undefined"
x == NULL #strange... and beware
is.null(x)
#let's look at that strange thing
x = 1 : 3
x[1]
x[0] #a vector with zero elements -- a "null vector" of sorts
length(x[0])
c() #remarkably... not what we expect...
x[NA]
x[NaN]
x[Inf]
x[4]
```

Now let's look at data types again. These types are the data types we learned in class.

```{r}
x = c(1, 2, 3, 4, 5)
class(x)
x = seq(1, 5)
class(x) #integer and numeric are approximately the same for purposes of this class
x = sample(rep(c(0, 1), 50))
x #binary variable?
x = c("a", "b", "c", "d")
x
class(x)
x = rep(x, 5)
x
class(x)
x = factor(x)
x
?factor
levels = c("low", "medium", "high")
x_char = sample(rep(levels, 10))
x = factor(x_char)
x #nominal categorical variable
x = factor(x_char, levels = levels)
x
x = factor(x_char, levels = levels, ordered = TRUE)
x #ordinal categorical variable
as.numeric(x) #coerce this variable to a number... result makes sense
```


Data which is of class factor can be used in mathematical models that we build just like numeric. It will, by default be reduced to binary variables.

## Realizations from r.v.'s

Let's now do some probability a la Math 241. Let's realize a few iid random variable models. Let's say $X_1, ..., X_{10} \iid$ ...

* binomial
* geometric
* poisson
* standard uniform --- all numbers between 0 and 1 equally likely
* exponential
* normal with mean 5 and standard deviation 2

```{r}
n = 10 #good programming practice to declare shared data

x = rbinom(n, size = 5, prob = 0.1)
dbinom(1, size = 5, p = 0.1)
pbinom(3, size = 5, p = 0.1)
qbinom(.75, size = 3, p = 0.1)
x #returns as a vector
?rbinom
x = rgeom(n, prob = 0.1)
x
?rgeom
x = rpois(n, lambda = 2)
x
?rpois
x = runif(n)
x
?runif
x = rexp(n, rate = 1)
x
?rexp
x = rnorm(n, mean = 5, sd = 2)
x
?rnorm
```

Randomness is an illustion in a computer. This is a whole topic in and of itself. But for our purposes, imagine random numbers being generated from a long but fixed list of numbers and the starting point is random. But you can make the randomness deterministic by setting the starting point. Then it's just the same numbers in the same order! This starting point is called a "seed" and it's an integer. 

```{r}
set.seed(1984)
rnorm(10)
set.seed(1984) #reset to that point in the list
rnorm(10) #same exact numbers!!!
set.seed(1984.1)
rnorm(10) #same - needs to be a different integer
set.seed(-1984)
rnorm(10) #different
set.seed(1983)
rnorm(10) #different
get.seed()
```

When writing programs that depend on randomness, you should set the seed at the top so the program is deterministic. Then it is easy to find bugs. When you switch the program to production, you just comment ou the set seed line. I will be doing this fairly often during demos for the rest of the semester.

Sorting is also a key skill in R:

```{r}
n = 50
x = rbinom(n, size = 20, prob = 0.2)
x
sort(x)
sort(x, decreasing = TRUE)
```

## Matrices

Now that we finished vectors, let's learn about the matrix object. Let's use the dimension of the learning problem in class.

```{r}
n = 100 #number of historical objects: the people
p = 3 #number of features about each

random_junk = round(runif(n * p), 2)
random_junk
X = matrix(random_junk, nrow = n, ncol = p)
X
X[1, 1] #the square brackets are still the getter but since there's two dimensions there's two indices
X[1, ] #first row
class(X[1, ]) #note: not matrix anymore
X[1, , drop = FALSE]
class(X[1, , drop = FALSE]) #still matrix... this matters sometimes and we will see an example of it
X[, 1] #first column
class(X[, 1]) #note: not matrix anymore
X[, 1, drop = FALSE]
class(X[, 1, drop = FALSE])
#other ways of subsetting matrices
X[1 : 4, ]
X[, 2 : 3]
X[, c(1, 3)]
X[seq(1, n, by = 2), ]
X[1, 1] = 1984 #the squarae brackets are still the setter
X[1, 1]
```

Sometimes handy by not recommended is getting / setting with only one index.

```{r}
head(X)
X[1] #first entry i.e. the 1,1 entry
X[2] #first entry i.e. the 2,1 entry (column major indexing, see https://cran.r-project.org/web/packages/reticulate/vignettes/arrays.html)
X[n * p] #last entry i.e. n,p entry
X[n * p + 1] #beyond last entry returns missing / illegal
X[-1] #returns everything but the first entry and it automatically vectorizes
X[1 : 100] #returns the first 100 i.e. the first column
X[X > 0.5] = NA #replace all values no matter what row / col they exist in
X
```


A few handy functions for numerical matrices:

```{r}
X = matrix(random_junk, nrow = n, ncol = p) #reset
rowSums(X)
colSums(X)
rowMeans(X)
colMeans(X)
```

## Programming! Conditional statements, loops, functions

We will now learn how to program in R.

* Conditional Statements (and ways to print to screen too).

```{r}
x = 3
if (x > 0){
  cat("The value of x is", x, "\n") #concatenate and print to screen
} else {
  print("The value of x is"); print(x) #each gets its own line
}

if (x > 0){
  ##
} 
else { ##NOTE: ELSE MUST GO ON SAME LINE AS THE IF'S CLOSE BRACE!!! This is not Java or C++!
  ##
}

x = -5
if (x > 0){
  cat("The value of x is", x, "\n") #concatenate and print to screen
} else if (x < -3) { ##NOTE: ELSE MUST GO ON SAME LINE AS THE IF'S CLOSE BRACE!!!
  print(paste("The value of x is", x, collapse = " ")) #paste with collapse is like a string join
} else {
  print("The value of x is"); print(x) #each gets its own line
}
```

There is a one line convenience function called ifelse:

```{r}
x = -10 : 10
b = -10 : 10
a = ifelse(x > 0, b, -1)
a
x = -5 : 5
a = ifelse(x > 0, 1, -1)
a
```



There is also a switch in R. It is _not_ an official keyword (despite RStudio's endorsement), but it is useful for non-vector character expressions:

```{r}
ch = "four"
# ch = c("one", "two")
switch(ch,
   "one" = 1,
   "two" = 2,
   "three" = 3,
   -99
)
```


* Looping

A for loop contains three ingredients: (a) a looping variable joined with keyword "in" to a (b) vector or list (which we will discuss later) and executes over an (c) expression recommended to be enclosed in curly braces:

```{r}
for (i in 1 : 10){
  j = i
  print(j)
}
```

Note that both the looping variable i still exists and the variable within the looped expression j still exists:

```{r}
i
j
```

Be careful of scoping in R. You may want to "clean up your namespace" whenever you can.

```{r}
for (i in 1 : 10){
  j = i
  print(j)
}
rm(i, j)
```

There is no need to save the values of the temp variables i, j going forward. They would vanish in C++ / Java when the for loop ended as it goes out of scope.

You can iterate over a vector as well Python-style or Java-generic-style:

```{r}
vec = seq(1, 10)
for (i in vec){
  print(i)
}
```

And the vector does not have to be numeric:

```{r}
vec = c("a", "b", "c")
for (ch in vec){
  print(ch)
}
```

There is also the while loop but not a "do-while" loop

```{r}
x = 1
while (x <= 10){
  print(x)
  x = x + 1 #note: there is no "++" incrementing in R
}
```

This can also be done using a break:

```{r}
x = 1
while (TRUE){
  print(x)
  x = x + 1 #note: there is no "++" incrementing in R
  if (x > 10){
    break
  }
}
```

The "repeat" keyword is the same as the while-true. It is slightly more esoteric but still good to know:

```{r}
x = 1
repeat {
  print(x)
  x = x + 1 #note: there is no "++" incrementing in R
  if (x > 10){
    break
  }
}
```

The final control keyword tells the loop to break the current iteration but continue to the next:

```{r}
xs = rep(NA, 10)
xs[3] = 3
xs[7] = 7
xs

tot = 0
for (x in xs){
  if (is.na(x)){
    next #move to next element in the list in the for loop
  }
  tot = tot + x
}
tot
```

Note: "real" R programmers avoid loops since they are slow. Instead they use the "apply" function (but not "sapply" or "lapply" which are not faster). We will cover these in the future. The latest trend is to use in-line C++ code (we probably won't get to this).

## More data types

Before we get back to modeling, it is worth knowing a couple more data structures in R. These are not "data [science] types", these are "[computer science] data types" that are used frequently in data science applications. 

The first are "lists" which are "ordered hashmaps" or "ordered dictionaries" or "hash tables". if You don't know what this is, you should read about this online as it should have been covered in a intro to CS class.

```{r}
dict = list()
dict$a = "first"
dict$b = "second"
dict$c = "third"
dict
length(dict)
names(dict) #keys
dict_unlisted = unlist(dict) #values
dict_unlisted
class(dict_unlisted) #i.e. a vector
#three ways to access values by key / ordered location
dict$a
dict[["a"]]
dict[[1]] #value of first entered key
class(dict[[1]])
dict[[1 : 2]] #bombs
#now let's try to access a value for a non-existent key / ordered location
dict$q
dict[["q"]]
dict[[4]] #bombs
#convenient means to subset the list
dict[1]
class(dict[1])
dict[1 : 2]
dict[1 : 4] #this is the reason this type of access is not recommended
dict = list("first", "second", "third") #no key => value... what happens?
dict #default keys are the numbers 1, 2, ...
dict[[1]]
dict = list("a" = "first", "b" = "second", "c" = "third") #key => value
dict
dict = list(a = "first", b = "second", c = "third") #key => value
dict
```

Lists conveniently allow all sorts of data types (as values only).

```{r}
varied_dict = list()
varied_dict$a = "first"
varied_dict$b = 2
varied_dict$c = 1 : 7
varied_dict$d = matrix(NA, nrow = 2, ncol = 2)
varied_dict[["some function"]] = function(x){x^2} #this key is not recommended
varied_dict
varied_dict$`some function` #note the tick marks (sometimes seen) needed due to spaces in key name
length(varied_dict)
names(varied_dict)
```

They have lots of uses in data science applications. We will likely see them in class and if not, you'll definitely see them in the real world. Note that data.frame objects are implemented as lists as well as many other common R objects.

Unfortunately, list can only truly accept characters as keys. If you really need more flexibility here, we will need a library (coming soon).


The second is arrays i.e. are multidimensional vectors

```{r}
x = array(1 : 5, 5)
x
X = array(1 : 25, dim = c(5, 5))
X
X = array(1 : 125, dim = c(5, 5, 5))
X
X[1, , ]
X[, 1, ]
X[, , 1]
X[1, 1, 1]
```

These can be associative arrays too and operate like a hash of vectors across arbitrary dimensions:

```{r}
X = array(1 : 125, 
          dim = c(5, 5, 5),
          dimnames = list(
            c("A", "B", "C", "D", "E"),
            c("I", "II", "III", "IV", "V"),
            c("blue", "red", "green", "yellow", "orange")
          ))
X
X["A", , ]
X[, "III", ]
X[, , "orange"]
X["C", , "orange"]
X["C", "IV", "orange"]
```


* Functions

```{r}
my_function = function(x){
  x
}
##You may be used to:
# int my_function(int x){
#  //something
#  return x;
# }
```

* Functions are objects in R. This is why you are actually assigning the function to an object.
* We don't need to declare type since R is not "statically typed" (higher level languages usually are not statically typed). Objects can be coerced into different types on the fly (R is "dynamically typed").
* No need for a "return" statement as the last line is the data that is returned. It is considered bad style to use "return" in R.

Let's make sure this works:

```{r}
my_function(3)
my_function("asd")
my_function(x = 3) #you can specify that 3 is the value for argument "x"
my_function(y = 3) #illegal argument
(function(x){x + 1})(3) #anonymous function
```

R is somewhat user friendly as it allows for default argument values, making those arguments optional to specify when calling the function:

```{r}
my_function = function(x = 1, y = 2, z = 3, p = 4, q = 5, r = 6){
  (x + y + z) / (p + q + r)
}
my_function() #default execution
my_function(p = 0) #one optional argument specified, others defaulted
my_function(y = -2, q = 0) #two optional arguments specified, others defaulted

my_function = function(x = 1, y, z = 3, p = 4, q = 5, r = 6){
  (x + y + z) / (p + q + r)
}
my_function() #no dice
my_function(1, 0) #required argument specified
my_function(0, q = 7) #required argument specified and one optional argument
rm(my_function) #can be deleted since it's an object
```

There are also common functional programming functions. 

* Reduce uses a binary function to successively combine the elements of a given vector and a possibly given initial value. 
* Filter extracts the elements of a vector for which a predicate (logical) function gives true. 
* Find and Position give the first or last such element and its position in the vector, respectively. 
* Map applies a function to the corresponding elements of given vectors. 

If you like this, there are many packages that extend this and organize it nicely e.g. `purrr` (we will get to packages next class).

```{r}
x = c(1, 2, 3, 4, 5)
Reduce(sum, x)
Filter(function(x){x <= 3}, x)
Find(function(x){x <= 3}, x)
Map(function(x){x + 100}, x) #what happened here?? Map will return a list (for flexibility)
```

## First modeling exercise

Before we model, let's fabricate the training data! Let's try to build a data matrix similar to the one in the class example. Let's imagine $n = 100$ and $x_1$ is salary, $x_2$ is a dummy variable for missing loan payment in their credit history, $x_3$ is a ordinal variable for crime type coded 0, 1, 2 or 3.

We can "make up" a dataset using the sampling we just learned.

```{r}
n = 100 #number of historical objects: the people
p = 3 #number of features about each

X = matrix(NA, nrow = n, ncol = p)
X
```

Some more useful matrix functions:

```{r}
nrow(X)
ncol(X)
dim(X)
length(X) #countless bugs!
c(X)
```

Why should we fill up this matrix with NA's? No technical reason; it is done for a practical reason. Every value is currently "missing" until it's filled in i.e. it will let you know if you didn't fill any of the values.

We can also name rows and columns. Each row is a historical person and each column is a feature about that person.

```{r}
colnames(X) = c(
  "salary", 
  "has_past_unpaid_loan", 
  "past_crime_severity"
)
colnames(X) #setter and a getter
fake_first_names = c(
  "Sophia", "Emma", "Olivia", "Ava", "Mia", "Isabella", "Riley", 
  "Aria", "Zoe", "Charlotte", "Lily", "Layla", "Amelia", "Emily", 
  "Madelyn", "Aubrey", "Adalyn", "Madison", "Chloe", "Harper", 
  "Abigail", "Aaliyah", "Avery", "Evelyn", "Kaylee", "Ella", "Ellie", 
  "Scarlett", "Arianna", "Hailey", "Nora", "Addison", "Brooklyn", 
  "Hannah", "Mila", "Leah", "Elizabeth", "Sarah", "Eliana", "Mackenzie", 
  "Peyton", "Maria", "Grace", "Adeline", "Elena", "Anna", "Victoria", 
  "Camilla", "Lillian", "Natalie", "Jackson", "Aiden", "Lucas", 
  "Liam", "Noah", "Ethan", "Mason", "Caden", "Oliver", "Elijah", 
  "Grayson", "Jacob", "Michael", "Benjamin", "Carter", "James", 
  "Jayden", "Logan", "Alexander", "Caleb", "Ryan", "Luke", "Daniel", 
  "Jack", "William", "Owen", "Gabriel", "Matthew", "Connor", "Jayce", 
  "Isaac", "Sebastian", "Henry", "Muhammad", "Cameron", "Wyatt", 
  "Dylan", "Nathan", "Nicholas", "Julian", "Eli", "Levi", "Isaiah", 
  "Landon", "David", "Christian", "Andrew", "Brayden", "John", 
  "Lincoln"
)
rownames(X) = fake_first_names
rownames(X) #setter and getter
X
```

Let's pretend "salary" is normally distributed with mean \$50,000 and standard error \$20,000. Let's make up some data

```{r}
X[, 1] = round(rnorm(n, 50000, 20000))
X
#another way to set this feature:
X[, "salary"] = round(rnorm(n, 50000, 20000))
X
```

A quick sidebar about vectors within matrices with row or column names:

```{r}
salaries = X[, 1] 
salaries #it's a vector with names
names(salaries) #access to its names
names(salaries)[3] = "Adam"
salaries
salaries[3]
salaries["Adam"]
sort(salaries)
#how do we sort salaries by name?
salaries[order(names(salaries))]
?order #it's like sort, but it returns indices
rm(salaries)
```

Are the salary values independent? Yes, or at least we assume so... Hopefully we will get to models in this semester where they are not independent.

We will eventually do visualization, but first let's take a look at a summary of this data:

```{r}
summary(X[, "salary"])
```

There are other base functions to know:

```{r}
mean(X[, "salary"]) #mean should be "average"!!
sum(X[, "salary"]) / length(X[, "salary"])
var(X[, "salary"])
sd(X[, "salary"])
median(X[, "salary"])
min(X[, "salary"])
max(X[, "salary"])
IQR(X[, "salary"])
```

There is also the convenient quantile and inverse quantile function

```{r}
quantile(X[, "salary"], probs = 0.5)
quantile(X[, "salary"], probs = c(.1, .9))
inverse_quantile_obj = ecdf(X[, "salary"]) #the "empirical" CDF
inverse_quantile_obj(50000)
inverse_quantile_obj(70000)
inverse_quantile_obj(0)
inverse_quantile_obj(-10000)
inverse_quantile_obj(200000)
```


Let's pretend "has_past_unpaid_loan" is benoulli distributed with probability 0.2

```{r}
X[, "has_past_unpaid_loan"] = rbinom(n, size = 1, prob = 0.2)
X
```

Is this a reasonable fabrication of this dataset? No... since salary and not paying back a loan are dependent r.v.'s. But... we will ignore this now.

It would be nice to see a summary of values. Would median and mean be appropriate here? No. For categorical variables, you should "table" them:

```{r}
table(X[, "has_past_unpaid_loan"])
```


Also, 50\% of people have no crime, 40\% have an infraction, 8\% a misdimeanor and 2\% a felony. Let's try to add this to the matrix. We first need to simulate this. Here's how:

```{r}
X[, "past_crime_severity"] = sample(
  c("no crime", "infraction", "misdimeanor", "felony"),
  size = n,
  replace = TRUE,
  prob = c(.50, .40, .08, .02)
)
X
```

Oh no - what happened?? Our matrix went all characters... The matrix type cannot handle numeric and categorical variables simultaneously! It would be nice to keep factor or character information in a matrix but this is not the spec.

Enter the key data type, the "data.frame" - this is the object that is used for modeling in the R ecosystem. It is essentially an upgraded matrix.

```{r}
X = data.frame(
  salary = round(rnorm(n, 50000, 20000)),
  has_past_unpaid_loan = rbinom(n, size = 1, prob = 0.2),
  past_crime_severity = sample(
    c("no crime", "infraction", "misdimeanor", "felony"),
    size = n,
    replace = TRUE,
    prob = c(.50, .40, .08, .02)
  )
)
rownames(X) = fake_first_names
X
```

RStudio gives us a nicer rendering of the information. You can open it up in a separate tab via:

```{r}
View(X)
```

and you can view summaries of each feature and data type of each feature via

```{r}
summary(X)
str(X)
```

Again, summary doesn't work for "has_past_unpaid_loan". We should convert it to factor and try again. Note the "$" operator which is now valid for data.frame objects.

```{r}
X$has_past_unpaid_loan = factor(X$has_past_unpaid_loan, labels = c("Never", ">=1"))
head(X) #make sure that worked
summary(X) #much better now!
str(X)
```

Now that we have two categorical variables, we can do a "cross tab":

```{r}
table(X$has_past_unpaid_loan)
table(X$past_crime_severity)
table(X$has_past_unpaid_loan, X$past_crime_severity) / 100
#to avoid needing the "X$" over and over, use the convenience "with"
with(X,
  table(has_past_unpaid_loan, past_crime_severity)
)
```

In our training set D, we are missing one final variable, the response! Let's add it and say that 90\% of people are creditworthy i.e. they paid back their loan:

```{r}
X$paid_back_loan = factor(rbinom(n, size = 1, prob = 0.9), labels = c("No", "Yes"))
head(X) #make sure that worked
summary(X) #much better now!
```

Conceptually - why does this make no sense at all??? y is independent of X --- what happens then? No function f can ever have any predictive / explanatory power! This is just a silly example to show you the data types. We will work with real data soon. Don't worry.

