---
title: "Practice Lecture 1 MATH 342W Queens College"
author: "Professor Adam Kapelner"
date: "February 1, 2021"
---

You can get started with this cheatsheet:
https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf

Acknowledgement: I would like to thank Professor Andreas Buja at Wharton as I cribbed lots of his notes. Here's the sources:
http://stat.wharton.upenn.edu/~buja/STAT-470-503-770/CHAPTERS/
http://stat.wharton.upenn.edu/~buja/STAT-961/

as well as Yonatan Rosenblatt of Ben Gurion University:
http://www.john-ros.com/Rcourse

## Basic R

What is R?

* An open-source, free standard for statistical and numerical programming
* interpreted, scripted and high level (like Python and not like e.g. C++)
* Has a tremendous developer community and industry community
* Very good for "one-shot" scripting - very common in data science
* New stats algorithms and methodologies appear here first!
* I joke that I'm an R developer for a living
* History: came from Bell Labs. C --> S --> R

What is RStudio? An IDE. It is not the only IDE for R, but it is the most popular at the time of this writing. It consists of the file window (what you're reading) and the console window (below) and a few other convenient windows which will be explained later.

File types:

* The standard R script has a .R extension
* This file if you note has an .Rmd extension meeaning "R Markdown". Markdown is a way of typestting simple documents with basic code directives. You can read up on this by yourself. "R Markdown" is a document type that combines R code within markdown documents where the R code is in "chunks".

Let's see some R! Since R is interpreted, you can run commands directly into its command line by using ctrl-enter line-by-line or ctrl-shift-enter for the entire "chunk":

```{r}
5
6
```

This is not the most exciting command. The above is called a "chunk" and it tells the R Markdown compiler that there is R code inside those tick marks.

Here's some R code commented:

```{r}
5 #This will produce the number 5
```

Everything beyond the "#" is a comment as is ignored during interpretation.

Let's start by using R as a calculator

```{r}
5 + 5
5 - 5
5 / 5
5 * 5
5^5 #or 5 ** 5 (not common)
5 / 3
100 * (1 + 0.03)^5
5 %% 3
5 * 5 %% 3
(5 * 5) %% 3 #note order of operations - use parentheses
```

You can have multiple statements per line. The semicolon ends a statement like C++ but it is omitted usually because if the line ends it is implied. 

```{r}
5 + 5; 5 - 5
```

Multiple statements per line is not generally done unless you are making a point.

You can also extend expressions for many lines if you have a dangling operator at the end of the line e.g. this plus sign:

```{r}
5+5+5
```
Avoid this unless the line gets so long it's difficult to read. Note the "+" at the end of the line so the interpreter knows to keep looking to the next line.

Let's take a look at how to represent numbers using scientific notation

```{R}
3,125 #commas are illegal and note that RStudio finds basic errors for us
3.125e3 #scientific notation
3.125E3
```


Every language has a means of invoking functions. Here are some included famous math functions. R uses parentheses to invoke functions. The quantities within the parentheses are known as "parameters" or "arguments" to the function.

```{r}
sqrt(2)
log(2)
log(2, 2)
log(2, base = 2) #This is preferred: specify the argument by name after the first argument.
abs(-1)
floor(1.2)
ceiling(1.2)
round(1.2345)
round(1.2345, 2)
log10(1000)
sin(pi) #pi is a hard-wired numerical constant, the only one in the language (I think)
#note the numerical error
acos(1)
exp(1)
```

We can control the number of digits via

```{r}
options(digits = 3)
3.123456789E-6
options(digits = 7) #the default
3.123456789E-6
```


There are lots of options. Look at the help file for a function by using the question mark:

```{r}
?options
```

Let's talk about basic data types:

```{r}
TRUE #logical
FALSE
T
F
class(TRUE)
1 #numeric, think of double type in C++ or Java
class(1)
1L
class(1L) #what's the L doing? It's forcing a certain type
```

There are more formal ways to "force" AKA "cast":

```{r}
class(as.integer(1))
class(as.numeric(1L)) #what's the L doing? It's forcing a certain type
1.L #what's this? Not an error but a warning
1e3 #scientific notation convenience
1E3
class(1E3)
"A"
class("A")
as.integer(TRUE)
as.numeric(TRUE)
as.integer(FALSE)
as.numeric(FALSE)
as.logical(0)
as.logical(1)
as.logical(0.1) #anything else besides zero...
as.character(1)
as.character(0.1)
as.logical("a") #more on this "character" later
as.logical("FALSE") #more on this "character" later
as.logical("T") #more on this "character" later
```

All of the above are actually 1-dim vectors. We will explore this later.

To create vectors, we use the "c" function which is short for "combine". This function takes any number of arguments

```{r}
c(1, 1, 2, 3, 5, 8)
```

We can also use shorthand "ladder" notation:

```{r}
1 : 3
1 : 100
-3 : 3
3 : -3
-(1 : 5)
```

The "seq" function (short for "sequence") is key for this too:

```{r}
seq(1, 9) #short for 1 : 9
seq(1, 9, by = 2)
seq(1, 9, length.out = 3)
seq(1, 9, length.out = 4)
```

and so is the "rep" function (short for "replicate"):

```{r}
rep(1, 10)
rep(c(1, 2, 3), 10)
rep(seq(1, 2, by = 0.1), 10)
```

Every language has a means to assign a piece of data to variables:

```{r}
x = 1 #I use this... only because I like using one less character and I like my code to read like other popular languages
x
x <- 1 #but this is actually preferred according to the the R spec
1 -> x #even those this is legal, please don't do this!!
x
T = "my string"
x_the_number = 1 #underscored variables preferred
xTheNumber = 1 #camel case is a different style
x.the.number = 1 #some people like this but I don't because it is confusing with other languages e.g. Java
```

And a means to see what variables have been assigned and remove these variables:

```{r}
ls() #who's out there
rm(x)
x
rm(x_the_number, xTheNumber, x.the.number)
rm(list = ls()) #clear the workspace completely!
```

Printing them out is important:

```{r}
print(x) #gives you the output numbered by the number of outputs
cat(x) #just prints and it is preferred
cat(x); cat(x) #it prints without a line break
cat(x, "\n") #you can have many arguments to cat and it will concatenate them
cat("x is", x, "\n"); cat("x is", x, "\n")
```
There is a means to use logical comparisons:

```{r}
x = 1
x == 1
x == 0
x != 1
x != 0
x > 1
x < 1
x >= 1
x <= 1
x = TRUE; y = FALSE
x & y #AND
x | y #OR
```

Previously, we used vector types without even knowing. Let's talk about some simple operations on vectors:

```{r}
x = seq(1, 100) 
length(x) #how long is this vector
head(x) #the first elements
tail(x) #the last elements
head(x, 11) #the first 11 elements
tail(x, 13) #the last 13 elements
x[1] #square brackets are the "getter" - it gets the value at that index. R has one-based indexing while C, C++, Python, Java, have zero-based indexing. Get used to it. SORRY!!!
x[1] = 1984 #square brackets also serve as the "setter" - it sets the value at that index
x[1]
x[6]
x[length(x)] #unfortunately no "-1" last index convenience...
x[c(2, 77)]
x[2 : 77]
x[seq(1, 100, by = 2)] #odd numbers
x[-1] #everything but 1
x[-(1 : 10)] #everything but 1 - 10
x[-seq(2, 100, by = 2)] #another way to see odd numbers
x = c(rep(1, 10), rep(2, 10), rep(3, 10))
unique(x) #only unique elements (in order in which they appear)
x[1 : 10] = 10 : 1 # you can set in vectors as well
x[1 : 10]
```

and comparisons on vectors

```{r}
x = 1 : 5
y = 1 : 5
x == y
all.equal(x, y)
```

Most operations in R are vector-operations and these are preferred because they are optimized:

```{r}
x = 1 : 5
x + 2
x * 2
x^2
```
A note about logical vectors:

```{r}
x = c(TRUE, TRUE)
y = c(TRUE, FALSE)
x = c(FALSE, FALSE)
!x
x & y
x | y
x && y #vector AND - confusing - recommend not to use this
x || y #vector OR - confusing - recommend not to use this
as.numeric(x)
sum(x) #does the numeric conversion for you
prod(x) #does the numeric conversion for you
any(x) #convenient function
all(x)
#sometimes a useful function
xor(TRUE, FALSE)
xor(TRUE, TRUE)
xor(FALSE, FALSE)
```

Sampling is very important:

```{r}
x = 1 : 100
sample(x, 3)
sample(x, 101)
sample(x, 3, replace = TRUE)
sort(sample(x, 101, replace = TRUE))
sample(x) #default is length of the vector i.e. just shuffles
```

There are many "illegal" computation values in R: NaN, Inf, -Inf,

```{r}
1 / 0 #unlike C, Java -- no error... handles this natively
1 / 0 == Inf #this works
0 / 0 #unlike C, Java -- no error... handles this natively
x = 0 / 0
x == NaN #beware!!
is.nan(x)
-1 / 0
1 + 1 / 0
1 + Inf
1 / Inf
1 / (-Inf)
1 / NaN
log(0)
log(-1)
sqrt(-1)
```

There are a couple more "illegalish" values that are non-computational: NA, NULL, empty vectors

```{r}
NA #this is the value of missing
x = NA
x == NA #beware!!
is.na(x)
x = NULL #special reserved primitive for data that is "undefined"
x == NULL #strange... and beware
is.null(x)
#let's look at that strange thing
x = 1 : 3
x[1]
x[0] #a vector with zero elements -- a "null vector" of sorts
length(x[0])
c() #remarkably... not what we expect...
x[NA]
x[NaN]
x[Inf]
x[4]
```

Now let's look at data types again. These types are the data types we learned in class.

```{r}
x = c(1, 2, 3, 4, 5)
class(x)
x = seq(1, 5)
class(x) #integer and numeric are approximately the same for purposes of this class
x = sample(rep(c(0, 1), 50))
x #binary variable?
x = c("a", "b", "c", "d")
x
class(x)
x = rep(x, 5)
x
class(x)
x = factor(x)
x
?factor
levels = c("low", "medium", "high")
x_char = sample(rep(levels, 10))
x = factor(x_char)
x #nominal categorical variable
x = factor(x_char, levels = levels)
x
x = factor(x_char, levels = levels, ordered = TRUE)
x #ordinal categorical variable
as.numeric(x) #coerce this variable to a number... result makes sense
```


Data which is of class factor can be used in mathematical models that we build just like numeric. It will, by default be reduced to binary variables.



