---
title: "Practice Lecture 9 MATH 390.4 Queens College"
author: "Professor Adam Kapelner"
date: "March 1, 2021"
---

## Null Model

```{r}
y = MASS::Boston$medv
mean(y)
```

We can use `lm`.

```{r}
mod = lm(y ~ 1)
summary(mod)
mod$fitted.values
```


Let's do a simple example of projection. Let's project $y$ onto the intercept column, the column of all 1's. What do you think will happen?

```{r}
ones = rep(1, length(y))
H = ones %*% t(ones) / sum(ones^2)
H[1 : 5, 1 : 5]
#in fact
unique(c(H))
```

The whole matrix is just one single value for each element! What is this value? It's 1 / 506 where 506 is $n$. So what's going to happen?

```{r}
y_proj_ones = H %*% y
head(y_proj_ones)
```

Projection onto the space of all ones makes the null model ($g = \bar{y}$). It's the same as the model of response = intercept + error, i.e. $y = \mu_y + \epsilon$. The OLS intercept estimate is clearly $\bar{y}$.


## OLS using categorical predictors

Note that historically this is called "Analysis of Variance" or "ANOVA" for short. But there is no difference to the computer, it still crunches the same matrices.

Let's get the cars data again:

```{r}
cars = MASS::Cars93
str(cars)
```

Let's try to model `Type`, a factor with 6 levels.

```{r}
table(cars$Type)
```

What will $\hay{y}$ look like? Should be the $\bar{y}$'s for each level. What is $p$? 6. Let' see:

```{r}
anova_mod = lm(Price ~ Type, cars)
coef(anova_mod)
summary(anova_mod)$r.squared
```

What's going on here? This one categorical variable got blown up into 5 features! How to interpret?? First need reference category. Which one is missing? `Compact`. So what is prediction for the compact type? The intercept. What is prediction of Large type? Intercept + Large, etc. How does this regression look. Let's see the model matrix:

```{r}
Xmm = model.matrix(Price ~ Type, cars)
head(Xmm, 20)
```

The predictor `Type` got "dummified". (Remember we spoke about this in lecture 2)? There are now 5 dummy variables each representing one of the levels and the reference level is omitted because it is accounted for in the intercept. Let's make sure this is exactly what's going on.

```{r}
y = cars$Price
Xt = t(Xmm) 
XtX = Xt %*% Xmm
XtXinv = solve(XtX)
b = XtXinv %*% Xt %*% y
b
yhat = Xmm %*% b
e = y - yhat
Rsq = (var(y) - var(e)) / var(y)
Rsq
```

And of course the coefficients and $R^2$ are identical to the output from `lm`.

If we want to do a more "pure ANOVA", we can get rid of the intercept and see the $\bar{y}$'s immediately. This is handled in R's formula designation by adding a zero:

```{r}
anova_mod = lm(Price ~ 0 + Type, cars)
coef(anova_mod)
```

Is this correct?

```{r}
mean(cars$Price[cars$Type == "Compact"])
mean(cars$Price[cars$Type == "Large"])
mean(cars$Price[cars$Type == "Midsize"])
mean(cars$Price[cars$Type == "Small"])
mean(cars$Price[cars$Type == "Sporty"])
mean(cars$Price[cars$Type == "Van"])
```

What does $R^2$ look like?

```{r}
summary(anova_mod)$r.squared
```

Remember this from last time? What happened? The $R^2$ calculation in `lm` is not accurate without the intercept. Keep this in mind. 

What does the design matrx (model matrix) look like? we can use the `model.matrix` function to generate the columns of $X$ from the data frame. The argument is the formula we wish to generate the model matrix for. Since model matrices don't require

```{r}
Xmm = model.matrix(~ 0 + Type, cars)
head(Xmm, 20)
```

Very similar. 

Regressions without an intercept are not recommended. Here's why. What if we were doing two factors? I want a linear model with both Type and Airbags:

```{r}
table(cars$AirBags)
```

Airags is another nominal categorical variable, this time with three levels.

We invoke the model as follows.

```{r}
anova_mod = lm(Price ~ Type + AirBags, cars)
coef(anova_mod)
summary(anova_mod)$r.squared
```

What are interpretations now? What is the "reference level"? It's actually two levels in one: Type = compact and Airbags = Driver \& Passenger. 

A deeper question: can we read off Type = Midsize and AirBags = none? No... this is a modeling "enhancement" we will discuss in a few lectures from now.

If we model it without an intercept,


```{r}
anova_mod = lm(Price ~ 0 + Type + AirBags, cars)
coef(anova_mod)
```

we only get $\bar{y}$'s for the first factor predictor crossed with the reference category of the second. So above `TypeCompact` refers to the average of Type = Compact and Airbags = Driver \& Passenger.

Now let's create a linear model using one categorical predictor and one continuous predictor. The combination is called for historical reasons "Analysis of Covariance" or "ANCOVA" for short.

Let's use `Type` and `Horsepower`:

```{r}
ancova_mod = lm(Price ~ Type + Horsepower, cars)
coef(ancova_mod)
summary(ancova_mod)$r.squared
```

Interpretation of estimated coefficients? Why did $R^2$ increase? (We will be explaining this in detail in the next unit).

What's going on the design / model matrix?

```{r}
head(model.matrix(Price ~ Type + Horsepower, cars))
```

Same as model matrix with just `Type`. Since `Horsepower` is continuous, it doesn't get dummified to more features.

What if we went back to the `Type` regression, left out the intercept, dummified and added the intercept back in?

```{r}
Xmm = model.matrix(Price ~ 0 + Type, cars)
Xmm = cbind(1, Xmm)
head(Xmm)
```

Are the columns linearly independent? No ... so when we try to get the hat matrix,

```{r}
Xmm %*% solve(t(Xmm) %*% Xmm) %*% t(Xmm)
```

You can't invert a non-invertible matrix!!

What does R do when using the linear model function:

```{r}
coef(lm(cars$Price ~ 0 + Xmm))
```

SOMEHOW: it doesn't complain since it handles the non-invertibility (we don't know why...) but we do see that it's busted. Look at the coefficients! One is missing! What is it doing?? It's just arbitrarily dropping one (just like recommended).

## Multivariate linear regression with the Hat Matrix

```{r}
X = as.matrix(cbind(1, MASS::Boston[, 1: 13]))
Xt = t(X)
XtXinv = solve(Xt %*% X)
b = XtXinv %*% t(X) %*% y
b
```

We can compute all predictions:

```{r}
H = X %*% XtXinv %*% t(X)

yhat = H %*% y
head(yhat)
```

Can you tell this is projected onto a 13 dimensionsal space from a 506 dimensional space? Not really... but it is...

Now let's project over and over...

```{r}
head(H %*% H %*% H %*% H %*% H %*% H %*% H %*% H %*% H %*% y)
```

Same thing! Once you project, you're there. That's the idempotency of $H$.

Let's make sure that it really does represent the column space of $X$. Let's try to project different columns of $X$:

```{r}
head(X[, 1, drop = FALSE])
head(H %*% X[, 1, drop = FALSE])

head(X[, 2, drop = FALSE])
head(H %*% X[, 2, drop = FALSE])

head(X[, 3, drop = FALSE])
head(H %*% X[, 3, drop = FALSE]) #why?? Numerical error...

#etc....
```

We can calculate the residuals:

```{r}
I = diag(nrow(X))
e = y - yhat
head(e)
e = (I - H) %*% y
head(e)
```

Let's do that projection over and over onto the complement of the column space of $X$:

```{r}
head((I - H) %*% (I - H) %*% (I - H) %*% (I - H) %*% (I - H) %*% (I - H) %*% y)
```


## Eigendecomposition in R

```{r}
B = array(seq(1, 4), dim = c(2, 2))
B
eigen_decomp = eigen(B)
V = eigen_decomp$vectors
V
v1 = V[, 1, drop = FALSE]
v2 = V[, 2, drop = FALSE]
lambdas = eigen_decomp$values
lambdas
lambda1 = lambdas[1]
lambda2 = lambdas[2]
B %*% v1
lambda1 * v1
B %*% v2 
lambda2 * v2
B %*% v2 == lambda2 * v2 #why not?
#diagonolization
V %*% diag(lambdas) %*% solve(V)
```

# QR Decomposition


Let's go back to the Boston data and regenerate all our quantities:

```{r}
y = MASS::Boston$medv
ybar = mean(y)
SST = sum((y - ybar)^2)

X = as.matrix(cbind(1, MASS::Boston[, 1: 13]))
n = nrow(X)
p_plus_one = ncol(X)
Xt = t(X)
XtXinv = solve(Xt %*% X)
b = XtXinv %*% Xt %*% y
b
yhat = X %*% b
# e = y - yhat
# SSE = sum(e^2)
SSR = sum((yhat - ybar)^2)
SSR
Rsq = SSR / SST
Rsq
```

Now let's do the QR decomposition and see if the projections work.

```{r}
qrX = qr(X)
Q = qr.Q(qrX)
R = qr.R(qrX)
dim(Q)
dim(R)
Matrix::rankMatrix(Q)
Matrix::rankMatrix(R)

sum(Q[, 1]^2) #normalized?
sum(Q[, 2]^2) #normalized?
Q[, 1] %*% Q[, 2] #orthogonal?
Q[, 7] %*% Q[, 13] #orthogonal?

Qt = t(Q)
yhat_via_Q = Q %*% Qt %*% y
head(yhat)
head(yhat_via_Q)
testthat::expect_equal(c(yhat), c(yhat_via_Q)) #needed to vectorize to make dimensions equal
```

Can we get the $b$ vector from the $Q$ matrix?

```{r}
b = Qt %*% y
b
```

Nope - this is not the same! Why not?

Each dimension gives one piece of SSR and thus one piece of R^2 i.e. SSR = SSR_1 + ... + SSR_p

Our definition of SSR removed the ybar i.e. the contribution of the intercept. So we will do so here. That is the first column of $Q$:

```{r}
head(Q)
```

Now we add up all the features besides the intercept

```{r}
partial_SSRs = array(NA, p_plus_one)
for (j in 2 : p_plus_one){
  qj = Q[, j, drop = FALSE]
  yhat_j = qj %*% t(qj) %*% y #the projection onto the jth dimension of Q
  partial_SSRs[j] = sum(yhat_j^2)
}
round(partial_SSRs)
sum(partial_SSRs, na.rm = TRUE)
SST
partial_Rsqs = partial_SSRs / SST
round(partial_Rsqs, 2)
sum(partial_Rsqs, na.rm = TRUE)

round(partial_Rsqs / Rsq, 2)
```

Some dimensions in this subspace matter more than others. We can do approximately the same regression with less than p features. Let's try this:

```{r}
partial_Rsqs_sorted = sort(partial_Rsqs, decreasing = TRUE)
partial_Rsqs_sorted_cumul = cumsum(partial_Rsqs_sorted)
partial_Rsqs_sorted_cumul
#sort Q by Rsq
Qsorted = Q[, order(partial_Rsqs, na.last = FALSE)]
#let's take the first 8
Qreduced = Qsorted[, 6 : 14]
mod = lm(y ~ Qreduced)
summary(mod)
```

Why was the first column of `Qsorted` dropped?


# Correlation zero means orthogonality

Let's generate some fake data. In this example we'll have one predictor which will be orthogonal to the centered response. We enforce the response to be centered by adding a column of 1's:

```{r}
n = 100; p = 2
Q = qr.Q(qr(cbind(1, matrix(rnorm(n * p), nrow = n))))
y = Q[, p + 1]
x = Q[, 2]
```

Let's make sure it's orthogonal:

```{r}
x %*% y
```

If they're orthogonal and y is mean-centered, what is the correlation?

```{r}
cor(x, y)
```

If the correlation is 0, what is $b_1$, the slope? It has to be zero. Thus $b_0$ has to be $bar{x}$. Since x was also orthogonalized to the vector of 1's, it's centered and hence has average = 0. So both intercept and slope are 0:

What is $b$?

```{r}
mod = lm(y ~ x)
coef(mod)
```

What is $R^2$? Since $x$ and $y$ are orthogonal... a projection onto the colspace of $X$ gets annhilated.

```{r}
summary(mod)$r.squared
```

# Random correlations are non-zero

```{r}
n = 100
x = rnorm(n)
x = x - mean(x)
y = rnorm(n)
y = y - mean(y)
```

In this setup, $x$ and $y$ are centered Gaussian random vectors. Are they orthogonal?

```{r}
x %*% y
theta_in_rad = acos(x %*% y / sqrt(sum(x^2) * sum(y^2)))
theta_in_rad * 180 / pi


theta_in_rad = acos(X[, 65] %*% y / sqrt(sum(X[, 65]^2) * sum(y^2)))
theta_in_rad * 180 / pi
```

Nope... what about correlated?

```{r}
cor(x, y)
cor(x, y)^2
```

They *nearly* uncorrelated but they still have some correlation. How is this possible? 

There is "random chance"" AKA "chance capitalization"!

What about the best fitting line?

```{r}
mod = lm(y ~ x)
coef(mod)
```

Slope is about -0.8 which is small but non-zero.

What is $R^2$? Since $x$ and $y$ are nearly orthogonal... a projection onto the colspace of $X$ gets nearly annhilated.

```{r}
summary(mod)$r.squared
```

but not entirely. Lesson learned: random noise can be correlated with the response $y$ and give you the illusion of fit!


# The monotonicity of SSR (or $R^2$) with more features

As p increases, $R^2$ goes up. Here's a nice exercise:

```{r}
n = 100
y = rnorm(n)
Rsqs = array(NA, n)

#we know that Rsq = 0 for the null model (i.e. just regressing on the intercept)
Rsqs[1] = 0

#create a matrix with the correct number of rows but no columns
X = matrix(NA, nrow = n, ncol = 0)
X = cbind(1, X)

#for every new p, tack on a new random continuos predictor:
for (p_plus_one in 2 : n){
  X = cbind(X, rnorm(n))
  Rsqs[p_plus_one] = summary(lm(y ~ X))$r.squared
}
```

Now let's plot it and see what happens:

```{r}
pacman::p_load(ggplot2)
base = ggplot(data.frame(p_plus_one = 1 : n, Rsq = Rsqs))
base + geom_line(aes(x = p_plus_one, y = Rsq))
```

With each additional predictor, what happens to $R^2$?

```{r}
pacman::p_load(latex2exp)
base + geom_line(aes(x = p_plus_one, y = c(0, diff(Rsq)))) + xlab("p + 1") + ylab(TeX("$\\Delta R^2$"))
```

$R^2$ always increases:

```{r}
all(diff(Rsqs) > 0)
```

How can this possibly be?? The $x$'s are not related to $y$ whatsoever!!

Chance capitalization prevails. Each additional predictor picks up another dimension to add to the column space of $X$. Eventually, the projection explains *all* the variance. If $n = p + 1$, that matrix is square and of full rank, hence $\hat{y} = y$ and all residuals $e = 0$ since it is merely solving $n$ linearly independent equations.

So here's an idea. To get a perfect fit, just augment your design matrix with $n - (p + 1)$ random vectors and you get $R^2 = 100\%$!! There must be something wrong with this!!

Even if $p$ is large and $<n$, you are getting a lot of fit for free. So there's something wrong with this too!

This is called overfitting.

## Overfitting

Let's see how overfitting increases generalized estimation error.

```{r}
bbeta = c(1, 2, 3, 4)

#build training data
n = 100
X = cbind(1, rnorm(n), rnorm(n), rnorm(n))
y = X %*% bbeta + rnorm(n, 0, 0.3)
#\amthbb{D}

#build test data
n_star = 100
X_star = cbind(1, rnorm(n_star), rnorm(n_star), rnorm(n_star))
y_star = X_star %*% bbeta + rnorm(n_star, 0, 0.3)

all_betas = matrix(NA, n, n)
all_betas[4, 1 : 4] = coef(lm(y ~ 0 + X))
in_sample_rmse_by_p = array(NA, n)
for (j in 5 : n){
  X = cbind(X, rnorm(n))
  lm_mod = lm(y ~ 0 + X)
  all_betas[j, 1 : j] = coef(lm_mod)
  y_hat = X %*% all_betas[j, 1 : j]
  in_sample_rmse_by_p[j] = sqrt(mean((y - y_hat)^2))
}
plot(1 : n, in_sample_rmse_by_p)


all_betas[4 : n, 1 : 4]
b_error_by_p = rowSums((all_betas[, 1 : 4] - matrix(rep(bbeta, n), nrow = n, byrow = TRUE))^2)
plot(1 : n, b_error_by_p)

#look at out of sample error in the case of only the first four features
oos_rmse_by_p = array(NA, n)
for (j in 4 : n){
  y_hat_star = X_star %*% all_betas[j, 1 : 4]
  oos_rmse_by_p[j] = sqrt(mean((y_star - y_hat_star)^2))
}
plot(1 : n, oos_rmse_by_p)

#look at out of sample error in the case of the random features too
oos_rmse_by_p = array(NA, n)
for (j in 5 : n){
  X_star = cbind(X_star, rnorm(n))
  y_hat_star = X_star %*% all_betas[j, 1 : j]
  oos_rmse_by_p[j] = sqrt(mean((y_star - y_hat_star)^2))
}
plot(1 : n, oos_rmse_by_p, ylim = c(0, 2))
```
