---
title: "Practice Lecture 8 MATH 390.4 Queens College"
author: "Professor Adam Kapelner"
date: "March 20, 2020"
---

#Extrapolation vs Interpolation

Let's take a look at the Galton Data again.

```{r}
pacman::p_load(HistData, ggplot2)
data(Galton)
mod = lm(child ~ parent, Galton)
b_0 = mod$coefficients[1]
b_1 = mod$coefficients[2]
ggplot(Galton, aes(x = parent, y = child)) + 
  geom_point() + 
  geom_jitter() +
  geom_abline(intercept = b_0, slope = b_1, color = "blue", size = 1) +
  xlim(63.5, 72.5) + 
  ylim(63.5, 72.5) +
  coord_equal(ratio = 1)
```

Let's say I want to predict child's height for parents' average height of 70in. All I do is:

```{r}
predict(mod, data.frame(parent = 70))
```

What if I want to predict for a parents' height of 5in. Is there any 12in tall human being? No... it is absurd. But nothing stops you from doing:

```{r}
predict(mod, data.frame(parent = 5))
```

That's [actually possible](https://www.guinnessworldrecords.com/news/2012/2/shortest-man-world-record-its-official!-chandra-bahadur-dangi-is-smallest-adult-of-all-time/).

Look at our linear model from Euclid's perspective:

```{r}
ggplot(Galton, aes(x = parent, y = child)) + 
  geom_point() + 
  geom_jitter() +
  geom_abline(intercept = b_0, slope = b_1, color = "blue", size = 1) +
  xlim(-20, 120) + 
  ylim(-20, 120) +
  coord_equal(ratio = 1)
```

What is a linear model with $p + 1 = 2$. It's just a line. When geometry was first formalised by Euclid in the Elements, he defined a line to have "breadthless length" with a straight line being a line "which lies evenly with the points on itself". By "breadthless" he meant infinite in either direction. There is no mathematical problem with predicting childrens' heights using negative parents' heights e.g.

```{r}
predict(mod, data.frame(parent = -5))
```

But this is absurd. So now we need to talk about a fundamental concept in data science we've been kind of avoiding and one that most people ignore. There are two types of prediction: interpolation and extrapolation. Interpolation is essentially the type of "prediction" we've been talking about this whole class. 

Extrapolation is totally different. It's what happens when you predict outside of the range of the covariate data you've seen in $\mathbb{D}$. Extrapolation is very dangerous - your models only work based on $\mathbb{D}$. Extrapolation is prediction outside of the range you've seen before which means. You better have a good theoretical reason as to why your $\mathbb{H}$ function class will extend outside that range. Because each $\mathbb{H}$ function class will extrapolate very very differently.



What happens during extrapolation? Let's look at the (a) linear model, (b) polynomial model with degree 2 and (c) polynomial with degree 13.

```{r}
degree_2_poly_mod = lm(child ~ poly(parent, 2, raw = TRUE), Galton)
b_poly_2 = coef(degree_2_poly_mod)
degree_13_poly_mod = lm(child ~ poly(parent, 13, raw = TRUE), Galton)
b_poly_13 = coef(degree_13_poly_mod)
b_poly_13[is.na(b_poly_13)] = 0

plot_function_degree_2 = function(x, b){
  b[1] + b[2] * x + b[3] * x^2
}
plot_function_degree_13 = function(x, b){
  b[1] + b[2] * x + b[3] * x^2 + b[4] * x^3 + b[5] * x^4 + b[6] * x^5 + b[7] * x^6 + b[8] * x^7 + b[9] * x^8 + b[10] * x^9  + b[11] * x^10 + b[12] * x^11 + b[13] * x^12 + b[14] * x^13
}

# xymin = 65
# xymax = 71
xymin = 50
xymax = 90
ggplot(Galton, aes(x = parent, y = child)) + 
  geom_point() + 
  geom_jitter() +
  geom_abline(intercept = b_0, slope = b_1, color = "blue") +
  coord_cartesian(xlim = c(xymin, xymax), ylim = c(xymin, xymax)) +
  stat_function(fun = plot_function_degree_2, args = list(b = b_poly_2), col = "red", xlim = c(xymin, xymax)) +
  stat_function(fun = plot_function_degree_13, args = list(b = b_poly_13), col = "green", xlim = c(xymin, xymax))
```

Polynomial models have *TERRIBLE* extrapolation risk - totally unpredictable.

# Overfitting with Polynomials

Can we achieve $R^2 = 100\%$ using polynomial regression? Yes. Here's an example in one dimension. These are called "interpolation polynomials". In one dimension, as long as the $x$ values are distinct, $n$ data point can be fit by a $n - 1$ degree polynomial. Here's an example with a few data points:

```{r}
set.seed(1003)
n = 5
beta_0 = 1
beta_1 = 0
beta_2 = 1
x = runif(n)

h_star_x = beta_0 + beta_1 * x + beta_2 * x^2
y = h_star_x + rnorm(n)
#scatterplot it
df = data.frame(x = x, y = y, h_star_x = h_star_x)
basic = ggplot(df, aes(x, y)) +
  geom_point()
basic
```

Now fit polynomial models:

```{r}
degree_4_poly_mod = lm(y ~ poly(x, 4, raw = TRUE))
b_poly_4 = coef(degree_4_poly_mod)

plot_function_degree_4 = function(x, b){
  b[1] + b[2] * x + b[3] * x^2 + b[4] * x^3 + b[5] * x^4
}

basic + stat_function(fun = plot_function_degree_4, args = list(b = b_poly_4), col = "purple")
```

Perfect fit!

```{r}
summary(degree_4_poly_mod)$r.squared
```

This is the same thing we've seen before! If $n = p + 1$, then the design matrix is square and there is no need to project onto a lower dimensional subspace. To estimate the linear model, one only needs to solve $n$ equations with $n$ unknowns.

My recommendations:
1) Keep polynomial degree low. Preferably 2. Anything past 2 is not interpretable anyway. We didn't talk about "interpretability" of models yet, but you get the idea.
2) Be very careful not to extrapolate: make sure future predictions have the measurements within range of the training data $\mathbb{D}$. Extrapolations are going to be very, very inaccurate. Polynomial regressions I'm sure have gotten data scientists fired before.


# The Grammar of graphics and ggplot

First load the package and the dataset of interest as a dataframe:

```{r}
pacman::p_load(ggplot2, quantreg)
cars = MASS::Cars93 #dataframe
```

ggplot is based on the "Grammar of Graphics", a concept invented by the Statistician / Computer Scientist Leland Wilkinson who worked on SPSS, Tableau and now he works at H20, software that analyzes big data. The reference of interest is [here](http://papers.rgrossman.com/proc-094.pdf). He drew on ideas from John Tukey (one of the great statistician of the previous generation) while he was at Bell Labs, Andreas Buja (one of my professors at Penn) and Jerome Friedman (the professor that taught my data mining course at Stanford - remember the diamonds story?) 

It is a language that allows us to describe the components of a graphic. Previously, graphics were done in one shot and it was clunky. ggplot is a library written by Hadley Wickham based on this concept. Wickham is probably the most famous person in statistical computing today. He has commit rights in R and is one of the architects of RStudio. He calls grammar of graphics "graphical poems". Here are the basic components:

* an underlying data frame
* an "aesthetic" that maps visualization axes in the eventual plot(s) to variables in the data frame
* a "layer" which is composed of
  - a geometric object
  - a statistical transformation
  - a position adjustment
* a "scale" for each aesthetic
* a "coordinate" system for each aesthetic
* optional "facets" (more graphics)
* optional "labels" for the title, axes title, points, etc.

Don't worry - everything has "smart defaults" in Wickham's implementation so you don't have to worry about most things. We will explore some of the features below. Here's a good [cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf).

ggplot is layered where each component is an object. The objects are added to each other since the "+" operator is overloaded to accept these additions. This is nice because each component can be saved and reused. The following initialized the graphics:

```{r}
ggplot(cars)
```

Nothing happened - except the plot window went gray (the smart default). This is already rendering a graphic, but since it hasn't been given any of the required information, it has nothing to display. Next we create an aesthetics indicating a one-way plot (one variable only).

```{r}
ggplot(cars) + 
  aes(Price)
```

Notice how it can understand the variable name as an object name.

Since we've given it an aesthetics object, it now knows which variable is the x axis (default). It already knows the ranges of the variable (a smart default) and a default scale and coordinate system (smart defaults).

Usually this is done in one step by passing the aesthetics object into the ggplot:

```{r}
ggplot(cars, aes(Price))
```

Now we need to pick a layer by specifying a geometry. This is a type of plot. Since the predictor type of price is continuous, let's pick the "histogram" using the `geom_histogram` function:

```{r}
ggplot(cars, aes(Price)) +
  geom_histogram()
```

This can be customized:

```{r}
ggplot(cars, aes(Price)) +
  geom_histogram(binwidth = 1, col = "darkgreen", fill = "blue", alpha = 0.4)
```

Want to save it for your latex?

```{r}
ggsave("plot.png")
system("open plot.png")
ggsave("plot.pdf")
system("open plot.pdf")
```

Here are some other options besides the histogram:

```{r}
ggplot(cars, aes(Price)) +
  geom_dotplot()
ggplot(cars, aes(Price)) +
  geom_area(stat = "bin", binwidth = 2)
ggplot(cars, aes(Price)) +
  geom_freqpoly()
ggplot(cars, aes(Price)) +
  geom_density(fill = "green", alpha = 0.4)

summary(cars)
```


Can we compare price based on different conditions? Yes, we can subset the data and use color and alpha:

```{r}
ggplot(cars, aes(Price)) +
  geom_density(data = subset(cars, Man.trans.avail == "Yes"), col = "grey", fill = "darkgreen", alpha = 0.4) +
  geom_density(data = subset(cars, Man.trans.avail == "No"), col = "grey", fill = "red", alpha = 0.4)
```

Sidebar: why are cars that have manual transmissions available cheaper?

We can look at this also using a histogram of the conditional distributions:

```{r}
ggplot(cars, aes(Price)) +
  geom_histogram(data = subset(cars, Man.trans.avail == "Yes"), binwidth = 1, col = "grey", fill = "darkgreen", alpha = 0.4) +
  geom_histogram(data = subset(cars, Man.trans.avail == "No"), binwidth = 1, col = "grey", fill = "red", alpha = 0.4)
```

What if the variable is not continuous e.g. Cylinders? We can use a bar graph / bar plot.

```{r}
ggplot(cars, aes(Cylinders)) +
  geom_bar()
```

This is essential frequency by level of the categorical variable.

Now let's move on to looking at one variable versus another variable. For example price by engine power:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price))
```

Since we've given it an aesthetics object, it now knows which variable is the x axis and which variable is the y axis. It already knows the ranges of the variables (a smart default) and a default scale and coordinate system (smart defaults).

Just as before, now we need to pick a layer by specifying a geometry. This is a type of plot. Let's pick the "scatterplot" using the `geom_point` function:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point()
```

Now we have a nice scatterplot. This function uses the inherited data, the inherited aesthetics. Since this "geometry" is a "layer", we can pass in options to the layer.

```{r}
base_and_aesthetics = ggplot(cars, aes(x = Horsepower, y = Price))
base_and_aesthetics + geom_point(col = "red", fill = "green", shape = 23, size = 3, alpha = 0.3)
```

Let's handle names of axes, title and ranges:

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics +
  ggtitle("Average Car Price vs. Engine Power", subtitle = "in the Cars93 dataset") +
  ylab("Price (in $1000's)")
base_and_aesthetics_with_titles +
  geom_point() +
  xlim(0, 400) +
  ylim(0, 50)
  
```

Let's transform the variables:

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  scale_x_continuous(trans = "log2")
```

Each unit increase on the x-axis now represent a doubling increase in x (although the whole scale only spans 3 units). But look at how the grid didn't keep up. Let's fix this:

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  scale_x_continuous(trans = "log2", breaks = round(seq(0, max(cars$Horsepower), length.out = 6)))
```

We can do the same to the y axis:

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  scale_y_continuous(trans = "log10")
  scale_x_continuous(trans = "log10", breaks = round(seq(0, max(cars$Horsepower), length.out = 6)))
```

Let's look at some more geometries.

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  geom_smooth()
```

Here, I've added two geometries on the same aesthetic! This attempts to explain the relationship $f(x)$ using smoothing. Let's go for more.

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  geom_smooth() +
  geom_rug()
```

This allows us to also see the marginal distributions of the two variables.

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  geom_smooth() +
  geom_quantile(col = "red") +
  geom_rug()
```

This fits a line and tries to indicate statistical significance of the line. We have *not* covered any statistics in this class yet (ironic!) ... so ignore how the window is generated.

Can we display more than two dimensions? Yes. Let's indicate a third dimension with shape (only works with factors).

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power and Transmission")
base_and_aesthetics_with_titles +
  geom_point(aes(shape = Man.trans.avail)) +
  geom_smooth() +
  geom_rug()
```

Can we display more than three dimensions? Yes. Let's indicate a fourth dimension with color.

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power, Transmission & Drivetrain")
base_and_aesthetics_with_titles +
  geom_point(aes(shape = Man.trans.avail, col = DriveTrain)) +
  geom_smooth() +
  geom_rug()
```

Can we go to a fifth dimension? Maybe?

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power, Transmission & Drivetrain")
base_and_aesthetics_with_titles +
  geom_point(aes(shape = Man.trans.avail, col = DriveTrain, size = Weight), alpha = 0.5) + #size?
  geom_smooth() +
  geom_rug()
```

A seventh? We can use text labels adjacent to the scatterplot's points.

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power, Transmission, Drivetrain,  Weight & #Cylinders")
base_and_aesthetics_with_titles +
  geom_point(aes(shape = Man.trans.avail, col = DriveTrain, alpha = Weight)) + #size?
  geom_text(aes(label = Cylinders), vjust = 1.5, col = "darkgrey", lineheight = 0.3, size = 3) +
  geom_smooth() +
  geom_rug()
```

Getting difficult to see what's going on.

Let's move away from the scatterplot to just density estimation:

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power") #reset the title
base_and_aesthetics_with_titles +
  geom_density2d()
```

Other alternatives:

```{r}
base_and_aesthetics_with_titles +
  geom_bin2d(binwidth = c(8, 3))
pacman::p_load(hexbin)
base_and_aesthetics_with_titles +
  geom_hex()
```

This is like a two-dimensional histogram where the bar / hexagon heights are seen with color.

What if the x-axis is categorical for example Cylinders versus price? Typical is the "box and whiskers" plot:

```{r}
ggplot(cars, aes(x = Cylinders, y = Price)) +
  geom_boxplot()
```

Clear relationship!

How about multiple subplots based on the subsetting we did in the histograms? This is called "faceting". Here are two bivariate visualizations laid horizontally:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point() +
  geom_smooth() +
  facet_grid(. ~ Man.trans.avail)
```

Or alternatively, vertically:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point() +
  geom_smooth() +
  facet_grid(Man.trans.avail ~ .)
```

And we can even double-subset:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point() +
  facet_grid(Man.trans.avail ~ Origin)
```

And we can even triple-subset or more:

```{r}
cars$MedWeight = ifelse(cars$Weight > median(cars$Weight), ">MedWeight", "<MedWeight")
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point() +
  facet_grid(Man.trans.avail ~ Origin + MedWeight, scales = "free")
```

These three varibles seem somewhat independent.

There are other primitives like `geom_abline` which graphs a line and `geom_segment` we will see today. Note that if you want plots rendered within functions or loops you have to explicitly call the `plot` function:

```{r}
for (nsim in 1 : 3){
  graphics_obj = ggplot(data.frame(x = rnorm(1000))) + 
    geom_histogram(aes(x))
  graphics_obj
}
```

versus:

```{r}
for (nsim in 1 : 3){
  graphics_obj = ggplot(data.frame(x = rnorm(1000))) + geom_histogram(aes(x))
  plot(graphics_obj)
}
```


Lastly, ggplot offers lots of nice customization themes:

```{r}
graphics_obj = base_and_aesthetics_with_titles +
  geom_point() +
  geom_smooth() +
  geom_quantile(col = "red") +
  geom_rug()
graphics_obj + theme_bw()
graphics_obj + theme_dark()
graphics_obj + theme_classic()

```

Packages offer even more:

```{r}
pacman::p_load(forcats, lazyeval, ggthemes)
graphics_obj + theme_economist()
graphics_obj + theme_stata()
graphics_obj + theme_tufte()
```

and of course, the whimsical one and only:


```{r}
pacman::p_load(xkcd, extrafont)
download.file("http://simonsoftware.se/other/xkcd.ttf", dest = "xkcd.ttf", mode = "wb")
#MAC
# system("mv xkcd.ttf /Library/Fonts")
# font_import(path = "/Library/Fonts", pattern = "xkcd", prompt = FALSE)
# fonts()
# fonttable()
# loadfonts()
#WINDOWS
font_import(path = ".", pattern = "xkcd", prompt = FALSE)
fonts()
fonttable()

loadfonts(device="win")

graphics_obj + theme_xkcd()
```

# C++ and R

R goes back to 1995 when it was adapted from S (written in 1976 by John Chambers at Bell Labs) with minor modifications. The core of base R is written in C and Fortran. These two languages are the fastest known languages (how to measure "fastest" is a huge debate). Thus, base R is very fast. For instance the `sort` function is as fast as C/Fortran since it immediately calls compiled C/Fortran routines.

However, R code itself that you write is "interpreted" which means it is not compiled until you run it. And it has to compile on-the-fly, making it very slow. Prior to v3.4 (April, 2017) it was even slower since the code wasn't JIT compiled. All this "real CS" stuff you can learn in another class..

One notable place to observe this slowness relative to other languages is in looping. For example:


```{r}
f = function(){
  SIZE = 1e7
  v = array(NA, SIZE)
  for (i in 1 : SIZE){
    v[i] = i
  }
  v
}

v = f()
```

How long does this take?

```{r}
# install.packages("Rcpp")
pacman::p_load_current_gh("hadley/lineprof")
lineprof(f())

system.time({
  f()
})
#on my office computer for SIZE = 10,000,000:
# user  system elapsed 
# 7.13    0.20    7.35 
```

Take a simple function that computes square roots on each element:

```{r}
sqrt_vector = function(v){
  v_new = array(NA, length(v))
  for (i in 1 : length(v)){
    v_new[i] = sqrt(v[i])
  }
  v_new
}

system.time({
  sqrt_vector(v)
})
```

Does the apply function help?

```{r}
system.time({
  apply(v, MARGIN = 1, FUN = sqrt)
})
```

Strange that this takes so long? So it doesn't help... it hurts.

How much faster in C++ should this be?

Enter the `Rcpp` package - a way to compile little bits (or lotta bits) of C++ on the fly.

```{r}
pacman::p_load(Rcpp)
```


Let's write this for loop function to sqrt-ize and compile it and then save it into our namespace to be called like a regular function.

```{r}
cppFunction('
  NumericVector sqrt_vector_cpp(NumericVector v) {
    int n = v.size();
    NumericVector v_new(n);
    for (int i = 0; i < n; i++) { //indices from 0...n-1 not 1...n!
      v_new[i] = sqrt(v[i]);
    }
    return v_new;
  }
')
```

What do these two functions look like?

```{r}
sqrt_vector
sqrt_vector_cpp
```

One shows the R code and then says it is bytecode-compiled (go to an advanced CS class). The other just says we `.Call` some C++ function in a certain address and the argument to be inputted.

What is the gain in runtime?

```{r}
system.time({
  sqrt_vector_cpp(v)
})
```

WOW. 10x!!! Can't beat that with a stick...

Let's do a not-so-contrived example...

Matrix distance... Let's compute the distances of all pairs of rows in a dataset. I will try to code the R as efficiently as possible by using vector subtraction so there is only two for loops. The C++ function will have an additional loop to iterate over the features in the observations.

```{r}
#a subset of the diamonds data
X_diamonds = as.matrix(ggplot2::diamonds[1 : 3000, c("carat", "depth", "table", "x", "y", "z")])

compute_distance_matrix = function(X){
  n = nrow(X)
  D = matrix(NA, n, n)
  for (i_1 in 1 : (n - 1)){
    for (i_2 in (i_1 + 1) : n){
      D[i_1, i_2] = sqrt(sum((X[i_1, ] - X[i_2, ])^2))
    }
  }
  D
}

cppFunction('
  NumericMatrix compute_distance_matrix_cpp(NumericMatrix X) {
    int n = X.nrow();
    int p = X.ncol();
    NumericMatrix D(n, n);
    std::fill(D.begin(), D.end(), NA_REAL);

    for (int i_1 = 0; i_1 < (n - 1); i_1++){
      //Rcout << "computing for row #: " << (i_1 + 1) << "\\n";
      for (int i_2 = i_1 + 1; i_2 < n; i_2++){
        double sqd_diff = 0;
        for (int j = 0; j < p; j++){
          sqd_diff += pow(X(i_1, j) - X(i_2, j), 2); //by default the cmath library in std is loaded
        }
        D(i_1, i_2) = sqrt(sqd_diff); //by default the cmath library in std is loaded
      }
    }
    return D;
  }
')
```

```{r}
system.time({
  D = compute_distance_matrix(X_diamonds)
})
round(D[1 : 5, 1 : 5], 2)
```

Slow...

```{r}
system.time({
  D = compute_distance_matrix_cpp(X_diamonds)
})
round(D[1 : 5, 1 : 5], 2)
```

Absolutely lightning... 100x faster on my laptop than R's runtime.

Writing functions as strings that compile is annoying. It is better to have separate files. For instance...

```{r}
sourceCpp("distance_matrix.cpp")
```

Here are a list of the data structures in Rcpp: https://teuder.github.io/rcpp4everyone_en/070_data_types.html#vector-and-matrix

Another place where C++ pays the rent is recursion. Here is a quicksort implementation in R taken from 

```{r}
quicksort_R <- function(arr) {
  # Pick a number at random.
  mid <- sample(arr, 1)

  # Place-holders for left and right values.
  left <- c()
  right <- c()
  
  # Move all the smaller values to the left, bigger values to the right.
  lapply(arr[arr != mid], function(d) {
    if (d < mid) {
      left <<- c(left, d)
    }
    else {
      right <<- c(right, d)
    }
  })
  
  if (length(left) > 1) {
    left <- quicksort_R(left)
  }
  
  if (length(right) > 1) {
    right <- quicksort_R(right)
  }
  
  # Finally, return the sorted values.
  c(left, mid, right)
}
```

Let's create a random array to test these sorts on:

```{r}
n = 5e5
x = rnorm(n)
```


Let's profile the pure R sort function:

```{r}
system.time({
  x_sorted_pure_R = quicksort_R(x)
})
```

Let's profile R's `sort` function.

```{r}
system.time({
  x_sorted_base_R = sort(x)
})
```

Let's just ensure our method worked...

```{r}
pacman::p_load(testthat)
expect_equal(x_sorted_pure_R, x_sorted_base_R)
```

Basically infinitely faster. Let's make our own C++ implementation.

```{r}
sourceCpp("quicksort.cpp")
```

and profile it:

```{r}
system.time({
  x_sorted_cpp = quicksort_cpp(x)
})
```

Let's just ensure this method worked...

```{r}
pacman::p_load(testthat)
expect_equal(x_sorted_cpp, x_sorted_base_R)
```

Why is our C++ slower than `sort`. Because `sort` is also in C++ or Fortran and it's been likely optimized and reoptimized up to wazoo for decades. Also, Rcpp's data structures may be slower than base R's data structures. There may be some speed lost to translating to `NumericVector` from `double[]` or something like that.

Can you call R from Rcpp? You bet:

```{r}
cppFunction('
  NumericVector rnorm_cpp_R(int n, double mean, double sd){
      // get a pointer to R\'s rnorm() function
      Function f("rnorm");   
  
      // Next code is interpreted as rnorm(n, mean, sd)
      return f(n, Named("sd")=sd, _["mean"]=mean);
  }
')

rnorm_cpp_R(5, 1, .01)
```

A few math functions are implemented for you already:

```{r}
evalCpp('R::qnorm(0.5, 0, 1, 1, 0)')
```

Further, there are many common functions that are already wrapped for you via "Rcpp-sugar" which was the Rcpp's author's attempt to make Rcpp a whole lot easier, see [here](http://dirk.eddelbuettel.com/code/rcpp/Rcpp-sugar.pdf).

```{r}
evalCpp('rnorm(10, 100, 3)')
```

If you want blazing fast linear algebra, check out package `RcppArmadillo` which is a wrapper around Apache's Armadillo (namespace is "arma" in the code), an optimized linear algebra package in C++. Here is an example taken from [here](https://scholar.princeton.edu/sites/default/files/q-aps/files/slides_day4_am.pdf):

```{r}
pacman::p_load(RcppArmadillo, microbenchmark, testthat)

cppFunction('
  arma::mat ols_cpp(arma::mat X, arma::mat y){
    arma::mat Xt = X.t();
    return solve(Xt * X, Xt * y);
  }
', depends = "RcppArmadillo")

n = 500
D = data.frame(int = rep(1, n), x1 = rnorm(n), x2 = rnorm(n), x3 = rnorm(n), y = rnorm(n))
X = as.matrix(D[, 1 : 4])
y = as.matrix(D[, 5])

#does the function work?
expect_equal(as.numeric(ols_cpp(X, y)), as.numeric(solve(t(X) %*% X) %*% t(X) %*% y))

microbenchmark(
  R_via_lm = lm(y ~ 0 + ., data = D),
  R_matrix_multiplication = solve(t(X) %*% X) %*% t(X) %*% y,
  cpp = ols_cpp(X, y),
    times = 100
)
```
About 4x faster than R's optimized linear algebra routines. Supposedly it can go even faster if you enable parallelization within Armadillo.


Here are the places where Rcpp should be used (from https://teuder.github.io/rcpp4everyone_en/010_Rcpp_merit.html)

* Loop operations in which later iterations depend on previous iterations.
* Accessing each elements of a vector/matrix.
* Recurrent function calls within loops.
* Changing the size of vectors dynamically.
* Operations that need advanced data structures and algorithms (we don't do this in this class)
