---
title: "Practice Lecture 10 MATH 390.4 Queens College"
author: "Professor Adam Kapelner"
date: "March 8, 2021"
---

# QR Decomposition

Let's go back to the Boston data and regenerate all our quantities:

```{r}
y = MASS::Boston$medv
ybar = mean(y)
SST = sum((y - ybar)^2)

X = as.matrix(cbind(1, MASS::Boston[, 1: 13]))
n = nrow(X)
p_plus_one = ncol(X)
Xt = t(X)
XtXinv = solve(Xt %*% X)
b = XtXinv %*% Xt %*% y
b
yhat = X %*% b
# e = y - yhat
# SSE = sum(e^2)
SSR = sum((yhat - ybar)^2)
SSR
Rsq = SSR / SST
Rsq
```

Now let's do the QR decomposition and see if the projections work.

```{r}
qrX = qr(X)
Q = qr.Q(qrX)
R = qr.R(qrX)
dim(Q)
dim(R)
Matrix::rankMatrix(Q)
Matrix::rankMatrix(R)

sum(Q[, 1]^2) #normalized?
sum(Q[, 2]^2) #normalized?
Q[, 1] %*% Q[, 2] #orthogonal?
Q[, 7] %*% Q[, 13] #orthogonal?

Qt = t(Q)
yhat_via_Q = Q %*% Qt %*% y
head(yhat)
head(yhat_via_Q)
testthat::expect_equal(c(yhat), c(yhat_via_Q)) #needed to vectorize to make dimensions equal
```

Can we get the $b$ vector from the $Q$ matrix?

```{r}
b = Qt %*% y
b
```

Nope - this is not the same! Why not?

Each dimension gives one piece of SSR and thus one piece of R^2 i.e. SSR = SSR_1 + ... + SSR_p

Our definition of SSR removed the ybar i.e. the contribution of the intercept. So we will do so here. That is the first column of $Q$:

```{r}
head(Q)
```

Now we add up all the features besides the intercept

```{r}
partial_SSRs = array(NA, p_plus_one)
for (j in 2 : p_plus_one){
  qj = Q[, j, drop = FALSE]
  yhat_j = qj %*% t(qj) %*% y #the projection onto the jth dimension of Q
  partial_SSRs[j] = sum(yhat_j^2)
}
round(partial_SSRs)
sum(partial_SSRs, na.rm = TRUE)
SST
partial_Rsqs = partial_SSRs / SST
round(partial_Rsqs, 2)
sum(partial_Rsqs, na.rm = TRUE)

round(partial_Rsqs / Rsq, 2)
```

Some dimensions in this subspace matter more than others. We can do approximately the same regression with less than p features. Let's try this:

```{r}
partial_Rsqs_sorted = sort(partial_Rsqs, decreasing = TRUE)
partial_Rsqs_sorted_cumul = cumsum(partial_Rsqs_sorted)
partial_Rsqs_sorted_cumul
#sort Q by Rsq
Qsorted = Q[, order(partial_Rsqs, na.last = FALSE)]
#let's take the first 8
Qreduced = Qsorted[, 6 : 14]
mod = lm(y ~ Qreduced)
summary(mod)
```

Why was the first column of `Qsorted` dropped?


# Correlation zero means orthogonality

Let's generate some fake data. In this example we'll have one predictor which will be orthogonal to the centered response. We enforce the response to be centered by adding a column of 1's:

```{r}
n = 100; p = 2
Q = qr.Q(qr(cbind(1, matrix(rnorm(n * p), nrow = n))))
y = Q[, p + 1]
x = Q[, 2]
```

Let's make sure it's orthogonal:

```{r}
x %*% y
```

If they're orthogonal and y is mean-centered, what is the correlation?

```{r}
cor(x, y)
```

If the correlation is 0, what is $b_1$, the slope? It has to be zero. Thus $b_0$ has to be $bar{x}$. Since x was also orthogonalized to the vector of 1's, it's centered and hence has average = 0. So both intercept and slope are 0:

What is $b$?

```{r}
mod = lm(y ~ x)
coef(mod)
```

What is $R^2$? Since $x$ and $y$ are orthogonal... a projection onto the colspace of $X$ gets annhilated.

```{r}
summary(mod)$r.squared
```

# Random correlations are non-zero

```{r}
n = 100
x = rnorm(n)
x = x - mean(x)
y = rnorm(n)
y = y - mean(y)
```

In this setup, $x$ and $y$ are centered Gaussian random vectors. Are they orthogonal?

```{r}
x %*% y
theta_in_rad = acos(x %*% y / sqrt(sum(x^2) * sum(y^2)))
theta_in_rad * 180 / pi


theta_in_rad = acos(X[, 65] %*% y / sqrt(sum(X[, 65]^2) * sum(y^2)))
theta_in_rad * 180 / pi
```

Nope... what about correlated?

```{r}
cor(x, y)
cor(x, y)^2
```

They *nearly* uncorrelated but they still have some correlation. How is this possible? 

There is "random chance"" AKA "chance capitalization"!

What about the best fitting line?

```{r}
mod = lm(y ~ x)
coef(mod)
```

Slope is about -0.8 which is small but non-zero.

What is $R^2$? Since $x$ and $y$ are nearly orthogonal... a projection onto the colspace of $X$ gets nearly annhilated.

```{r}
summary(mod)$r.squared
```

but not entirely. Lesson learned: random noise can be correlated with the response $y$ and give you the illusion of fit!


# The monotonicity of SSR (or $R^2$) with more features

As p increases, $R^2$ goes up. Here's a nice exercise:

```{r}
n = 100
y = rnorm(n)
Rsqs = array(NA, n)

#we know that Rsq = 0 for the null model (i.e. just regressing on the intercept)
Rsqs[1] = 0

#create a matrix with the correct number of rows but no columns
X = matrix(NA, nrow = n, ncol = 0)
X = cbind(1, X)

#for every new p, tack on a new random continuos predictor:
for (p_plus_one in 2 : n){
  X = cbind(X, rnorm(n))
  Rsqs[p_plus_one] = summary(lm(y ~ X))$r.squared
}
```

Now let's plot it and see what happens:

```{r}
pacman::p_load(ggplot2)
base = ggplot(data.frame(p_plus_one = 1 : n, Rsq = Rsqs))
base + geom_line(aes(x = p_plus_one, y = Rsq))
```

With each additional predictor, what happens to $R^2$?

```{r}
pacman::p_load(latex2exp)
base + geom_line(aes(x = p_plus_one, y = c(0, diff(Rsq)))) + xlab("p + 1") + ylab(TeX("$\\Delta R^2$"))
```

$R^2$ always increases:

```{r}
all(diff(Rsqs) > 0)
```

How can this possibly be?? The $x$'s are not related to $y$ whatsoever!!

Chance capitalization prevails. Each additional predictor picks up another dimension to add to the column space of $X$. Eventually, the projection explains *all* the variance. If $n = p + 1$, that matrix is square and of full rank, hence $\hat{y} = y$ and all residuals $e = 0$ since it is merely solving $n$ linearly independent equations.

So here's an idea. To get a perfect fit, just augment your design matrix with $n - (p + 1)$ random vectors and you get $R^2 = 100\%$!! There must be something wrong with this!!

Even if $p$ is large and $<n$, you are getting a lot of fit for free. So there's something wrong with this too!

This is called overfitting.

## Overfitting

Let's see how overfitting increases generalized estimation error.

```{r}
bbeta = c(1, 2, 3, 4)

#build training data
n = 100
X = cbind(1, rnorm(n), rnorm(n), rnorm(n))
y = X %*% bbeta + rnorm(n, 0, 0.3)
#\amthbb{D}

#build test data
n_star = 100
X_star = cbind(1, rnorm(n_star), rnorm(n_star), rnorm(n_star))
y_star = X_star %*% bbeta + rnorm(n_star, 0, 0.3)

all_betas = matrix(NA, n, n)
all_betas[4, 1 : 4] = coef(lm(y ~ 0 + X))
in_sample_rmse_by_p = array(NA, n)
for (j in 5 : n){
  X = cbind(X, rnorm(n))
  lm_mod = lm(y ~ 0 + X)
  all_betas[j, 1 : j] = coef(lm_mod)
  y_hat = X %*% all_betas[j, 1 : j]
  in_sample_rmse_by_p[j] = sqrt(mean((y - y_hat)^2))
}
plot(1 : n, in_sample_rmse_by_p)


all_betas[4 : n, 1 : 4]
b_error_by_p = rowSums((all_betas[, 1 : 4] - matrix(rep(bbeta, n), nrow = n, byrow = TRUE))^2)
plot(1 : n, b_error_by_p)

#look at out of sample error in the case of only the first four features
oos_rmse_by_p = array(NA, n)
for (j in 4 : n){
  y_hat_star = X_star %*% all_betas[j, 1 : 4]
  oos_rmse_by_p[j] = sqrt(mean((y_star - y_hat_star)^2))
}
plot(1 : n, oos_rmse_by_p)

#look at out of sample error in the case of the random features too
oos_rmse_by_p = array(NA, n)
for (j in 5 : n){
  X_star = cbind(X_star, rnorm(n))
  y_hat_star = X_star %*% all_betas[j, 1 : j]
  oos_rmse_by_p[j] = sqrt(mean((y_star - y_hat_star)^2))
}
plot(1 : n, oos_rmse_by_p, ylim = c(0, 2))
```



## Eigendecomposition in R

```{r}
B = array(seq(1, 4), dim = c(2, 2))
B
eigen_decomp = eigen(B)
V = eigen_decomp$vectors
V
v1 = V[, 1, drop = FALSE]
v2 = V[, 2, drop = FALSE]
lambdas = eigen_decomp$values
lambdas
lambda1 = lambdas[1]
lambda2 = lambdas[2]
B %*% v1
lambda1 * v1
B %*% v2 
lambda2 * v2
B %*% v2 == lambda2 * v2 #why not?
#diagonolization
V %*% diag(lambdas) %*% solve(V)
```