---
title: "Practice Lecture 10 MATH 342W Queens College"
author: "Professor Adam Kapelner"
date: "March 3, 2021"
---

## Multivariate linear regression with the Hat Matrix

First let's do the null model to examine what the null hat matrix looks like. In this exercise, we will see that $g_0 = \bar{y}$ is really the OLS solution in the case of no features, only an intercept i.e. $b_0 = \bar{y}$.

```{r}
y = MASS::Boston$medv
mean(y)
```

We can use `lm`.

```{r}
mod = lm(y ~ 1)
coef(mod)
mod$fitted.values
```


Let's do a simple example of projection. Let's project $y$ onto the intercept column, the column of all 1's. What do you think will happen?

```{r}
ones = rep(1, length(y))
H = ones %*% t(ones) / sum(ones^2)
H[1 : 5, 1 : 5]
#in fact
unique(c(H))
1 / 506
```

The whole matrix is just one single value for each element! What is this value? It's 1 / 506 where 506 is $n$. So what's going to happen?

```{r}
y_proj_ones = H %*% y
head(y_proj_ones)
```

Projection onto the space of all ones makes the null model ($g = \bar{y}$). It's the same as the model of response = intercept + error, i.e. $y = \mu_y + \epsilon$. The OLS intercept estimate is clearly $\bar{y}$.

```{r}
y = MASS::Boston[, 14]
X = as.matrix(cbind(1, MASS::Boston[, 1 : 13]))
Xt = t(X)
XtXinv = solve(Xt %*% X)
b = XtXinv %*% t(X) %*% y
b
```

We can compute all predictions:

```{r}
H = X %*% XtXinv %*% t(X)
dim(h)
yhat = H %*% y
head(yhat)
```

Can you tell this is projected onto a 13 dimensionsal space from a 506 dimensional space? Not really... but it is...

Now let's project over and over...

```{r}
head(H %*% H %*% H %*% H %*% H %*% H %*% H %*% H %*% H %*% y)
```

Same thing! Once you project, you're there. That's the idempotency of $H$.

Let's make sure that it really does represent the column space of $X$. Let's try to project different columns of $X$:

```{r}
head(X[, 1, drop = FALSE])
head(H %*% X[, 1, drop = FALSE])

head(X[, 2, drop = FALSE])
head(H %*% X[, 2, drop = FALSE])

head(X[, 3, drop = FALSE])
head(H %*% X[, 3, drop = FALSE]) #why?? Numerical error...

head(X[, 1, drop = FALSE] * 3 + X[, 2, drop = FALSE] * 17)
head(H %*% (X[, 1, drop = FALSE] * 3 + X[, 2, drop = FALSE] * 17))

#etc....
```

We can calculate the residuals:

```{r}
e = y - yhat
head(e)
I = diag(nrow(X))
e = (I - H) %*% y
head(e)
```

Let's do that projection over and over onto the complement of the column space of $X$:

```{r}
head((I - H) %*% (I - H) %*% (I - H) %*% (I - H) %*% (I - H) %*% (I - H) %*% y)
```

