---
title: "Practice Lecture 14 MATH 342W Queens College"
author: "Professor Adam Kapelner"
date: "March 17, 2021"
---

# Orthogonal vs raw polynomials

I forgot to talk about this last class...

Why is orthonormal polynomial the default? You can argue that doing a QR decomposition on the polynomial expansion and employing Q in the design matrix will change b thereby making b uninterpretable! So why use orthonormal polynomials? Here's why:

```{r}
n = 1000
set.seed(1984)
X = data.frame(x = c(runif(n / 2, 0, 1e-2), runif(n / 2, 0, 1e6)))
d = 10
num_digits = 8
Xmm_orth = model.matrix(~ 0 + poly(x, d), X)
colnames(Xmm_orth)[1 : d] = 1 : d
Xmm_raw = model.matrix(~ 0 + poly(x, d, raw = TRUE), X)
colnames(Xmm_raw)[1 : d] = 1 : d
```

Let's look at the design matrix for small values of x:

```{r}
head(as.matrix(X))
round(head(Xmm_orth), num_digits)
round(head(Xmm_raw), num_digits)
```

You get numerical underflow almost immediately when using the raw polynomial computations (you get it by degree 4). And thus you can't even get the OLS estimates:


```{r}
y = rnorm(n)
solve(t(Xmm_raw) %*% Xmm_raw) %*% t(Xmm_raw) %*% y
```

Let's look at the design matrix for large values of x:

```{r}
tail(as.matrix(X))
round(tail(Xmm_orth), num_digits)
round(tail(Xmm_raw), num_digits)
```

You get numerical overflow in the design matrix (but it will happen later). But the second you start to use the design matrix with 10^59's inside...

```{r}
solve(t(Xmm_raw[800 : 1000, ]) %*% Xmm_raw[800 : 1000, ]) %*% t(Xmm_raw[800 : 1000, ]) %*% y
```

As opposed to

```{r}
solve(t(Xmm_orth) %*% Xmm_orth) %*% t(Xmm_orth) %*% y
```

No problem at all!!!

So that's the reason: numerical stability. But if you need interpretability, you need raw polynomials. But if you're interpreting the model, how do you even interpret beyond degree 2???

# Log transformations

We will be examining the diamonds dataset. Let's take a moment to read about our data. In real life, you will take more than just a moment to learn about the data. And more than just a moment to clean the data and do sanity checks. As these steps are more of the grunt work of data science (and are learned on the fly), I won't really cover them formally.

```{r}
rm(list = ls())
pacman::p_load(ggplot2) #this loads the diamonds data set too
?diamonds
dim(diamonds)
```

That's a huge $n$. So, let's expect things to take a bit longer when processing.

A natural increasing relationship will likely be found between weight and price. Let's see it visually:

```{r}
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point()
```


How good does a best guess linear relationship do?

```{r}
mod = lm(price ~ carat, diamonds)
b = coef(mod)
b
summary(mod)$r.squared
summary(mod)$sigma
```

What does the intercept say about extrapolation?

Let's see the best fit line $g(x)$ visually:

```{r}
ggplot(diamonds, aes(x = carat, y = price)) + geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green") +ylim(0, 50000)
```

Maybe the relationship between weight and price is not linear - but exponential? E.g. if the weight of a diamond doubles, maybe the price quadruples? Or linear increases in weight yield percentage increases in price. This seems plausible. 

Besides the plausibility of the functional form, there is an agnostic reason to employ log y as the predictive target. Let's first examine the univariate data!


```{r}
skimr::skim(diamonds$price)
```

Very large standard error and very long tail which can be seed more clearly here:

```{r}
ggplot(diamonds) + geom_histogram(aes(price), binwidth = 200)
```

Let's take a look at the distribution after logging:

```{r}
ggplot(diamonds) + geom_histogram(aes(x = log(price)), binwidth = 0.01)
```

Some strange artifacts appear. Why the gap? Why is it "cut" sharply at a maximum? These are questions to ask the one who collected the data. But let's get back to the log story...

Popular wisdom says logging this type of highly skewed-right distribution would possibly make the model "more linear in x". Put another way, it would be easier to "catch" (predict) the long tail since it won't be a long tail anymore after you log-transform. It would also prevent observations with large y's becoming "leverage points" i.e. points that unduly influence the model and thereby warp its ability to predict the average observation. If we have time later in the semester, we can learn about leverage points, but not now. In Econ 382 you learn more reasons for why you should use log the response, but those reasons are relevant for inference so we won't discuss them here.

Let's give the model with ln(y) a whirl. Maybe we'll even learn something about diamonds. The way to create such a model is to simply fit an OLS model to log y. This is called a log-linear model. Since this is a pretty standard thing to do so R's formula notation has it built-in as follows:

```{r}
log_linear_mod = lm(log(price) ~ carat, diamonds)
b = coef(log_linear_mod)
b
```

Let's see what this looks like.

```{r}
ggplot(diamonds, aes(x = carat, y = log(price))) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green") + ylim(6, 12)
```

It looks very bad if carat is large. That means our little theory about carats getting larger yielding multiples of price doesn't correspond to reality.

How did we do?

```{r}
summary(log_linear_mod)$r.squared
summary(log_linear_mod)$sigma
```

Look at that RMSE! That dropped like a rock! Is that real?

No. RMSE before is in the units of y. And now y is now in ln($). So this RMSE and the previous RMSE are *not* comparable.

The $R^2$ are *not* comparable either. Even though they're on a [0, 1] scale in both models, the SST's are different so you're measuring the proportion of a different variance.

Let's attempt to compare apples-apples.

```{r}
log_y_hat = log_linear_mod$fitted.values
y_hat = exp(log_y_hat)
e = diamonds$price - y_hat
SSE = sum(e^2)
SST = sum((diamonds$price - mean(diamonds$price))^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
```

What does this mean? It means this was a bad idea. Those residuals for large carats are insanely large. They're wrong on a log scale! Which means they're off by orders of magnitude. Working with logged y is dangerous business if you're wrong! Before you were off by a few thousand dollars; now you're off by millions. For example. Let's look at a large diamond:

```{r}
xstar = diamonds[diamonds$carat > 5, ][1, ]
xstar$price
predict(mod, xstar)
exp(predict(log_linear_mod, xstar))
```

That's a pretty bad residual!

How about log-log model? 

```{r}
log_log_linear_mod = lm(log(price) ~ log(carat), diamonds)
b = coef(log_log_linear_mod)
b
```

Let's see what it looks like:

```{r}
ggplot(diamonds, aes(x = log(carat), y = log(price))) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green")
```

Well look at that! That's a nice looking model. (Note that the slope coefficients in log-log models, i.e. b_2 here, are called "elasticity" in Econ 382 as it measures how the relative change in x affects the relative change in y).

How are our metrics?

```{r}
summary(log_log_linear_mod)$r.squared
summary(log_log_linear_mod)$sigma
```

Let's see apples-to-apples to the natural y model.

```{r}
log_y_hat = log_log_linear_mod$fitted.values
y_hat = exp(log_y_hat)
e = diamonds$price - y_hat
SSE = sum(e^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
```

This is on-par with the vanilla OLS model, but still doesn't "beat it". There was no guarantee that we would be "beat it" even though we used procedures that are reasonable and popular! My belief is that we really should be wary of the maximum price and employ that in the model. Maybe we'll do this in a lab exercise?

Let's repeat this entire exercise using the length of the diamond. The length of the diamond feature is confusingly named "x" in the dataset. It is an "x" but it's also the diamond's "x"!!!

```{r}
ggplot(diamonds, aes(x = x, y = price)) + 
  geom_point()
```

Besides the non-linear relationship, what else do you see? Mistakes in the dataset! Can a real diamond have zero length?? Yes. This is the real world. There are mistakes all the time.

Let's kill it! How many are we dealing with here?

```{r}
nrow(diamonds[diamonds$x == 0, ])
```


```{r}
diamonds = diamonds[diamonds$x != 0, ]
```

What's the deal with the x variable now?

```{r}
skimr::skim(diamonds$x)
```

How good does a best guess linear relationship do?

```{r}
mod = lm(price ~ x, diamonds)
b = coef(mod)
b
summary(mod)$r.squared
summary(mod)$sigma
```

Let's see the best fit line $g(x)$ visually:

```{r}
ggplot(diamonds, aes(x = x, y = price)) + geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green")
```

Again we got some bad extrapolation going on which we can't fix using a purely linear modeling strategy.

Let's log-linearize it and see how we do.

```{r}
log_linear_mod = lm(log(price) ~ x, diamonds)
b = coef(log_linear_mod)
ggplot(diamonds, aes(x = x, y = log(price))) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green")
```

How did we do? Ensure it's apples-apples.

```{r}
log_y_hat = log_linear_mod$fitted.values
y_hat = exp(log_y_hat)
e = diamonds$price - y_hat
SSE = sum(e^2)
SST = sum((diamonds$price - mean(diamonds$price))^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
```

Still not better. Log-log?

```{r}
log_log_linear_mod = lm(log(price) ~ log(x), diamonds)
b = coef(log_log_linear_mod)
ggplot(diamonds, aes(x = log(x), y = log(price))) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green")
```

How did we do? 

```{r}
log_y_hat = log_log_linear_mod$fitted.values
y_hat = exp(log_y_hat)
e = diamonds$price - y_hat
SSE = sum(e^2)
SST = sum((diamonds$price - mean(diamonds$price))^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
```

We did it. We found a log transformation that seems to give higher predictive power than the vanilla linear model on the raw repsonse and raw feature.

This brings up the whole idea of "model selection". We went hunting for models until we found one that's better. We will hopefully do model selection today...

Transforming y is a big decision as it changes the response metric! The rule of thumb is it is easier to model a response metric that has less extreme values (especially when using linear models) as the extreme values have a big impact on slope coefficients and can distort the best fit line due to the least squares minimization (hence the popularity of logging the response).

Let's see if we get anywhere with this using all the features in this model.

```{r}
lm_y = lm(price ~ ., diamonds)
lm_ln_y = lm(log(price) ~ ., diamonds)
summary(lm_y)$r.squared
summary(lm_y)$sigma

#now for the log-linea model
y_hat = exp(lm_ln_y$fitted.values)
e = diamonds$price - y_hat
SSE = sum(e^2)
SST = sum((diamonds$price - mean(diamonds$price))^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
``` 

This is pretty convincing evidence that this transformation does a better job (at least in our linear modeling context).

Let's look at one prediction:

```{r}
predict(lm_y, diamonds[12345, ])
exp(predict(lm_ln_y, diamonds[12345, ]))
diamonds$price[12345]
```

Again, we should be careful when you use $g$ after logging, you will have to exponentiate the result (middle line above). 

Small point: this exponentiation is known to create bias because $E[Y]$ is different from $exp(E[ln(y)])$ (for those who took 368 - remember Jensen's inequality?) For the purposes of this class, this can be ignored since we are evaluating g on its own merits and we're doing so honestly.

If you like this stuff, there are a whole bunch of transformations out there that are even cooler than the natural log. 

# Linear Models with Feature Interactions

Let's go back to modeling price with weight. Let us add a third variable to this plot, color, a metric about the "yellowness" of the diamond. This is an ordinal categorical variable ranging from D (most clear i.e. best) to J (most yellow in this dataset i.e. worst).


```{r}
pacman::p_load(ggplot2)
base = ggplot(diamonds, aes(x = carat, y = price)) 
base +
  geom_point(aes(col = color)) + scale_color_brewer(type = "div")
```

We can split the data on color to see it more clearly:

```{r}
base +
  geom_point() +
  facet_wrap(~ color, ncol = 3) + 
  aes(color = color) + scale_color_brewer(type = "div")
```


What do we see here? It looks like the slope of the price vs. carat linear model is slightly affected by color. For instance, the "D" color diamonds' price increases much faster as weight increases than the "E" color diamonds' price increases in weight, etc. Why do you think this is?

We can picture two of these linear models below by fitting two submodels, one for D and one for J:

```{r}
mod_D = lm(price ~ carat, diamonds[diamonds$color == "D", ])
b_D = coef(mod_D)
mod_J = lm(price ~ carat, diamonds[diamonds$color == "J", ])
b_J = coef(mod_J)
b_D
b_J
```

Let's see it on the plot:

```{r}
base +
  geom_point(aes(col = color)) + scale_color_brewer(type = "div") +
  geom_abline(intercept = b_D[1], slope = b_D[2], col = "blue", lwd = 2) +
  geom_abline(intercept = b_J[1], slope = b_J[2], col = "red", lwd = 2)
```

This indicates a separate intercept and carat-slope for each color. How is this done? Interacting carat and slope. The formula notation has the `*` operator for this. It is multiplication in formula land after all!

```{r}
mod = lm(price ~ color, diamonds)
coef(mod)
mod = lm(price ~ carat * color, diamonds)
coef(mod) #beware: strange naming convention on the interaction terms for an ordered factor
diamonds$color = factor(diamonds$color, ordered = FALSE)
mod = lm(price ~ carat * color, diamonds)
coef(mod) #much better...
```

The reference category is color D. This means every other color should start lower and have a lower slope. This is about what we see above.

How much of a better model is this than a straight linear model?

```{r}
mod_vanilla = lm(price ~ carat + color, diamonds)
summary(mod_vanilla)$r.squared
summary(mod_vanilla)$sigma
summary(mod)$r.squared
summary(mod)$sigma
```

You can get more predictive accuracy out of this. We added a degree of freedom? Is this gain real? Yes. With one more feature and $n = 54,000$ there is no chance this gain came from overfitting noise. Add 10,000 garbage features, yes, there will be overgitting.

Let's take a look at carat with another variable, depth, a continuous predictor. High depth indicates diamonds are skinny and tall; low depth indicates diamonds are flat like a pancake.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(col = depth), lwd = 1, alpha = 0.5) + scale_colour_gradientn(colours = rainbow(5))
```

It seems people like flatter diamonds and are willing to pay more per carat. Let's see this in the regression:

```{r}
mod = lm(price ~ carat + depth, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ carat * depth, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
```

If carat increases by one unit, how much does price increase by? A tiny amount of increase.

How about cut?


```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(col = cut), lwd = 0.5) + scale_color_brewer(type = "div")
```

Likely something here.

```{r}
mod = lm(price ~ carat, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ carat + cut, diamonds)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ carat * cut, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
```

Yes.

Can we include all these interactions?

```{r}
mod = lm(price ~ carat + color + depth + cut, diamonds)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ carat * (color + depth + cut), diamonds)
summary(mod)$r.squared
summary(mod)$sigma
coef(mod)
```

A decent gain once again.

What does the design matrix look like there? What is $p$?

```{r}
diamonds$cut = factor(diamonds$cut, ordered = FALSE)
Xmm = model.matrix(price ~ carat * (color + depth + cut), diamonds)
head(Xmm)
```


Can we take a look at interactions of two categorical variables?


```{r}
plot1 = ggplot(diamonds, aes(x = cut, y = color)) +
  geom_jitter(aes(col = price), lwd = 0.5) + scale_colour_gradientn(colours = rainbow(5))
plot1
```

Cool animation possible. May not work because it needs a ton of packages...

```{r}
pacman:::p_load_gh("dgrtwo/gganimate")
plot1 + transition_time(price)
```

Not so clear what's going on here. Let's see what the regressions say:


```{r}
mod = lm(price ~ color + cut, diamonds)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ color * cut, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
```

Not too much gain.
