---
title: "Practice Lecture 14 MATH 342W Queens College"
author: "Professor Adam Kapelner"
date: "March 17, 2021"
---

# Orthogonal vs raw polynomials

I forgot to talk about this last class...

Why is orthonormal polynomial the default? You can argue that doing a QR decomposition on the polynomial expansion and employing Q in the design matrix will change b thereby making b uninterpretable! So why use orthonormal polynomials? Here's why:

```{r}
n = 1000
set.seed(1984)
X = data.frame(x = c(runif(n / 2, 0, 1e-2), runif(n / 2, 0, 1e6)))
d = 10
num_digits = 8
Xmm_orth = model.matrix(~ 0 + poly(x, d), X)
colnames(Xmm_orth)[1 : d] = 1 : d
Xmm_raw = model.matrix(~ 0 + poly(x, d, raw = TRUE), X)
colnames(Xmm_raw)[1 : d] = 1 : d
```

Let's look at the design matrix for small values of x:

```{r}
head(as.matrix(X))
round(head(Xmm_orth), num_digits)
round(head(Xmm_raw), num_digits)
```

You get numerical underflow almost immediately when using the raw polynomial computations (you get it by degree 4). And thus you can't even get the OLS estimates:


```{r}
y = rnorm(n)
solve(t(Xmm_raw) %*% Xmm_raw) %*% t(Xmm_raw) %*% y
```

Let's look at the design matrix for large values of x:

```{r}
tail(as.matrix(X))
round(tail(Xmm_orth), num_digits)
round(tail(Xmm_raw), num_digits)
```

You get numerical overflow in the design matrix (but it will happen later). But the second you start to use the design matrix with 10^59's inside...

```{r}
solve(t(Xmm_raw[800 : 1000, ]) %*% Xmm_raw[800 : 1000, ]) %*% t(Xmm_raw[800 : 1000, ]) %*% y
```

As opposed to

```{r}
solve(t(Xmm_orth) %*% Xmm_orth) %*% t(Xmm_orth) %*% y
```

No problem at all!!!

So that's the reason: numerical stability. But if you need interpretability, you need raw polynomials. But if you're interpreting the model, how do you even interpret beyond degree 2???

# Log transformations

We will be examining the diamonds dataset. Let's take a moment to read about our data. In real life, you will take more than just a moment to learn about the data. And more than just a moment to clean the data and do sanity checks. As these steps are more of the grunt work of data science (and are learned on the fly), I won't really cover them formally.

```{r}
rm(list = ls())
pacman::p_load(ggplot2) #this loads the diamonds data set too
?diamonds
dim(diamonds)
```

That's a huge $n$. So, let's expect things to take a bit longer when processing.

A natural increasing relationship will likely be found between weight and price. Let's see it visually:

```{r}
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point()
```


How good does a best guess linear relationship do?

```{r}
mod = lm(price ~ carat, diamonds)
b = coef(mod)
b
summary(mod)$r.squared
summary(mod)$sigma
```

What does the intercept say about extrapolation?

Let's see the best fit line $g(x)$ visually:

```{r}
ggplot(diamonds, aes(x = carat, y = price)) + geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green")
```

Maybe the relationship between weight and price is not linear - but exponential? E.g. if the weight of a diamond doubles, maybe the price quadruples? Or linear increases in weight yield percentage increases in price. This seems plausible. 

Besides the plausibility of the functional form, there is an agnostic reason to employ log y as the predictive target. Let's first examine the univariate data!


```{r}
skimr::skim(diamonds$price)
```

Very large standard error and very long tail which can be seed more clearly here:

```{r}
ggplot(diamonds) + geom_histogram(aes(price), binwidth = 200)
```

Let's take a look at the distribution after logging:

```{r}
ggplot(diamonds) + geom_histogram(aes(x = log(price)), binwidth = 0.01)
```

Some strange artifacts appear. Why the gap? Why is it "cut" sharply at a maximum? These are questions to ask the one who collected the data. But let's get back to the log story...

Popular wisdom says logging this type of highly skewed-right distribution would possibly make the model "more linear in x". Put another way, it would be easier to "catch" (predict) the long tail since it won't be a long tail anymore after you log-transform. It would also prevent observations with large y's becoming "leverage points" i.e. points that unduly influence the model and thereby warp its ability to predict the average observation. If we have time later in the semester, we can learn about leverage points, but not now. In Econ 382 you learn more reasons for why you should use log the response, but those reasons are relevant for inference so we won't discuss them here.

Let's give the model with ln(y) a whirl. Maybe we'll even learn something about diamonds. The way to create such a model is to simply fit an OLS model to log y. This is called a log-linear model. Since this is a pretty standard thing to do so R's formula notation has it built-in as follows:

```{r}
log_linear_mod = lm(log(price) ~ carat, diamonds)
b = coef(log_linear_mod)
b
```

Let's see what this looks like.

```{r}
ggplot(diamonds, aes(x = carat, y = log(price))) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green") + ylim(6, 12)
```

It looks very bad if carat is large. That means our little theory about carats getting larger yielding multiples of price doesn't correspond to reality.

How did we do?

```{r}
summary(log_linear_mod)$r.squared
summary(log_linear_mod)$sigma
```

Look at that RMSE! That dropped like a rock! Is that real?

No. RMSE is in the units of y. And y is now in ln($). So this RMSE and the previous RMSE are *not* comparable.

The $R^2$ are not comparable either. Even though they're on a [0, 1] scale in both models, the SST's are different so you're measuring the proportion of a different variance.

Let's attempt to compare apples-apples.

```{r}
log_y_hat = log_linear_mod$fitted.values
y_hat = exp(log_y_hat)
e = diamonds$price - y_hat
SSE = sum(e^2)
SST = sum((diamonds$price - mean(diamonds$price))^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
```

What does this mean? It means this was a bad idea. Those residuals for large carats are insanely large. They're wrong on a log scale! Which means they're off by orders of magnitude. Working with logged y is dangerous business if you're wrong! Before you were off by a few thousand dollars; now you're off by millions. For example. Let's look at a large diamond:

```{r}
xstar = diamonds[diamonds$carat > 5, ][1, ]
xstar$price
predict(mod, xstar)
exp(predict(log_linear_mod, xstar))
```

That's a pretty bad residual!

How about log-log model? 

```{r}
log_log_linear_mod = lm(log(price) ~ log(carat), diamonds)
b = coef(log_log_linear_mod)
b
```

Let's see what it looks like:

```{r}
ggplot(diamonds, aes(x = log(carat), y = log(price))) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green")
```

Well look at that! That's a nice looking model. (Note that the slope coefficients in log-log models, i.e. b_2 here, are called "elasticity" in Econ 382 as it measures how the relative change in x affects the relative change in y).

How are our metrics?

```{r}
summary(log_log_linear_mod)$r.squared
summary(log_log_linear_mod)$sigma
```

Let's see apples-to-apples.

```{r}
log_y_hat = log_log_linear_mod$fitted.values
y_hat = exp(log_y_hat)
e = diamonds$price - y_hat
SSE = sum(e^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
```

This is on-par with the vanilla OLS model, but still doesn't "beat it". There was no guarantee that we would be "beat it" even though we used procedures that are reasonable and popular! My belief is that we really should be wary of the maximum price and employ that in the model. Maybe we'll do this in a lab exercise?

Let's repeat this entire exercise using the length of the diamond. The length of the diamond feature is confusingly named "x" in the dataset. It is an "x" but it's also the diamond's "x"!!!

```{r}
ggplot(diamonds, aes(x = x, y = price)) + 
  geom_point()
```

Besides the non-linear relationship, what else do you see? Mistakes in the dataset! Can a real diamond have zero length?? Yes. This is the real world. There are mistakes all the time.

Let's kill it! How many are we dealing with here?

```{r}
nrow(diamonds[diamonds$x == 0, ])
```


```{r}
diamonds = diamonds[diamonds$x != 0, ]
```

What's the deal with the x variable now?

```{r}
skimr::skim(diamonds$x)
```

How good does a best guess linear relationship do?

```{r}
mod = lm(price ~ x, diamonds)
b = coef(mod)
b
summary(mod)$r.squared
summary(mod)$sigma
```

Let's see the best fit line $g(x)$ visually:

```{r}
ggplot(diamonds, aes(x = x, y = price)) + geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green")
```

Again we got some bad extrapolation going on which we can't fix using a purely linear modeling strategy.

Let's log-linearize it and see how we do.

```{r}
log_linear_mod = lm(log(price) ~ x, diamonds)
b = coef(log_linear_mod)
ggplot(diamonds, aes(x = x, y = log(price))) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green")
```

How did we do? Ensure it's apples-apples.

```{r}
log_y_hat = log_linear_mod$fitted.values
y_hat = exp(log_y_hat)
e = diamonds$price - y_hat
SSE = sum(e^2)
SST = sum((diamonds$price - mean(diamonds$price))^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
```

Still not better. Log-log?

```{r}
log_log_linear_mod = lm(log(price) ~ log(x), diamonds)
b = coef(log_log_linear_mod)
ggplot(diamonds, aes(x = log(x), y = log(price))) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green")
```

How did we do? 

```{r}
log_y_hat = log_log_linear_mod$fitted.values
y_hat = exp(log_y_hat)
e = diamonds$price - y_hat
SSE = sum(e^2)
SST = sum((diamonds$price - mean(diamonds$price))^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
```

We did it. We found a log transformation that seems to give higher predictive power than the vanilla linear model on the raw repsonse and raw feature.

This brings up the whole idea of "model selection". We went hunting for models until we found one that's better. We will hopefully do model selection today...

Transforming y is a big decision as it changes the response metric! The rule of thumb is it is easier to model a response metric that has less extreme values (especially when using linear models) as the extreme values have a big impact on slope coefficients and can distort the best fit line due to the least squares minimization (hence the popularity of logging the response).

Let's see if we get anywhere with this using all the features in this model.

```{r}
lm_y = lm(price ~ ., diamonds)
lm_ln_y = lm(log(price) ~ ., diamonds)
summary(lm_y)$r.squared
summary(lm_y)$sigma

#now for the log-linea model
y_hat = exp(lm_ln_y$fitted.values)
e = diamonds$price - y_hat
SSE = sum(e^2)
SST = sum((diamonds$price - mean(diamonds$price))^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
``` 

This is pretty convincing evidence that this transformation does a better job (at least in our linear modeling context).

Let's look at one prediction:

```{r}
predict(lm_y, diamonds[12345, ])
exp(predict(lm_ln_y, diamonds[12345, ]))
diamonds$price[12345]
```

Again, we should be careful when you use $g$ after logging, you will have to exponentiate the result (middle line above). 

Small point: this exponentiation is known to create bias because $E[Y]$ is different from $exp(E[ln(y)])$ (for those who took 368 - remember Jensen's inequality?) For the purposes of this class, this can be ignored since we are evaluating g on its own merits and we're doing so honestly.

If you like this stuff, there are a whole bunch of transformations out there that are even cooler than the natural log. 

# Linear Models with Feature Interactions

Let's go back to modeling price with weight. Let us add a third variable to this plot, color, a metric about the "yellowness" of the diamond. This is an ordinal categorical variable ranging from D (most clear i.e. best) to J (most yellow in this dataset i.e. worst).


```{r}
pacman::p_load(ggplot2)
base = ggplot(diamonds, aes(x = carat, y = price)) 
base +
  geom_point(aes(col = color)) + scale_color_brewer(type = "div")
```

We can split the data on color to see it more clearly:

```{r}
base +
  geom_point() +
  facet_wrap(~ color, ncol = 5) + 
  aes(color = color) + scale_color_brewer(type = "div")
```


What do we see here? It looks like the slope of the price vs. carat linear model is slightly affected by color. For instance, the "D" color diamonds' price increases much faster as weight increases than the "E" color diamonds' price increases in weight, etc. Why do you think this is?

We can picture two of these linear models below by fitting two submodels, one for D and one for J:

```{r}
mod_D = lm(price ~ carat, diamonds[diamonds$color == "D", ])
b_D = coef(mod_D)
mod_J = lm(price ~ carat, diamonds[diamonds$color == "J", ])
b_J = coef(mod_J)
b_D
b_J
```

Let's see it on the plot:

```{r}
base +
  geom_point(aes(col = color)) + scale_color_brewer(type = "div") +
  geom_abline(intercept = b_D[1], slope = b_D[2], col = "blue", lwd = 2) +
  geom_abline(intercept = b_J[1], slope = b_J[2], col = "red", lwd = 2)
```

This indicates a separate intercept and carat-slope for each color. How is this done? Interacting carat and slope. The formula notation has the `*` operator for this. It is multiplication in formula land after all!

```{r}
mod = lm(price ~ color, diamonds)
coef(mod)
mod = lm(price ~ carat * color, diamonds)
coef(mod) #beware: strange naming convention on the interaction terms for an ordered factor
diamonds$color = factor(diamonds$color, ordered = FALSE)
mod = lm(price ~ carat * color, diamonds)
coef(mod) #much better...
```

The reference category is color D. This means every other color should start lower and have a lower slope. This is about what we see above.

How much of a better model is this than a straight linear model?

```{r}
mod_vanilla = lm(price ~ carat + color, diamonds)
summary(mod_vanilla)$r.squared
summary(mod_vanilla)$sigma
summary(mod)$r.squared
summary(mod)$sigma
```

You can get more predictive accuracy out of this. We added a degree of freedom? Is this gain real? Yes. With one more feature and $n = 54,000$ there is no chance this gain came from overfitting noise. Add 10,000 garbage features, yes, there will be overgitting.

Let's take a look at carat with another variable, depth, a continuous predictor. High depth indicates diamonds are skinny and tall; low depth indicates diamonds are flat like a pancake.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(col = depth), lwd = 1, alpha = 0.5) + scale_colour_gradientn(colours = rainbow(5))
```

It seems people like flatter diamonds and are willing to pay more per carat. Let's see this in the regression:

```{r}
mod = lm(price ~ carat, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ carat * depth, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
```

If carat increases by one unit, how much does price increase by?

Is this better than the model without the interaction?

```{r}
mod = lm(price ~ carat + depth, diamonds)
summary(mod)$r.squared
summary(mod)$sigma
```

A tiny amount of increase.

How about cut?


```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(col = cut), lwd = 0.5) + scale_color_brewer(type = "div")
```

Likely something here.

```{r}
mod = lm(price ~ carat, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ carat + cut, diamonds)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ carat * cut, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
```

Yes.

Can we include all these interactions?

```{r}
mod = lm(price ~ carat * (color + depth + cut), diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ carat + color + depth + cut, diamonds)
summary(mod)$r.squared
summary(mod)$sigma
```

A decent gain once again.

What does the design matrix look like there? What is $p$?

```{r}
Xmm = model.matrix(price ~ carat * (color + depth + cut), diamonds)
head(Xmm) #you can see the strange naming convention here on cut since cut is an ordered factor (weird...)
class(diamonds$cut)
```


Can we take a look at interactions of two categorical variables?


```{r}
plot1 = ggplot(diamonds, aes(x = cut, y = color)) +
  geom_jitter(aes(col = price), lwd = 0.5) + scale_colour_gradientn(colours = rainbow(5))
plot1
```

Cool animation possible. May not work because it needs a ton of packages...

```{r}
pacman:::p_load_gh("dgrtwo/gganimate")
plot1 + transition_time(price)
```

Not so clear what's going on here. Let's see what the regressions say:


```{r}
mod = lm(price ~ color + cut, diamonds)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ color * cut, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
```

Not too much gain.

## For MA students only: Eigendecomposition in R and the Projection Matrix

```{r}
B = array(seq(1, 4), dim = c(2, 2))
B
eigen_decomp = eigen(B)
V = eigen_decomp$vectors
V
v1 = V[, 1, drop = FALSE]
v2 = V[, 2, drop = FALSE]
lambdas = eigen_decomp$values
lambdas
lambda1 = lambdas[1]
lambda2 = lambdas[2]
B %*% v1
lambda1 * v1
B %*% v2 
lambda2 * v2
B %*% v2 == lambda2 * v2 #why not?
#diagonolization
V %*% diag(lambdas) %*% solve(V)
```

Let's take a look at a projection matrix

```{r}
X = model.matrix(medv ~ ., MASS::Boston)
head(X)
H = X %*% solve(t(X) %*% X) %*% t(X)
dim(H)
eigen_decomp = eigen(H)
lambdas = eigen_decomp$values
head(lambdas, 50)
lambdas = as.numeric(eigen_decomp$values) #coerce to real numbers (good enough approximation)
head(lambdas, 50)
lambdas_rounded = round(lambdas, 10)
head(lambdas_rounded, 50)
table(lambdas_rounded)
sum(lambdas_rounded) #i.e. the rank(H) = # cols in X
```

It turns out all eigenvalues of a projection matrix are either 0 or 1. What are the eigenvectors? Only the first ncol(X) = 14 reconstruct the space. Are they the same as the X columns?

```{r}
V = eigen_decomp$vectors
dim(V)
V = matrix(as.numeric(V), ncol = ncol(V))
dim(V)
V[1 : 20, 1 : ncol(X)]

V_X_angles = matrix(NA, ncol(X), ncol(X))
for (i in 1 : ncol(X)){
  for (j in 1 : ncol(X)){
    if (j > i){
      V_X_angles[i, j] = acos(V[, i] %*% X[, j] / sqrt(sum(V[, i]^2) * sum(X[, j]^2))) * 180 / pi
    }
  }
}
V_X_angles
```

No... and they don't need to be. They just need to represent the same space i.e. their columns need to span the same 14-dimensional subspace. You can check this by just comparing the orthogonal projection matrix which is unique for each subspace.

```{r}
V_X = V[, 1 : ncol(X)]
H_V = V_X %*% solve(t(V_X) %*% V_X) %*% t(V_X)
dim(H_V)
testthat::expect_equal(c(H_V), c(H))
```

And they're the same as expected.

# Popular Add-on Packages to R

## Piping

Take a look at this one-liner:

```{r}
set.seed(1984)
mean(head(round(sample(rnorm(1000)), digits = 2)))
```

This is hard to read. Of course we can make it easier by using breaklines e.g.

```{r}
mean(
  head(
    round(
      sample(
        rnorm(1000)
      ), digits = 2)
    )
)
```

But it doesn't make it much easier to read. And it probably makes it harder to write.

Enter an idea taken from unix / linux. Output of one function is input to next function. It is the inverse of the usual "order of operations". Let's see how this works.

We first load the piping library:

```{r}
pacman::p_load(magrittr) 
```

The package is named after Rene Magritte, the Belgian surrealist artist because he wrote [(Ceci n'est pas un pipe)](https://en.wikipedia.org/wiki/The_Treachery_of_Images) on a painting of a pipe.

In pipe format this would look like:

```{r}
set.seed(1984)
rnorm(1000) %>% #the pipe operator
  sample %>% 
  round(digits = 2) %>% #the first argument is passed in automatically.
  head %>%
  mean
```

That's it! There's nothing more to it other than a gain in readability.

What if we wanted to do something like `mean(rnorm(1000) + 1)`? This `rnorm(1000) %>% +1 %>% mean` doesn't work because I imagine because the basic arithmetic operators couldn't be parsed like normal while there was a pipe. So they invented special pipe functions for this:

```{r}
rnorm(1000) %>% add(1) %>% mean
```

There are other exceptions to the rule too which you'll figure out if you adopt the pipe.

Unfortunately... the world at large hasn't completely accepted this as a way to write R. So feel free to use for yourself. But be carefuly when using this style with others. 

Also note the code you write with pipes will be slower than the normal syntax.

# The Grammar of graphics and ggplot

First load the package and the dataset of interest as a dataframe:

```{r}
pacman::p_load(ggplot2, quantreg)
cars = MASS::Cars93 #dataframe
```

ggplot is based on the "Grammar of Graphics", a concept invented by the Statistician / Computer Scientist Leland Wilkinson who worked on SPSS, Tableau and now he works at H20, software that analyzes big data. The reference of interest is [here](http://papers.rgrossman.com/proc-094.pdf). He drew on ideas from John Tukey (one of the great statistician of the previous generation) while he was at Bell Labs, Andreas Buja (one of my professors at Penn) and Jerome Friedman (the professor that taught my data mining course at Stanford - remember the diamonds story?) 

It is a language that allows us to describe the components of a graphic. Previously, graphics were done in one shot and it was clunky. ggplot is a library written by Hadley Wickham based on this concept. Wickham is probably the most famous person in statistical computing today. He has commit rights in R and is one of the architects of RStudio. He calls grammar of graphics "graphical poems". Here are the basic components:

* an underlying data frame
* an "aesthetic" that maps visualization axes in the eventual plot(s) to variables in the data frame
* a "layer" which is composed of
  - a geometric object
  - a statistical transformation
  - a position adjustment
* a "scale" for each aesthetic
* a "coordinate" system for each aesthetic
* optional "facets" (more graphics)
* optional "labels" for the title, axes title, points, etc.

Don't worry - everything has "smart defaults" in Wickham's implementation so you don't have to worry about most things. We will explore some of the features below. Here's a good [cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf).

ggplot is layered where each component is an object. The objects are added to each other since the "+" operator is overloaded to accept these additions. This is nice because each component can be saved and reused. The following initialized the graphics:

```{r}
ggplot(cars)
```

Nothing happened - except the plot window went gray (the smart default). This is already rendering a graphic, but since it hasn't been given any of the required information, it has nothing to display. Next we create an aesthetics indicating a one-way plot (one variable only).

```{r}
ggplot(cars) + 
  aes(Price)
```

Notice how it can understand the variable name as an object name.

Since we've given it an aesthetics object, it now knows which variable is the x axis (default). It already knows the ranges of the variable (a smart default) and a default scale and coordinate system (smart defaults).

Usually this is done in one step by passing the aesthetics object into the ggplot:

```{r}
ggplot(cars, aes(Price))
```

Now we need to pick a layer by specifying a geometry. This is a type of plot. Since the predictor type of price is continuous, let's pick the "histogram" using the `geom_histogram` function:

```{r}
ggplot(cars, aes(Price)) +
  geom_histogram()
```

This can be customized:

```{r}
ggplot(cars, aes(Price)) +
  geom_histogram(binwidth = 1, col = "darkgreen", fill = "blue", alpha = 0.4)
```

Want to save it for your latex?

```{r}
ggsave("plot.png")
system("open plot.png")
ggsave("plot.pdf")
system("open plot.pdf")
```

Here are some other options besides the histogram:

```{r}
ggplot(cars, aes(Price)) +
  geom_dotplot()
ggplot(cars, aes(Price)) +
  geom_area(stat = "bin", binwidth = 2)
ggplot(cars, aes(Price)) +
  geom_freqpoly()
ggplot(cars, aes(Price)) +
  geom_density(fill = "green", alpha = 0.4)

summary(cars)
```


Can we compare price based on different conditions? Yes, we can subset the data and use color and alpha:

```{r}
ggplot(cars, aes(Price)) +
  geom_density(data = subset(cars, Man.trans.avail == "Yes"), col = "grey", fill = "darkgreen", alpha = 0.4) +
  geom_density(data = subset(cars, Man.trans.avail == "No"), col = "grey", fill = "red", alpha = 0.4)
```

Sidebar: why are cars that have manual transmissions available cheaper?

We can look at this also using a histogram of the conditional distributions:

```{r}
ggplot(cars, aes(Price)) +
  geom_histogram(data = subset(cars, Man.trans.avail == "Yes"), binwidth = 1, col = "grey", fill = "darkgreen", alpha = 0.4) +
  geom_histogram(data = subset(cars, Man.trans.avail == "No"), binwidth = 1, col = "grey", fill = "red", alpha = 0.4)
```

What if the variable is not continuous e.g. Cylinders? We can use a bar graph / bar plot.

```{r}
ggplot(cars, aes(Cylinders)) +
  geom_bar()
```

This is essential frequency by level of the categorical variable.

Now let's move on to looking at one variable versus another variable. For example price by engine power:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price))
```

Since we've given it an aesthetics object, it now knows which variable is the x axis and which variable is the y axis. It already knows the ranges of the variables (a smart default) and a default scale and coordinate system (smart defaults).

Just as before, now we need to pick a layer by specifying a geometry. This is a type of plot. Let's pick the "scatterplot" using the `geom_point` function:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point()
```

Now we have a nice scatterplot. This function uses the inherited data, the inherited aesthetics. Since this "geometry" is a "layer", we can pass in options to the layer.

```{r}
base_and_aesthetics = ggplot(cars, aes(x = Horsepower, y = Price))
base_and_aesthetics + geom_point(col = "red", fill = "green", shape = 23, size = 3, alpha = 0.3)
```

Let's handle names of axes, title and ranges:

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics +
  ggtitle("Average Car Price vs. Engine Power", subtitle = "in the Cars93 dataset") +
  ylab("Price (in $1000's)")
base_and_aesthetics_with_titles +
  geom_point() +
  xlim(0, 400) +
  ylim(0, 50)
  
```

Let's transform the variables:

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  scale_x_continuous(trans = "log2")
```

Each unit increase on the x-axis now represent a doubling increase in x (although the whole scale only spans 3 units). But look at how the grid didn't keep up. Let's fix this:

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  scale_x_continuous(trans = "log2", breaks = round(seq(0, max(cars$Horsepower), length.out = 6)))
```

We can do the same to the y axis:

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  scale_y_continuous(trans = "log10")
  scale_x_continuous(trans = "log10", breaks = round(seq(0, max(cars$Horsepower), length.out = 6)))
```

Let's look at some more geometries.

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  geom_smooth()
```

Here, I've added two geometries on the same aesthetic! This attempts to explain the relationship $f(x)$ using smoothing. Let's go for more.

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  geom_smooth() +
  geom_rug()
```

This allows us to also see the marginal distributions of the two variables.

```{r}
base_and_aesthetics_with_titles +
  geom_point() +
  geom_smooth() +
  geom_quantile(col = "red") +
  geom_rug()
```

This fits a line and tries to indicate statistical significance of the line. We have *not* covered any statistics in this class yet (ironic!) ... so ignore how the window is generated.

Can we display more than two dimensions? Yes. Let's indicate a third dimension with shape (only works with factors).

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power and Transmission")
base_and_aesthetics_with_titles +
  geom_point(aes(shape = Man.trans.avail)) +
  geom_smooth() +
  geom_rug()
```

Can we display more than three dimensions? Yes. Let's indicate a fourth dimension with color.

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power, Transmission & Drivetrain")
base_and_aesthetics_with_titles +
  geom_point(aes(shape = Man.trans.avail, col = DriveTrain)) +
  geom_smooth() +
  geom_rug()
```

Can we go to a fifth dimension? Maybe?

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power, Transmission & Drivetrain")
base_and_aesthetics_with_titles +
  geom_point(aes(shape = Man.trans.avail, col = DriveTrain, size = Weight), alpha = 0.5) + #size?
  geom_smooth() +
  geom_rug()
```

A seventh? We can use text labels adjacent to the scatterplot's points.

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power, Transmission, Drivetrain,  Weight & #Cylinders")
base_and_aesthetics_with_titles +
  geom_point(aes(shape = Man.trans.avail, col = DriveTrain, alpha = Weight)) + #size?
  geom_text(aes(label = Cylinders), vjust = 1.5, col = "darkgrey", lineheight = 0.3, size = 3) +
  geom_smooth() +
  geom_rug()
```

Getting difficult to see what's going on.

Let's move away from the scatterplot to just density estimation:

```{r}
base_and_aesthetics_with_titles = base_and_aesthetics_with_titles +
  ggtitle("Average Car Price by Power") #reset the title
base_and_aesthetics_with_titles +
  geom_density2d()
```

Other alternatives:

```{r}
base_and_aesthetics_with_titles +
  geom_bin2d(binwidth = c(8, 3))
pacman::p_load(hexbin)
base_and_aesthetics_with_titles +
  geom_hex()
```

This is like a two-dimensional histogram where the bar / hexagon heights are seen with color.

What if the x-axis is categorical for example Cylinders versus price? Typical is the "box and whiskers" plot:

```{r}
ggplot(cars, aes(x = Cylinders, y = Price)) +
  geom_boxplot()
```

Clear relationship!

How about multiple subplots based on the subsetting we did in the histograms? This is called "faceting". Here are two bivariate visualizations laid horizontally:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point() +
  geom_smooth() +
  facet_grid(. ~ Man.trans.avail)
```

Or alternatively, vertically:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point() +
  geom_smooth() +
  facet_grid(Man.trans.avail ~ .)
```

And we can even double-subset:

```{r}
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point() +
  facet_grid(Man.trans.avail ~ Origin)
```

And we can even triple-subset or more:

```{r}
cars$MedWeight = ifelse(cars$Weight > median(cars$Weight), ">MedWeight", "<MedWeight")
ggplot(cars, aes(x = Horsepower, y = Price)) +
  geom_point() +
  facet_grid(Man.trans.avail ~ Origin + MedWeight, scales = "free")
```

These three varibles seem somewhat independent.

There are other primitives like `geom_abline` which graphs a line and `geom_segment` we will see today. Note that if you want plots rendered within functions or loops you have to explicitly call the `plot` function:

```{r}
for (nsim in 1 : 3){
  graphics_obj = ggplot(data.frame(x = rnorm(1000))) + 
    geom_histogram(aes(x))
  graphics_obj
}
```

versus:

```{r}
for (nsim in 1 : 3){
  graphics_obj = ggplot(data.frame(x = rnorm(1000))) + geom_histogram(aes(x))
  plot(graphics_obj)
}
```


Lastly, ggplot offers lots of nice customization themes:

```{r}
graphics_obj = base_and_aesthetics_with_titles +
  geom_point() +
  geom_smooth() +
  geom_quantile(col = "red") +
  geom_rug()
graphics_obj + theme_bw()
graphics_obj + theme_dark()
graphics_obj + theme_classic()

```

Packages offer even more:

```{r}
pacman::p_load(forcats, lazyeval, ggthemes)
graphics_obj + theme_economist()
graphics_obj + theme_stata()
graphics_obj + theme_tufte()
```

and of course, the whimsical one and only:


```{r}
pacman::p_load(xkcd, extrafont)
download.file("http://simonsoftware.se/other/xkcd.ttf", dest = "xkcd.ttf", mode = "wb")
#MAC
# system("mv xkcd.ttf /Library/Fonts")
# font_import(path = "/Library/Fonts", pattern = "xkcd", prompt = FALSE)
# fonts()
# fonttable()
# loadfonts()
#WINDOWS
font_import(path = ".", pattern = "xkcd", prompt = FALSE)
fonts()
fonttable()

loadfonts(device="win")

graphics_obj + theme_xkcd()
```

# C++ and R

R goes back to 1995 when it was adapted from S (written in 1976 by John Chambers at Bell Labs) with minor modifications. The core of base R is written in C and Fortran. These two languages are the fastest known languages (how to measure "fastest" is a huge debate). Thus, base R is very fast. For instance the `sort` function is as fast as C/Fortran since it immediately calls compiled C/Fortran routines.

However, R code itself that you write is "interpreted" which means it is not compiled until you run it. And it has to compile on-the-fly, making it very slow. Prior to v3.4 (April, 2017) it was even slower since the code wasn't JIT compiled. All this "real CS" stuff you can learn in another class..

One notable place to observe this slowness relative to other languages is in looping. For example:


```{r}
f = function(){
  SIZE = 1e7
  v = array(NA, SIZE)
  for (i in 1 : SIZE){
    v[i] = i
  }
  v
}

v = f()
```

How long does this take?

```{r}
# install.packages("Rcpp")
pacman::p_load_current_gh("hadley/lineprof")
lineprof(f())

system.time({
  f()
})
#on my office computer for SIZE = 10,000,000:
# user  system elapsed 
# 7.13    0.20    7.35 
```

Take a simple function that computes square roots on each element:

```{r}
sqrt_vector = function(v){
  v_new = array(NA, length(v))
  for (i in 1 : length(v)){
    v_new[i] = sqrt(v[i])
  }
  v_new
}

system.time({
  sqrt_vector(v)
})
```

Does the apply function help?

```{r}
system.time({
  apply(v, MARGIN = 1, FUN = sqrt)
})
```

Strange that this takes so long? So it doesn't help... it hurts.

How much faster in C++ should this be?

Enter the `Rcpp` package - a way to compile little bits (or lotta bits) of C++ on the fly.

```{r}
pacman::p_load(Rcpp)
```


Let's write this for loop function to sqrt-ize and compile it and then save it into our namespace to be called like a regular function.

```{r}
cppFunction('
  NumericVector sqrt_vector_cpp(NumericVector v) {
    int n = v.size();
    NumericVector v_new(n);
    for (int i = 0; i < n; i++) { //indices from 0...n-1 not 1...n!
      v_new[i] = sqrt(v[i]);
    }
    return v_new;
  }
')
```

What do these two functions look like?

```{r}
sqrt_vector
sqrt_vector_cpp
```

One shows the R code and then says it is bytecode-compiled (go to an advanced CS class). The other just says we `.Call` some C++ function in a certain address and the argument to be inputted.

What is the gain in runtime?

```{r}
system.time({
  sqrt_vector_cpp(v)
})
```

WOW. 10x!!! Can't beat that with a stick...

Let's do a not-so-contrived example...

Matrix distance... Let's compute the distances of all pairs of rows in a dataset. I will try to code the R as efficiently as possible by using vector subtraction so there is only two for loops. The C++ function will have an additional loop to iterate over the features in the observations.

```{r}
#a subset of the diamonds data
X_diamonds = as.matrix(ggplot2::diamonds[1 : 3000, c("carat", "depth", "table", "x", "y", "z")])

compute_distance_matrix = function(X){
  n = nrow(X)
  D = matrix(NA, n, n)
  for (i_1 in 1 : (n - 1)){
    for (i_2 in (i_1 + 1) : n){
      D[i_1, i_2] = sqrt(sum((X[i_1, ] - X[i_2, ])^2))
    }
  }
  D
}

cppFunction('
  NumericMatrix compute_distance_matrix_cpp(NumericMatrix X) {
    int n = X.nrow();
    int p = X.ncol();
    NumericMatrix D(n, n);
    std::fill(D.begin(), D.end(), NA_REAL);

    for (int i_1 = 0; i_1 < (n - 1); i_1++){
      //Rcout << "computing for row #: " << (i_1 + 1) << "\\n";
      for (int i_2 = i_1 + 1; i_2 < n; i_2++){
        double sqd_diff = 0;
        for (int j = 0; j < p; j++){
          sqd_diff += pow(X(i_1, j) - X(i_2, j), 2); //by default the cmath library in std is loaded
        }
        D(i_1, i_2) = sqrt(sqd_diff); //by default the cmath library in std is loaded
      }
    }
    return D;
  }
')
```

```{r}
system.time({
  D = compute_distance_matrix(X_diamonds)
})
round(D[1 : 5, 1 : 5], 2)
```

Slow...

```{r}
system.time({
  D = compute_distance_matrix_cpp(X_diamonds)
})
round(D[1 : 5, 1 : 5], 2)
```

Absolutely lightning... 100x faster on my laptop than R's runtime.

Writing functions as strings that compile is annoying. It is better to have separate files. For instance...

```{r}
sourceCpp("distance_matrix.cpp")
```

Here are a list of the data structures in Rcpp: https://teuder.github.io/rcpp4everyone_en/070_data_types.html#vector-and-matrix

Another place where C++ pays the rent is recursion. Here is a quicksort implementation in R taken from 

```{r}
quicksort_R <- function(arr) {
  # Pick a number at random.
  mid <- sample(arr, 1)

  # Place-holders for left and right values.
  left <- c()
  right <- c()
  
  # Move all the smaller values to the left, bigger values to the right.
  lapply(arr[arr != mid], function(d) {
    if (d < mid) {
      left <<- c(left, d)
    }
    else {
      right <<- c(right, d)
    }
  })
  
  if (length(left) > 1) {
    left <- quicksort_R(left)
  }
  
  if (length(right) > 1) {
    right <- quicksort_R(right)
  }
  
  # Finally, return the sorted values.
  c(left, mid, right)
}
```

Let's create a random array to test these sorts on:

```{r}
n = 5e5
x = rnorm(n)
```


Let's profile the pure R sort function:

```{r}
system.time({
  x_sorted_pure_R = quicksort_R(x)
})
```

Let's profile R's `sort` function.

```{r}
system.time({
  x_sorted_base_R = sort(x)
})
```

Let's just ensure our method worked...

```{r}
pacman::p_load(testthat)
expect_equal(x_sorted_pure_R, x_sorted_base_R)
```

Basically infinitely faster. Let's make our own C++ implementation.

```{r}
sourceCpp("quicksort.cpp")
```

and profile it:

```{r}
system.time({
  x_sorted_cpp = quicksort_cpp(x)
})
```

Let's just ensure this method worked...

```{r}
pacman::p_load(testthat)
expect_equal(x_sorted_cpp, x_sorted_base_R)
```

Why is our C++ slower than `sort`. Because `sort` is also in C++ or Fortran and it's been likely optimized and reoptimized up to wazoo for decades. Also, Rcpp's data structures may be slower than base R's data structures. There may be some speed lost to translating to `NumericVector` from `double[]` or something like that.

Can you call R from Rcpp? You bet:

```{r}
cppFunction('
  NumericVector rnorm_cpp_R(int n, double mean, double sd){
      // get a pointer to R\'s rnorm() function
      Function f("rnorm");   
  
      // Next code is interpreted as rnorm(n, mean, sd)
      return f(n, Named("sd")=sd, _["mean"]=mean);
  }
')

rnorm_cpp_R(5, 1, .01)
```

A few math functions are implemented for you already:

```{r}
evalCpp('R::qnorm(0.5, 0, 1, 1, 0)')
```

Further, there are many common functions that are already wrapped for you via "Rcpp-sugar" which was the Rcpp's author's attempt to make Rcpp a whole lot easier, see [here](http://dirk.eddelbuettel.com/code/rcpp/Rcpp-sugar.pdf).

```{r}
evalCpp('rnorm(10, 100, 3)')
```

If you want blazing fast linear algebra, check out package `RcppArmadillo` which is a wrapper around Apache's Armadillo (namespace is "arma" in the code), an optimized linear algebra package in C++. Here is an example taken from [here](https://scholar.princeton.edu/sites/default/files/q-aps/files/slides_day4_am.pdf):

```{r}
pacman::p_load(RcppArmadillo, microbenchmark, testthat)

cppFunction('
  arma::mat ols_cpp(arma::mat X, arma::mat y){
    arma::mat Xt = X.t();
    return solve(Xt * X, Xt * y);
  }
', depends = "RcppArmadillo")

n = 500
D = data.frame(int = rep(1, n), x1 = rnorm(n), x2 = rnorm(n), x3 = rnorm(n), y = rnorm(n))
X = as.matrix(D[, 1 : 4])
y = as.matrix(D[, 5])

#does the function work?
expect_equal(as.numeric(ols_cpp(X, y)), as.numeric(solve(t(X) %*% X) %*% t(X) %*% y))

microbenchmark(
  R_via_lm = lm(y ~ 0 + ., data = D),
  R_matrix_multiplication = solve(t(X) %*% X) %*% t(X) %*% y,
  cpp = ols_cpp(X, y),
    times = 100
)
```

About 4x faster than R's optimized linear algebra routines. Supposedly it can go even faster if you enable parallelization within Armadillo. I couldn't get that demo to work.

Here are the places where Rcpp should be used (from https://teuder.github.io/rcpp4everyone_en/010_Rcpp_merit.html)

* Loop operations in which later iterations depend on previous iterations.
* Accessing each element of a vector/matrix.
* Recurrent function calls within loops.
* Changing the size of vectors dynamically.
* Operations that need advanced data structures and algorithms (we don't do this in this class).
