---
title: "Practice Lecture 23 MATH 342W Queens College"
author: "Professor Adam Kapelner"
date: "April 28, 2021"
---

# Bias-Variance Decomposition of Generalization Error

Let's try to fit a quadratic $f$ with a linear model and examine bias-variance tradeoff.

```{r}
rm(list = ls())
xmin = 0
xmax = 5
n_train = 20
n_test = 1000
sigma = 1

Nsim = 1000

training_gs = matrix(NA, nrow = Nsim, ncol = 2)
x_trains = matrix(NA, nrow = Nsim, ncol = n_train)
y_trains = matrix(NA, nrow = Nsim, ncol = n_train)
all_oos_residuals = matrix(NA, nrow = Nsim, ncol = n_test)
for (nsim in 1 : Nsim){
  #simulate dataset $\mathbb{D}$
  x_train = runif(n_train, xmin, xmax)
  delta_train = rnorm(n_train, 0, sigma) #Assumption I: mean zero and Assumption II: homoskedastic
  y_train = x_train^2 + delta_train
  x_trains[nsim, ] = x_train
  y_trains[nsim, ] = y_train
  
  #fit a model g | x's, delta's and save it
  g_model = lm(y_train ~ ., data.frame(x = x_train))
  training_gs[nsim, ] = coef(g_model)
  
  #generate oos dataset according to the same data generating process (DGP) 
  x_test = runif(n_test, xmin, xmax)
  delta_test = rnorm(n_test, 0, sigma)
  y_test = x_test^2 + delta_test
  #predict oos using the model and save the oos residuals
  y_hat_test = predict(g_model, data.frame(x = x_test))
  all_oos_residuals[nsim, ] = y_test - y_hat_test
}
```

Graph a few:

```{r}
pacman::p_load(ggplot2)
ggplot() + 
  geom_point(aes(x, y), data = data.frame(x = x_trains[1, ], y = y_trains[1, ]), col = "blue") +
  geom_point(aes(x, y), data = data.frame(x = x_trains[2, ], y = y_trains[2, ]), col = "darkgreen") +
  geom_point(aes(x, y), data = data.frame(x = x_trains[3, ], y = y_trains[3, ]), col = "red")
```

The blue dataset is one possible $\mathbb{D}$, the green dataset is another possible $\mathbb{D}$ and the red dataset is another possible $\mathbb{D}$. This illustrated "dataset-dataset variability".

Take a look at the mse:

```{r}
mse = mean(c(all_oos_residuals)^2)
mse
```

This should be irreducible error plus bias-squared plus variance.

Let's visualize the bias

```{r}
resolution = 10000
g_average = colMeans(training_gs)
f = function(x){x^2}
x = seq(xmin, xmax, length.out = resolution)
f_x_df = data.frame(x = x, f = f(x))
ggplot(f_x_df, aes(x, f)) + 
  geom_line(col = "green") + 
  geom_abline(intercept = g_average[1], slope = g_average[2], col = "red")
```

What is the average bias of $g$?

```{r}
x = seq(xmin, xmax, length.out = resolution)
g_avg_x = g_average[1] + g_average[2] * x
f_x = x^2
biases = f_x - g_avg_x
expe_bias_g_sq = mean(biases^2)
expe_bias_g_sq
```

What is the variance? Let's look at all lines:

```{r}
plot_obj = ggplot() + 
  xlim(xmin, xmax) + ylim(xmin^2, xmax^2)

for (nsim in 1 : min(Nsim, 100)){ #otherwise takes too long
  plot_obj = plot_obj + geom_abline(intercept = training_gs[nsim, 1], slope = training_gs[nsim, 2], col = "blue")
}

plot_obj +
  geom_abline(intercept = g_average[1], slope = g_average[2], col = "red", lwd = 2) + 
  geom_line(data = f_x_df, aes(x, f), col = "green", size = 2)
```

Now calculate this average variance:

```{r}
x = seq(xmin, xmax, length.out = resolution)

expe_g_x = g_average[1] + g_average[2] * x

var_x_s = array(NA, Nsim)
for (nsim in 1 : Nsim){
  g_x = training_gs[nsim, 1] + training_gs[nsim, 2] * x
  var_x_s[nsim] = mean((g_x - expe_g_x)^2)
}

expe_var_g = mean(var_x_s)
expe_var_g
```

Now check the equivalence

```{r}
mse
sigma^2
expe_bias_g_sq
expe_var_g
sigma^2 + expe_bias_g_sq + expe_var_g
```

This is not exactly equal due to numerical error.

Let's try the whole thing again using a quadratic regression!

```{r}
training_gs = matrix(NA, nrow = Nsim, ncol = 3)
all_residuals = matrix(NA, nrow = Nsim, ncol = n_test)

for (nsim in 1 : Nsim){
  #simulate dataset $\mathbb{D}$
  x_train = runif(n_train, xmin, xmax)
  delta_train = rnorm(n_train, 0, sigma)
  y_train = x_train^2 + delta_train
  
  #fit a model g | x's, delta's and save it
  g_model = lm(y_train ~ poly(x, 2, raw = TRUE), data.frame(x = x_train))
  training_gs[nsim, ] = coef(g_model)
  
  #generate oos dataset and save residuals on oos data
  x_test = runif(n_test, xmin, xmax)
  delta_test = rnorm(n_test, 0, sigma)
  y_test = x_test^2 + delta_test
  y_hat_test = predict(g_model, data.frame(x = x_test))
  all_residuals[nsim, ] = y_test - y_hat_test
}
```

Take a look at the mse:

```{r}
mse = mean(c(all_residuals)^2)
mse
```

Much lower! Why? Bias went down. 

Let's visualize the bias

```{r}
g_average = colMeans(training_gs)
f = function(x){x^2}
x = seq(xmin, xmax, length.out = resolution)
ggplot(f_x_df, aes(x, f)) + 
  geom_line(col = "green") + 
  stat_function(fun = function(x){g_average[1] + g_average[2] * x + g_average[3] * x^2}, col = "red")
```

Not much! What is the average bias of $g$?

```{r}
x = seq(xmin, xmax, length.out = resolution)
g_avg_x = g_average[1] + g_average[2] * x + g_average[3] * x^2
f_x = x^2
biases = f_x - g_avg_x
expe_bias_g_sq = mean(biases^2)
expe_bias_g_sq
```

What is the variance? Let's look at all lines:

```{r}
plot_obj = ggplot(data.frame(x = x)) + 
  xlim(xmin, xmax) + ylim(xmin^2, xmax^2)

for (nsim in 1 : min(Nsim, 50)){ #otherwise takes too long
  plot_obj = plot_obj + geom_line(data = data.frame(x = x, y = training_gs[nsim, 1] + training_gs[nsim, 2] * x + training_gs[nsim, 3] * x^2), mapping = aes(x, y), col = "blue")
}

plot_obj +
  stat_function(fun = function(x){g_average[1] + g_average[2] * x + g_average[3] * x^2}, col = "red", lwd = 2) + 
  geom_line(data = f_x_df, aes(x, f), col = "green", size = 2)
```

Now calculate this average variance:

```{r}
x = seq(xmin, xmax, length.out = resolution)

expe_g_x = g_average[1] + g_average[2] * x + g_average[3] * x^2

var_x_s = array(NA, Nsim)
for (nsim in 1 : Nsim){
  g_x = training_gs[nsim, 1] + training_gs[nsim, 2] * x + training_gs[nsim, 3] * x^2
  var_x_s[nsim] = mean((g_x - expe_g_x)^2)
}

expe_var_g = mean(var_x_s)
expe_var_g
```

Now check the equivalence

```{r}
mse
sigma^2 
expe_bias_g_sq
expe_var_g
sigma^2 + expe_bias_g_sq + expe_var_g
```


Try it again with quintic polynomials!

```{r}
training_gs = matrix(NA, nrow = Nsim, ncol = 6)
all_residuals = matrix(NA, nrow = Nsim, ncol = n_test)

for (nsim in 1 : Nsim){
  #simulate dataset $\mathbb{D}$
  x_train = runif(n_train, xmin, xmax)
  delta_train = rnorm(n_train, 0, sigma)
  y_train = x_train^2 + delta_train
  
  #fit a model g | x's, delta's and save it
  g_model = lm(y_train ~ poly(x, 5, raw = TRUE), data.frame(x = x_train))
  training_gs[nsim, ] = coef(g_model)
  
  #generate oos dataset and save residuals on oos data
  x_test = runif(n_test, xmin, xmax)
  delta_test = rnorm(n_test, 0, sigma)
  y_test = x_test^2 + delta_test
  y_hat_test = predict(g_model, data.frame(x = x_test))
  all_residuals[nsim, ] = y_test - y_hat_test
}
```

Take a look at the mse:

```{r}
mse = mean(c(all_residuals)^2)
mse
```

Much higher! Why? Variance went up!

Let's visualize the bias

```{r}
g_average = colMeans(training_gs)
f = function(x){x^2}
x = seq(xmin, xmax, length.out = resolution)
ggplot(f_x_df, aes(x, f)) + 
  geom_line(col = "darkgreen") + 
  stat_function(fun = function(x){g_average[1] + g_average[2] * x + g_average[3] * x^2 + g_average[4] * x^3 + g_average[5] * x^4 + g_average[6] * x^5}, col = "red")
```

Not much! Now acutllay compute the average bias squared of $g$:

```{r}
x = seq(xmin, xmax, length.out = resolution)
g_avg_x = g_average[1] + g_average[2] * x + g_average[3] * x^2 + g_average[4] * x^3 + g_average[5] * x^4 + g_average[6] * x^5
f_x = x^2
biases = f_x - g_avg_x
expe_bias_g_sq = mean(biases^2)
expe_bias_g_sq
```

This appears to have increased over last time by a nominal amount ... but it's only because we're not running the regression infinite times. Remember this "expectation" is only an average.

What is the variance? Let's look at all lines:

```{r}
plot_obj = ggplot(data.frame(x = x)) + 
  xlim(xmin, xmax) + ylim(xmin^2, xmax^2)

for (nsim in 1 : min(Nsim, 30)){ #otherwise takes too long
  plot_obj = plot_obj + geom_line(data = data.frame(x = x, y = training_gs[nsim, 1] + training_gs[nsim, 2] * x + training_gs[nsim, 3] * x^2 + training_gs[nsim, 4] * x^3 + training_gs[nsim, 5] * x^4 + training_gs[nsim, 6] * x^5), mapping = aes(x, y), col = "blue")
}

plot_obj +
  stat_function(fun = function(x){g_average[1] + g_average[2] * x + g_average[3] * x^2 + g_average[4] * x^3 + g_average[5] * x^4 + g_average[6] * x^5}, col = "red", lwd = 2) + 
  ylim(-10, 50) +
  geom_line(data = f_x_df, aes(x, f), col = "green", size = 2)
```

It looks awful!!!

Now actually compute the average variance numerically:

```{r}
x = seq(xmin, xmax, length.out = resolution)

expe_g_x = g_average[1] + g_average[2] * x + g_average[3] * x^2 + g_average[4] * x^3 + g_average[5] * x^4 + g_average[6] * x^5

var_x_s = array(NA, Nsim)
for (nsim in 1 : Nsim){
  g_x = training_gs[nsim, 1] + training_gs[nsim, 2] * x + training_gs[nsim, 3] * x^2 + training_gs[nsim, 4] * x^3 + training_gs[nsim, 5] * x^4 + training_gs[nsim, 6] * x^5
  var_x_s[nsim] = mean((g_x - expe_g_x)^2)
}

expe_var_g = mean(var_x_s)
expe_var_g
```

Any more complexity than you need allows for overfitting!

Now check the equivalence

```{r}
mse
sigma^2
expe_bias_g_sq
expe_var_g
sigma^2 + expe_bias_g_sq + expe_var_g
```

# Bias - Variance Decomposition of MSE in Regression Trees

Let's return to the simulated sine curve data which we used to introduce regression trees.

```{r}
rm(list = ls())
n_train = 100
n_test = 500
xmin = 0
xmax = 10
sigma = 0.2
Nsim = 250
```

And load the tree package:

```{r}
options(java.parameters = "-Xmx4000m")
pacman::p_load(YARF, tidyverse, magrittr)
```

Now let's generate lots of different datasets and fit many tree models. Note there's a new argument `calculate_oob_error = FALSE`. This is here for speed only. Ignore this for now as we will go over what this means later in detail.

```{r}
training_gs = list() #storing entire objects - need a hash
all_residuals = matrix(NA, nrow = Nsim, ncol = n_test)

for (nsim in 1 : Nsim){
  #simulate dataset $\mathbb{D}$
  x_train = runif(n_train, xmin, xmax)
  delta_train = rnorm(n_train, 0, sigma)
  y_train = sin(x_train) + delta_train #f(x) = sin(x)
  # ggplot(data.frame(x = x, y = y), aes(x, y)) + geom_point(lwd=0.6) 
  
  #fit a model g | x's, delta's and save it
  g_model = YARFCART(data.frame(x = x_train), y_train, 
                     bootstrap_indices = 1 : n_train, calculate_oob_error = FALSE)
  training_gs[[nsim]] = g_model
  
  #generate oos dataset and save residuals on oos data
  x_test = runif(n_test, xmin, xmax)
  delta_test = rnorm(n_test, 0, sigma) #mean zero, variance sigsq always (not dependent on value of x)
  y_test = sin(x_test) + delta_test
  y_hat_test = predict(g_model, data.frame(x = x_test))
  all_residuals[nsim, ] = y_test - y_hat_test
}
```

Let's look at the last data set to remind ourselves of the problem setting:

```{r}
ggplot(data.frame(x = x_train, y = y_train)) + 
  geom_point(aes(x, y))
```


What does the storage of all the models look like?

```{r}
head(training_gs, 2)
```

Take a look at the mse:

```{r}
mse = mean(c(all_residuals)^2)
mse
```

Let's visualize the bias

```{r}
resolution = 1000
#generate x and the truth
x = seq(xmin, xmax, length.out = resolution)
f_x = sin(x)

#now estimate the expectation of g by averaging all the different models
g_avg_x = array(0, resolution)
for (nsim in 1 : Nsim){
  g_nsim = training_gs[[nsim]]
  g_avg_x = g_avg_x + predict(g_nsim, data.frame(x = x))
}
g_avg_x = g_avg_x / Nsim #average of all models

#now plot
ggplot(data.frame(x = x, f = f_x, expe_g = g_avg_x)) + 
  geom_line(aes(x, f), col = "darkgreen") + 
  geom_line(aes(x, expe_g), col = "red")
```

Not much! Now actually compute the average bias squared of $g$:

```{r}
biases = f_x - g_avg_x
expe_bias_g_sq = mean(biases^2)
expe_bias_g_sq
```

This is small - why??

It's because trees are so expressive and have such model complexity that they can nail almost any true $f$ function!

That means the MSE save the irreducible noise is coming from the variance. Let's look at the variance vs f(x):

```{r}
plot_obj = ggplot(data.frame(x = x)) + 
  xlim(xmin, xmax) #+ ylim(xmin^2, xmax^2)

num_trees_to_visualize = 30
for (nsim in 1 : min(Nsim, num_trees_to_visualize)){
  g_nsim = training_gs[[nsim]]
  g_x = predict(g_nsim, data.frame(x = x))
  plot_obj = plot_obj + geom_line(data = data.frame(x = x, y = g_x), aes(x = x, y = y), col = "blue")
}

plot_obj +
  geom_line(data = data.frame(x = x, expe_g = g_avg_x), mapping = aes(x, expe_g), col = "red", lwd = 1) #note: expe_g = f(x) since no bias.... but technically I should be using a separate f(x) data frame
```

It looks awful!!!

Now actually compute the average variance numerically:

```{r}
x = seq(xmin, xmax, length.out = resolution)

var_x_s = array(NA, min(Nsim, 50))
for (nsim in 1 : min(Nsim, 50)){ #otherwise takes too long
  g_nsim = training_gs[[nsim]]
  g_x = predict(g_nsim, data.frame(x = x))
  var_x_s[nsim] = mean((g_x - g_avg_x)^2)
}

expe_var_g = mean(var_x_s)
expe_var_g
```

Any more complexity than you need allows for overfitting!

Now check the equivalence

```{r}
mse
sigma^2
expe_bias_g_sq
expe_var_g
sigma^2 + expe_bias_g_sq + expe_var_g
```

# Model Averaging via Bootstrap Aggregation ("Bagging")

Now let's generate one dataset (only ONE)!

```{r}
x_train = runif(n_train, xmin, xmax)
delta_train = rnorm(n_train, 0, sigma)
y_train = sin(x_train) + delta_train
ggplot(data.frame(x = x_train, y = y_train)) + geom_point(aes(x, y))
```

Spend a moment to appreciate that this is all we're going to have. There is a lot of irreducible noise that is going to prevent us from finding the sine curve.

Now we create many bootstrap samples to create similiar but different datasets and fit many tree models:

```{r}
bootstrapped_gs = list() #storing entire objects - need a hash

num_trees = 250
for (t in 1 : num_trees){ #M = num_trees
  #fit a model g | x's, delta's and save it
  bootstrap_indices_t = sample(1 : n_train, replace = TRUE)
  g_model = YARFCART(data.frame(x = x_train[bootstrap_indices_t]), y_train[bootstrap_indices_t], 
                     calculate_oob_error = FALSE, bootstrap_indices = 1 : n_train) #this is confusing! But will become clear later...
  bootstrapped_gs[[t]] = g_model
}
```

Now let's aggregate all models constructed with bootstrap samples together by averaging and see what it looks like relative to real $f$:

```{r}
#generate x and the truth
x = seq(xmin, xmax, length.out = resolution)
f_x = sin(x)

#create the bagged model predictions
g_bagged = array(0, resolution)
for (t in 1 : num_trees){
  g_t = bootstrapped_gs[[t]]
  g_bagged = g_bagged + predict(g_t, data.frame(x = x))
}
g_bagged = g_bagged / num_trees #average of all models

#now plot
ggplot(data.frame(x = x, f = f_x, g_bagged = g_bagged)) + 
  geom_line(aes(x, f), col = "green") + 
  geom_line(aes(x, g_bagged), col = "blue")
```
  
That's pretty good considering the plot of the raw data from above.

We just compared one tree to a bagged-tree (i.e. average of many trees where each tree was fit on a bootstrap sample of the original data). Let's see if the bagged tree is truly better than one tree. The way to do this is generate lots of datasets and try to estimate mse (with bias and variance).

We fit the bag of trees before manually with a for loop. But of course there is a convenience to do this, the function `YARFBAG`. And then we run the same simulation as we did above with one tree.

```{r}
Nsim = 250
training_gs = list() #storing entire objects - need a hash
all_residuals = matrix(NA, nrow = Nsim, ncol = n_test)

for (nsim in 1 : Nsim){
  #simulate dataset $\mathbb{D}$
  x_train = runif(n_train, xmin, xmax)
  delta_train = rnorm(n_train, 0, sigma)
  y_train = sin(x_train) + delta_train
  # ggplot(data.frame(x = x, y = y), aes(x, y)) + geom_point(lwd=0.6) 
  
  #fit a model bagged tree g | x's, delta's and save it
  g_model = YARFBAG(data.frame(x = x_train), y_train, num_trees = num_trees, calculate_oob_error = FALSE) #automatically does the bootstrap internally
  training_gs[[nsim]] = g_model
  
  #generate oos dataset and save residuals on oos data
  x_test = runif(n_test, xmin, xmax)
  delta_test = rnorm(n_test, 0, sigma) #mean zero, variance sigsq always (not dependent on value of x)
  y_test = sin(x_test) + delta_test
  y_hat_test = predict(g_model, data.frame(x = x_test))
  all_residuals[nsim, ] = y_test - y_hat_test
}
```

What do all the the bagged models look like?

```{r}
head(training_gs, 2)
```

Take a look at the mse:

```{r}
mse = mean(c(all_residuals)^2)
mse
```

Let's visualize the bias

```{r}
resolution = 1000
#generate x and the truth
x = seq(xmin, xmax, length.out = resolution)
f_x = sin(x)

#now estimate the expectation of g by averaging all the different models
g_avg_x = array(0, resolution)
for (nsim in 1 : Nsim){
  g_nsim = training_gs[[nsim]]
  g_avg_x = g_avg_x + predict(g_nsim, data.frame(x = x))
}
g_avg_x = g_avg_x / Nsim #average of all models

#now plot
ggplot(data.frame(x = x, f = f_x, expe_g = g_avg_x)) + 
  geom_line(aes(x, f), col = "darkgreen") + 
  geom_line(aes(x, expe_g), col = "red")
```

If a single tree was unbiased then an average of trees will be unbiased.

Now actually compute the average bias squared of $g$:

```{r}
biases = f_x - g_avg_x
expe_bias_g_sq = mean(biases^2)
expe_bias_g_sq
```

Almost zilch.

Now... variance?

```{r}
plot_obj = ggplot(data.frame(x = x)) + 
  xlim(xmin, xmax) #+ ylim(xmin^2, xmax^2)

for (nsim in 1 : min(Nsim, 30)){ #otherwise takes too long
  g_nsim = training_gs[[nsim]]
  g_x = predict(g_nsim, data.frame(x = x))
  plot_obj = plot_obj + geom_line(data = data.frame(x = x, y = g_x), aes(x = x, y = y), col = "blue")
}

plot_obj +
  geom_line(data = data.frame(x = x, expe_g = g_avg_x), mapping = aes(x, expe_g), col = "red", lwd = 1.5)
```

There is variance, but less than previously when only one tree was used.

Now actually compute the average variance numerically:

```{r}
x = seq(xmin, xmax, length.out = resolution)

var_x_s = array(NA, min(Nsim, 50))
for (nsim in 1 : min(Nsim, 50)){ #otherwise takes too long
  g_nsim = training_gs[[nsim]]
  g_x = predict(g_nsim, data.frame(x = x))
  var_x_s[nsim] = mean((g_x - g_avg_x)^2)
}

expe_var_g_bagged = mean(var_x_s)
expe_var_g_bagged
```

Any more complexity than you need allows for overfitting!

Now check the equivalence

```{r}
mse
sigma^2
expe_bias_g_sq
expe_var_g_bagged
sigma^2 + expe_bias_g_sq + expe_var_g_bagged
```

We have a better algorithm! Why? Because there is reduction of correlation between the bootstrapped models:

```{r}
rho = expe_var_g_bagged / expe_var_g
rho
```

That's a gain of almost 50% in the MSE component that we have control over! And... it's for FREEEEEE.

# Bagged Trees vs. a Linear Model: Behold the power of machine learning!!

Let's look at the boston housing data first.

```{r}
boston = MASS::Boston

prop_test = 0.1
test_indices = sample(1 : nrow(boston), round(prop_test * nrow(boston)))
boston_test = boston[test_indices, ]
y_test = boston_test$medv
X_test = boston_test
X_test$medv = NULL
train_indices = setdiff(1 : nrow(boston), test_indices)
boston_train = boston[train_indices, ]
y_train = boston_train$medv
X_train = boston_train
X_train$medv = NULL

mod_lin = lm(y_train ~ ., X_train)
y_hat_test_lin = predict(mod_lin, X_test)
s_e_lin = sd(y_test - y_hat_test_lin)
s_e_lin

num_trees = 500
mod_bag = YARFBAG(X_train, y_train, num_trees = num_trees, calculate_oob_error = FALSE)
y_hat_test_bag = predict(mod_bag, X_test)
s_e_bag = sd(y_test - y_hat_test_bag)
s_e_bag

cat("oos standard error reduction:", (1 - s_e_bag / s_e_lin) * 100, "%\n")
```


Whole lot of room to improve! That extra 26% or so of unexplained variance is split between bias and irreducible error due to unknown information. The bagged trees can get rid of the bias and minimize variance while doing so. And it did a great job!

Now the diamonds data:

```{r}
n_train = 1000

training_indices = sample(1 : nrow(diamonds), n_train)
diamonds_train = diamonds[training_indices, ]
y_train = diamonds_train$price
X_train = diamonds_train
X_train$price = NULL


test_indices = setdiff(1 : nrow(diamonds), training_indices)
diamonds_test = diamonds[test_indices, ]
y_test = diamonds_test$price
X_test = diamonds_test
X_test$price = NULL


mod_lin = lm(y_train ~ ., X_train)
y_hat_test_lin = predict(mod_lin, X_test)
s_e_lin = sd(y_test - y_hat_test_lin)
s_e_lin

mod_bag = YARFBAG(X_train, y_train, num_trees = num_trees, calculate_oob_error = FALSE)
y_hat_test_bag = predict(mod_bag, X_test)
s_e_bag = sd(y_test - y_hat_test_bag)
s_e_bag

cat("oos standard error reduction:", (1 - s_e_bag / s_e_lin) * 100, "%\n")
```

```{r}
summary(mod_lin)$r.squared
```

Seemingly not a whole lot of room to improve! But there is when you measure error in RMSE. That extra 8% of unexplained variance means a whole lot of RMSE and it is split between bias and irreducible error due to unknown information. But it did the best it can. It is likely what's remaining is due to information we don't measure.








