---
title: "Practice Lecture 21 MATH 342W Queens College"
author: "Professor Adam Kapelner"
date: "April 21, 2021"
---

# Joins

Another one of the core data munging skills is joining data frames together. In the real world, databases consist of multiple dataframes called "tables" and design matrices are built by gathering data from among many tables. To illustrate this, we load two datasets from the package `nycflights13`, one dataset about weather and one about airports:

```{r}
pacman::p_load(nycflights13, data.table, tidyverse, magrittr)
data(weather)
summary(weather)
data(airports)
summary(airports)

head(weather)
head(airports)
```

Note how the weather and airports datasets contain a common feature: name of airport. It is called `FAA` in airports and `origin` in weather.

First we rename the column in weather to match the column in airports:

```{r}
weather %<>% 
  rename(faa = origin)
```

We also pare down the datasets so we can see the joins more clearly:

```{r}
airports %<>% 
  select(faa, lat, lon)
weather %<>% 
  select(faa, time_hour, temp, humid, wind_speed, pressure, wind_gust)
head(airports)
head(weather)
airports_dt = data.table(airports)
weather_dt = data.table(weather)
```

Some features just aren't measured that often e.g. `wind_gust`.

Let's do some joins. First "left". This is likely the most common because it's usually how we conceptualize what we're doing in our heads.

```{r}
airports_and_weather = left_join(airports, weather, by = "faa")
airports_and_weather %>% sample_n(500)


airports_and_weather_dt = merge(airports_dt, weather_dt, by = "faa", all.x = TRUE)
airports_and_weather_dt = merge(airports_dt, weather_dt, all.x = TRUE) #note this works too since it knows faa is the only column in common but not recommended since specifying "by" is more clear
airports_and_weather_dt[sample(1 : .N, 500)]
```

Now "right"

```{r}
airports_and_weather = right_join(airports, weather, by = "faa")
airports_and_weather %>% sample_n(500)
airports_and_weather_dt = merge(airports_dt, weather_dt, by = "faa", all.y = TRUE)
airports_and_weather_dt = merge(airports_dt, weather_dt, all.y = TRUE)
airports_and_weather_dt[sample(1 : .N, 500)]
```


```{r}
airports_and_weather = inner_join(airports, weather, by = "faa")
airports_and_weather %>% sample_n(500)
airports_and_weather_dt = merge(airports_dt, weather_dt, by = "faa")
airports_and_weather_dt = merge(airports_dt, weather_dt)
airports_and_weather_dt[sample(1 : .N, 500)]
```

And full, keeping all the rows. We use a subset to show how this works:

```{r}
airports_without_EWR = airports %>%
  filter(faa != "EWR")
airports_without_EWR_dt = data.table(airports_without_EWR)
airports_without_EWR_and_weather = full_join(airports_without_EWR, weather, by = "faa")
airports_without_EWR_and_weather %>% sample_n(500)
airports_without_EWR_and_weather_dt = merge(airports_without_EWR_dt, weather_dt, by = "faa", all = TRUE)
airports_without_EWR_and_weather_dt = merge(airports_without_EWR_dt, weather_dt, all = TRUE)
airports_without_EWR_and_weather_dt[sample(.N, 500)]
```

There is also `semi_join` and `anti_join` that do the opposite of joining. In my experience, these use cases are limited.


# Using Probability Estimation to do Classification

First repeat quickly (a) load the adult data (b) do a training / test split and (c) build the logisitc model.

```{r}
pacman::p_load_gh("coatless/ucidata")
data(adult)
adult = na.omit(adult) #kill any observations with missingness

set.seed(1)
train_size = 5000
train_indices = sample(1 : nrow(adult), train_size)
adult_train = adult[train_indices, ]
y_train = adult_train$income
X_train = adult_train
X_train$income = NULL

test_size = 5000
test_indices = sample(setdiff(1 : nrow(adult), train_indices), test_size)
adult_test = adult[test_indices, ]
y_test = adult_test$income
X_test = adult_test
X_test$income = NULL

logistic_mod = glm(income ~ ., adult_train, family = "binomial")
p_hats_train = predict(logistic_mod, adult_train, type = "response")
p_hats_test = predict(logistic_mod, adult_test, type = "response")
```

Let's establish a rule: if the probability estimate is greater than or equal to 50%, let's classify the observation as positive, otherwise 0.

```{r}
y_hats_train = factor(ifelse(p_hats_train >= 0.5, ">50K", "<=50K"))
```

How did this "classifier" do in-sample?

```{r}
mean(y_hats_train != y_train)
table(y_train, y_hats_train)
```

Let's see the same thing oos:

```{r}
y_hats_test = factor(ifelse(p_hats_test >= 0.5, ">50K", "<=50K"))
mean(y_hats_test != y_test)
oos_conf_table = table(y_test, y_hats_test)
oos_conf_table
```

A tad bit worse. Here are estimates of the future performance for each class:

```{r}
n = sum(oos_conf_table)
fp = oos_conf_table[1, 2]
fn = oos_conf_table[2, 1]
tp = oos_conf_table[2, 2]
tn = oos_conf_table[1, 1]
num_pred_pos = sum(oos_conf_table[, 2])
num_pred_neg = sum(oos_conf_table[, 1])
num_pos = sum(oos_conf_table[2, ])
num_neg = sum(oos_conf_table[1, ])
precision = tp / num_pred_pos
cat("precision", round(precision * 100, 2), "%\n")
recall = tp / num_pos
cat("recall", round(recall * 100, 2), "%\n")
false_discovery_rate = 1 - precision
cat("false_discovery_rate", round(false_discovery_rate * 100, 2), "%\n")
false_omission_rate = fn / num_pred_neg
cat("false_omission_rate", round(false_omission_rate * 100, 2), "%\n")
```

This whole classifier hinged on the decision of 50%! What if we change it??

# Asymmetric Cost Classifiers

Let's establish a *new* rule: if the probability estimate is greater than or equal to 90%, let's classify the observation as positive, otherwise 0.

```{r}
y_hats_test = factor(ifelse(p_hats_test >= 0.9, ">50K", "<=50K"))
mean(y_hats_test != y_test)
oos_conf_table = table(y_test, y_hats_test)
oos_conf_table
```

Of course the misclassification error went up! But now look at the confusion table! The second column represents all $\hat{y} = 1$ and there's not too many of them! Why? You've made it *much* harder to classify something as positive. Here's the new additional performance metrics now:

```{r}
n = sum(oos_conf_table)
fp = oos_conf_table[1, 2]
fn = oos_conf_table[2, 1]
tp = oos_conf_table[2, 2]
tn = oos_conf_table[1, 1]
num_pred_pos = sum(oos_conf_table[, 2])
num_pred_neg = sum(oos_conf_table[, 1])
num_pos = sum(oos_conf_table[2, ])
num_neg = sum(oos_conf_table[1, ])
precision = tp / num_pred_pos
cat("precision", round(precision * 100, 2), "%\n")
recall = tp / num_pos
cat("recall", round(recall * 100, 2), "%\n")
false_discovery_rate = 1 - precision
cat("false_discovery_rate", round(false_discovery_rate * 100, 2), "%\n")
false_omission_rate = fn / num_pred_neg
cat("false_omission_rate", round(false_omission_rate * 100, 2), "%\n")
```

We don't make many false discoveries but we make a lot of false omissions! It's a tradeoff...
