\documentclass[12pt]{article}

\include{preamble}

\newtoggle{professormode}
\toggletrue{professormode} %STUDENTS: DELETE or COMMENT this line



\title{MATH 390.4 / 650.2 Spring 2020 \\ Philosophy of Modeling Draft and Final Paper}

\author{Professor Adam Kapelner} %STUDENTS: write your name here

\iftoggle{professormode}{
\date{Draft Due 6PM Monday KY604 by email, April 6, 2020 \\ \vspace{0.5cm} \small (this document last updated \currenttime~on \today)}
}


\begin{document}
\maketitle

\noindent Pick one of the prompts and argue for one of the choices. 

\begin{enumerate}[(a)]
\item The laws of physics are absolutely true / \\
The laws of physics are \textit{not} absolutely true / \\
The laws of physics are \textit{perhaps} absolutely true.
\item There exists an all-powerful, all-knowing and timeless being or intelligence / \\
There does \textit{not} exist an all-powerful, all-knowing and timeless being or intelligence / \\
There \textit{perhaps} exists an all-powerful, all-knowing and timeless being or intelligence.
\item Global warming is real / \\
Global warming is \textit{not} real / \\
Global warming is \textit{perhaps} real.
\item The optimal human diet for health is [insert a diet here] / \\
The optimal human diet for health is \textit{not} [insert a diet here] / \\
The optimal human diet for health is \textit{perhaps} [insert a diet here].
\item People who smoke contract lung cancer / \\
People who smoke do \textit{not} contract lung cancer / \\
People who smoke \textit{perhaps} contract lung cancer.
\item Autonomous vehicles will navigate perfectly / \\
Autonomous vehicles will \textit{not} navigate perfectly / \\
Autonomous vehicles will \textit{perhaps} navigate perfectly.
\item The government understands the optimal level of taxation / \\
The government does \textit{not} understand the optimal level of taxation / \\
The government \textit{perhaps} understands the optimal level of taxation.
\item The medical establishment understand the human body / \\
The medical establishment does \textit{not} understand the human body / \\
The medical establishment \textit{perhaps} understands the human body.
\item Credit scores like those given by Experian, TransUnion and Equifax are useful / \\
Credit scores like those given by Experian, etc. are \textit{useless} / \\
Credit scores like those given by Experian, etc. are \textit{perhaps} useful.
\item What makes a marriage successful is understood / \\
What makes a marriage successful is \emph{not} understood / \\
What makes a marriage successful is \emph{somewhat} understood.
\item What makes a virus a dangerous pandemic is understood / \\
What makes a virus a dangerous pandemic is \emph{not} understood / \\
What makes a virus a dangerous pandemic is \emph{somewhat} understood.
\end{enumerate}

\noindent It is your job to interpret the above prompts and redefine them in your own words, i.e. you must \textit{limit}, \textit{scope} and \textit{concretize} the above which are \textit{deliberately open-ended} and nebulous. To argue your point, you will need to formulate a mathematical model with phenomenon(s) that can be measured (and you must explain how you measure / sample in the real world) and characteristics of the units under consideration. You will need to clearly define what \qu{models} are, how it could be \textbf{useful} in your context, how your model is mathematical and discuss the limitations of the modeling enterprise in general. 

In context of your prompt and modeling choice, you must appropriately explain the concepts we learned / will learn in class, including but not limited to $t ,f, g, g_0, h^*, \delta, \mathcal{E}, e, \mathbb{D}, \mathcal{H}, \mathcal{A}$, $t, z_1, \ldots, z_t, n, p$, $X$, $x_{\cdot 1}, \ldots, x_{\cdot p}$, $x_{1 \cdot}, \ldots, x_{n \cdot}, \mathcal{X}, y, \mathcal{Y}$, supervised learning, the three sources of error, the fidelity of the measurement process, extrapolation, interpolation, stationarity, overfitting, validation (in-sample vs. out-of-sample), generalization error, model selection, machine learning, etc. 

You are welcome to bring outside sources about philosophy of modeling as well as sources which help make your arguments in support of a prompt. Please cite them appropriately using APA style e.g. (Johnson et al., 1999) and enter them into a bibilography where the reference elements are formatted in APA style.\\
~\\
\textbf{Specs}: Your essay must be typed and must be at least 10 pages double-spaced with one inch margin, 12pt Times (or Computer Modern if using \LaTeX) and be appropriately organized. You choose a title which will be atop the first page. No need for a title page. Sectioning is at your preference. But, one of the choices in one of the prompts must be a powerful sentence in either the beginning or the end of the essay. The bibliography does not count towards the page limit. Keep footnotes to a minimum and do not use endnotes. You must email me a  \inred{PDF} of your paper. Microsoft Word allows saving as PDF, so this is no problem.

\subsection*{Revision (Due 11:59PM Friday, April 24 by email)}

You will receive comments on your draft after your submission. You will have the opportunity to revise your draft once and resubmit as a PDF. Resubmit by responding to the email with my comments and attaching your revision. You will receive a grade for this assignment assessed by the performance on your revision (not your initial submission).

\pagenumbering{gobble}

\end{document}