\documentclass[12pt]{article}

\usepackage[margin=1.0in]{geometry}
\input{../../syllabi/preamble}

\newcommand{\coursedept}{Math}
\newcommand{\coursenumber}{342W}
\newcommand{\coursenumbercrosslisted}{/ 650.4~}
\newcommand{\semester}{Spring}
\newcommand{\numcredits}{6}
\newcommand{\lectimeandloc}{Mon and Wed 5-6:50PM / on zoom}
\newcommand{\requiredlabtimeandloc}{Required Lab Time / Loc 		& Thurs 9-10:50AM / on zoom \\}
\newcommand{\tataofficehourtimeandloc}{TA 	/ TA Office Hours / Loc 			& Tzipora Horowitz / Wed 6:55-7:55PM / on zoom}
\newcommand{\coursewebpageurl}{https://github.com/kapelner/QC_\coursedept_\coursenumber_\semester_\the\year}
\newcommand{\coursewebpagelink}{\href{\coursewebpageurl}{course homepage}}
\newcommand{\slackurl}{https://QC\coursedept\coursenumber\semester\the\year.slack.com/}
\newcommand{\slacklink}{\href{\slackurl}{slack}}
\newcommand{\numtheoryhws}{4--7}
\newcommand{\extrahwzero}{\item provide a link to your public repository on github (this means you need to sign up for github first)}
\newcommand{\hwzerodue}{Wednesday, Feb 3 11:59PM}
\newcommand{\lastdatetimetohandinhomeworks}{May 18 at noon}

\input{../../syllabi/_header}

\section*{Course Overview}

MATH 342W. Data Science via Machine Learning and Statistical Modeling. 6 hr.; 4 cr. Prereq.: MATH 241 (intro to probablity and statistics), MATH 231 (intro to linear algebra), CSCI 111 (intro to programming) or equivalents. Recommended: ECON 382 (intro to economentrics) or equivalent. Philosophy of modeling and learning using data. Prediction via the ordinary linear model including orthogonal projections, sum of squares identity, $R^2$ and RMSE. Polynomial and interaction regressions. Prediction with machine learning including neural nets (the perceptron), support vector machines and the tree methods CART, bagged trees and Random Forests. Probability estimation using logistic regression, asymmetric cost classifiers and the ROC / DET performance curves. Underfitting vs. overfitting and the bias-variance decomposition / tradeoff. Model validation including out of sample techniques such as cross validation and bootstrap validation. Correlation vs. causation, causal models, lurking variables and interpretations of linear model coefficients. Extrapolation. The \texttt{R} language will be taught formally from the ground and up (its use will be a substantial part of the homework) as well as visualization using the \texttt{ggplot} library and manipulation using the \texttt{dplyr} and \texttt{data.table} libraries. \pagebreak

You should be familiar with the following before entering the class:

\begin{itemize}
\itemsep -0.0em 
\item Basic Probability Theory: conditional probability, in/dependence, identical distributedness
\item Modeling with discrete random variables: Bernoulli, Binomial%, Poisson, Geometric, Negative Binomial, Uniform Discrete and others
\item Expectation and variance
%\item Modeling with continuous random variables: Exponential, Uniform and Normal
%\item Frequentist confidence intervals and hypothesis testing for one-sample proportions
%\item Basic visualization of data: plots, histograms, bar charts
\item Linear algebra: Vectors, matrices, transpose, inverse, rank, nullity
\item Programming: basic data types, vectors, arrays, control flow (for, while, if, else), functions
\end{itemize}

\noindent We will review the above \textit{throughout the semester} when needed and we will do so rapidly. \\

\textbf{This is not your typical mathematics course.} This course will do lots of modeling of real-world situations using data via the \texttt{R} statistical language.



\section*{Course Materials}

We will be using many reference texts and three popular books which you will read portions from. However the main materials are the course notes. You should always supplement concepts from class by reading up on them online; \href{https://en.wikipedia.org}{wikipedia} I find the best for this. 

\paragraph{Theory Reference:} It is not necessary to have these two books, but it is recommended. The first is \qu{Learning from Data: A Short Course} by Abu-Mostafa, Magdon-Ismael and Lin which can be purchased used on \href{https://www.amazon.com/Learning-Data-Yaser-S-Abu-Mostafa/dp/1600490069}{Amazon}. We will also be using portions from \qu{Deep Learning} by Goodfellow, Bengio and Courville that can be purchased on \href{https://www.amazon.com/Deep-Learning-Adaptive-Computation-Machine/dp/0262035618}{Amazon} and read for free at \url{http://www.deeplearningbook.org/}.

\paragraph{Popular Books:} We will also be reading the non-fiction novel \qu{The Signal and the Noise} by Nate Silver which can also be purchased on \href{https://www.amazon.com/Signal-Noise-Many-Predictions-Fail-but/dp/0143125087}{Amazon}. This is \textit{required} --- you will have homework questions directly from this book. We will also be reading \qu{Preditive Analytics, Data Mining and Big Data} by Steven Finlay that can be purchased on \href{https://www.amazon.com/Predictive-Analytics-Data-Mining-Misconceptions/dp/1349478687}{Amazon} and it is also available online from the \href{https://link-springer-com.queens.ezproxy.cuny.edu/book/10.1057%2F9781137379283}{Queens College library system}. 

\paragraph{Computer Software:} You need your own personal computer, laptop preferred. We will be using \texttt{R} which is a free, open source statistical programming language and console available for all operating systems. Please download the latest version from: \url{http://cran.mirrors.hoobly.com/}. You will be expected to do programming. I recommend the IDE \texttt{RStudio} available for free at \url{https://www.rstudio.com/products/rstudio/download/}.

\paragraph{Source Control:} You will be expected to use \texttt{git} and have a \url{github.com} account with a repository named \texttt{QC\_MATH\_342}. You will use this repository to submit coding homework assignments (and theory assignments if you use \LaTeX).


\paragraph{Book on \texttt{R}:} We will be making some use of \qu{R for Data Science} by Wickham and Grolemund which can be purchased on \href{https://www.amazon.com/R-Data-Science-Hadley-Wickham/dp/1491910399}{Amazon} or read online at \url{http://r4ds.had.co.nz/}.


\input{../../syllabi/_the650section}

\input{../../syllabi/_announcements_on_slack}

\input{../../syllabi/_use_of_slack}

\section*{Class Meetings}

There are 42 scheduled meetings. Of these, 26 will be lectures, 10 will be labs, 2 will be midterm exams (see the schedule on page~\pageref{subsec:exam_schedule}) which are in class and 2 will be review periods before the exams, 1 will be a review lab and 1 will be help with the final paper. I am \inred{canceling} Monday, May 17 (the last meeting) due to a Jewish holiday. This meeting would have been for help on the final paper. We will decide when to have this session during finals week that fits the majority's schedule.

\subsection*{Lectures}

Lectures will be on zoom and will be split usually into two periods: theory and practice. The first is a standard \qu{chalkboard} lecture where we learn concepts and the second will be using the \qu{computer/projector} to see the concepts in action in the \texttt{R} language. %I have a no computer / tablet / phone policy during the theory component of the lectures (only pen / pencil and paper) but you are highly recommended to have the laptop during the second part.

\input{../../syllabi/_zoom_policies}

\input{_lecture_schedule}

\input{../../syllabi/_lecture_upload}


\subsection*{Labs}

Labs  will be on zoom during the Thursday morning meetings. Sometimes we will spend some of the two hours doing a practice lecture but the majority of this time will be your time. You will take turns \qu{driving} the coding in front of the class, working on exercises that you will finish for homework. Thus we will spend most a lot of time talking through problem solving skills in data science.

\section*{Homework}

Homework will be split into \textit{theory} and \textit{practice} (called \qu{labs}). This course will be the \qu{writing in the major course} next year. Thus, a portion of each theory and practice homework will involve writing \textit{English} and being graded on \textit{English}.

\subsection*{Theory Homework}

\input{../../syllabi/_theory_hws_text}
\input{../../syllabi/_theory_hws_submission_text}



\subsection*{Practice Homework (Labs)}

These will almost exclusively consist of short and medium coding exercises in \texttt{R}. Most of the assignment will be done for you and your peers during the Friday lab session.


\input{../../syllabi/_philosophy_hws}

\input{../../syllabi/_time_spent_hws}

\input{../../syllabi/_late_hw_policy}

\input{../../syllabi/_latex_hw_bonus_policy}

\input{../../syllabi/_hw_ec_policy}

\input{../../syllabi/_hw_0}

\section*{Writing Assignments}

There will be two writing assingments. (1) A \qu{philosophy of modeling} essay. Here you will coalesce the non-mathematical material that is crucial to this class. The purpose is to make you truly understand the modeling process and its limitations from start to finish. (2) A final project. Here you will use build a predictive model using a dataset. This is the capstone project for the entire data science and statistics major and it is where you will tie everything together.

This class will soon be the writing in the major course. Thus, writing is a major part of the curriculum herein.

\section*{Examinations}

\input{../../syllabi/_examination_text}

Since the is the capstone course, there is no final exam, but a large final project. There will be two midterm exams and the schedule is below.

\subsection*{Exam and Major Assignment Schedule}\label{subsec:exam_schedule}

\begin{itemize}
\itemsep -0.0em 
\item Midterm examination I will be Thurs, March 25 in class with the first review session on the Wednesday prior
\item The philosophy of modeling paper's first draft is due Sunday, Mar 21 at 11:59PM
\item Midterm examination II will be Thurs, May 13 in class with a review on the Wednesday prior
\item The final project is due Sunday, May 23 11:59PM
\end{itemize}

\subsection*{Exam Policies and Materials}

\input{../../syllabi/_examination_policies}


I also allow \qu{cheat sheets} on examinations. For both midterms, you are allowed to bring \ingreen{two} 8.5'' $\times$ 11'' sheet of paper (front and back). \inred{Four sheets single sided are not allowed.} On this paper you can write anything you would like which you believe will help you on the exam. %For the final, you are allowed to bring three 8.5'' $\times$ 11'' sheet of paper (front and back). \inred{Six sheets single sided are not allowed.} I will be handing back the cheat sheets so you can reuse your midterm cheat sheets for the final if you wish. 




\input{../../syllabi/_cheating_on_exams_and_missing_exams}
\input{../../syllabi/_special_services}

\input{../../syllabi/_class_participation}

\input{../../syllabi/_zoom_attendance}


\section*{Grading and Grading Policy}\label{sec:grading}

Your course grade will be calculated based on the percentages as follows: 

\begin{table}[h]
\centering
\begin{tabular}{l|l}
Theory Homework & 9\% \\
Labs & 14\% \\
Midterm Examination I & 18\%\\
Midterm Examination II* & 18\%\\
Philosophy of Modeling Paper & 9\% \\
Final Project with Writeup & 22\% \\
Class participation & 5\% \\
Attendance & 5\% 
\end{tabular}
\end{table}
\FloatBarrier

\noindent *The second midterm is not cumulative. It only covers material \textit{after} midterm I.


\input{../../syllabi/_advanced_course_grade_distribution}

\input{../../syllabi/_grade_checking_on_gradesly}

\input{../../syllabi/_auditing_policy}




\end{document}
