---
title: "Lab 4"
author: "Your Name Here"
output: pdf_document
date: "11:59PM March 10, 2021"
---

Load up the famous iris dataset. We are going to do a different prediction problem. Imagine the only input x is Species and you are trying to predict y which is Petal.Length. A reasonable prediction is the average petal length within each Species. Prove that this is the OLS model by fitting an appropriate `lm` and then using the predict function to verify.

```{r}
#TO-DO
```

Construct the design matrix with an intercept, $X$, without using `model.matrix`.

```{r}
#TO-DO
```

Find the hat matrix $H$ for this regression.

```{r}
#TO-DO
```

Verify this hat matrix is symmetric using the `expect_equal` function in the package `testthat`.

```{r}
#TO-DO
```

Verify this hat matrix is idempotent using the `expect_equal` function in the package `testthat`.

```{r}
#TO-DO
```

Using the `diag` function, find the trace of the hat matrix.

```{r}
#TO-DO
```

It turns out the trace of a hat matrix is the same as its rank! But we don't have time to prove these interesting and useful facts..

For masters students: create a matrix $X_\perp$.

```{r}
#TO-DO
```

Using the hat matrix, compute the $\hat{y}$ vector and using the projection onto the residual space, compute the $e$ vector and verify they are orthogonal to each other.

```{r}
#TO-DO
```

Compute SST, SSR and SSE and $R^2$ and then show that SST = SSR + SSE.

```{r}
#TO-DO
```

Find the angle $\theta$ between $y$ - $\bar{y}1$ and $\hat{y} - \bar{y}1$ and then verify that its cosine squared is the same as the $R^2$ from the previous problem.

```{r}
#TO-DO
```

Project the $y$ vector onto each column of the $X$ matrix and test if the sum of these projections is the same as yhat.

```{r}
#TO-DO
```

Construct the design matrix without an intercept, $X$, without using `model.matrix`.

```{r}
#TO-DO
```

Find the OLS estimates using this design matrix. It should be the sample averages of the petal lengths within species.

```{r}
#TO-DO
```

Verify the hat matrix constructed from this design matrix is the same as the hat matrix constructed from the design matrix with the intercept. (Fact: orthogonal projection matrices are unique).

```{r}
#TO-DO
```

Project the $y$ vector onto each column of the $X$ matrix and test if the sum of these projections is the same as yhat.

```{r}
#TO-DO
```

Convert this design matrix into $Q$, an orthonormal matrix.

```{r}
#TO-DO
```

Project the $y$ vector onto each column of the $Q$ matrix and test if the sum of these projections is the same as yhat.

```{r}
#TO-DO
```

Find the $p=3$ linear OLS estimates if $Q$ is used as the design matrix using the `lm` method. Is the OLS solution the same as the OLS solution for $X$?

```{r}
#TO-DO
```

Use the predict function and ensure that the predicted values are the same for both linear models: the one created with $X$ as its design matrix and the one created with $Q$ as its design matrix.

```{r}
#TO-DO
```


Clear the workspace and load the boston housing data and extract $X$ and $y$. The dimensions are $n=506$ and $p=13$. Create a matrix that is $(p + 1) \times (p + 1)$ full of NA's. Label the columns the same columns as X. Do not label the rows. For the first row, find the OLS estimate of the $y$ regressed on the first column only and put that in the first entry. For the second row, find the OLS estimates of the $y$ regressed on the first and second columns of $X$ only and put them in the first and second entries. For the third row, find the OLS estimates of the $y$ regressed on the first, second and third columns of $X$ only and put them in the first, second and third entries, etc. For the last row, fill it with the full OLS estimates.

```{r}
#TO-DO
```

Why are the estimates changing from row to row as you add in more predictors?

#TO-DO

Create a vector of length $p+1$ and compute the R^2 values for each of the above models. 

```{r}
#TO-DO
```

Is R^2 monotonically increasing? Why?

#TO-DO
