---
title: "Lab 7"
author: "Your Name Here"
output: pdf_document
date: "11:59PM April 22, 2021"
---

We will get some experience with speeding up R code using C++ via the `Rcpp` package.

First, clear the workspace and load the `Rcpp` package.

```{r}
#TO-DO
```

Create a variable `n` to be 10 and a vaiable `Nvec` to be 100 initially. Create a random vector via `rnorm` `Nvec` times and load it into a `Nvec` x `n` dimensional matrix.

```{r}
#TO-DO
```

Write a function `all_angles` that measures the angle between each of the pairs of vectors. You should measure the vector on a scale of 0 to 180 degrees with negative angles coerced to be positive.

```{r}
#TO-DO
```

Plot the density of these angles.

```{r}
#TO-DO
```

Write an Rcpp function `all_angles_cpp` that does the same thing. Use an IDE if ou want, but write it below in-line.

```{r}
#TO-DO
```

Test the time difference between these functions for `n = 1000` and `Nvec = 100, 500, 1000, 5000`.  Store the results in a matrix.

```{r}
#TO-DO
```

Plot the divergence of performance (in log seconds) over n using a line geometry. Use two different colors for the R and CPP functions. Make sure there's a color legend on your plot.

```{r}
#TO-DO
```

Let `Nvec = 10000` and vary `n` to be 10, 100, 1000. Plot the density of angles for all three values of `n` on one plot using color to signify `n`. Make sure you have a color legend. This is not easy.

```{r}
#TO-DO
```

Write an R function `nth_fibonnaci` that finds the nth Fibonnaci number via recursion but allows you to specify the starting number. For instance, if the sequency started at 1, you get the familiar 1, 1, 2, 3, 5, etc. But if it started at 0.01, you would get 0.01, 0.01, 0.02, 0.03, 0.05, etc.

```{r}
#TO-DO
```

Write an Rcpp function `nth_fibonnaci_cpp` that does the same thing. Use an IDE if ou want, but write it below in-line.

```{r}
#TO-DO
```

Time the difference in these functions for n = 100, 200, ...., 1500 while starting the sequence at the smallest possible floating point value in R. Store the results in a matrix.

```{r}
#TO-DO
```

Plot the divergence of performance (in log seconds) over n using a line geometry. Use two different colors for the R and CPP functions. Make sure there's a color legend on your plot.

```{r}
#TO-DO
```

