---
title: "Lab 9"
author: "Your Name Here"
output: pdf_document
date: "11:59PM May 10, 2021"
---

Here we will learn about trees, bagged trees and random forests. You can use the `YARF` package if it works, otherwise, use the `randomForest` package (the standard).

Let's take a look at the simulated sine curve data from practice lecture 12. Below is the code for the data generating process:

```{r}
rm(list = ls())
n = 500
sigma = 0.3
x_min = 0
x_max = 10
f_x = function(x){sin(x)}
y_x = function(x, sigma){f_x(x) + rnorm(n, 0, sigma)}
x_train = runif(n, x_min, x_max)
y_train = y_x(x_train, sigma)
```

Plot an example dataset of size 500:

```{r}
pacman::p_load(ggplot2)
#TO-DO
```

Create a test set of size 500 as well

```{r}
#TO-DO
```

Locate the optimal node size hyperparameter for the regression tree model. I believe you can use `randomForest` here by setting `ntree = 1`, `replace = FALSE`, `sampsize = n` (`mtry` is already set to be 1 because there is only one feature) and then you can set `nodesize`. Plot nodesize by out of sample s_e. Plot.

```{r}
#TO-DO
```

Plot the regression tree model g(x) with the optimal node size.

```{r}
#TO-DO
```

Provide the bias-variance decomposition of this DGP fit with this model. It is a lot of code, but it is in the practice lectures. If your three numbers don't add up within two significant digits, increase your resolution.

```{r}
#TO-DO
```


```{r}
rm(list = ls())
```

Take a sample of n = 2000 observations from the diamonds data.

```{r}
#TO-DO
```

find the bootstrap s_e for a RF model using 1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000 trees. If you are using the `randomForest` package, you can calculate oob residuals via `e_oob = y_train - rf_mod$predicted`. Plot.

```{r}
#TO-DO
```

Using the diamonds data, find the oob s_e for a bagged-tree model using 1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000 trees. If you are using the `randomForest` package, you can create the bagged tree model via setting an argument within the RF constructor function. Plot.

```{r}
#TO-DO
```


What is the percentage gain / loss in performance of the RF model vs bagged trees model?

```{r}
#TO-DO
```


Plot oob s_e by number of trees for both RF and bagged trees using a long data frame.

```{r}
#TO-DO
```

Build RF models for 500 trees using different `mtry` values: 1, 2, ... the maximum. That maximum will be the number of features assuming that we do not binarize categorical features if you are using `randomForest` or the number of features assuming binarization of the categorical features if you are using `YARF`. Calculate oob s_e for all mtry values. Plot.

```{r}
#TO-DO
```

Plot oob s_e by mtry.

```{r}
#TO-DO
```

```{r}
rm(list = ls())
```


Take a sample of n = 2000 observations from the adult data.

```{r}
#TO-DO
```

Using the adult data, find the bootstrap misclassification error for an RF model using 1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000 trees.

```{r}
#TO-DO
```

Using the adult data, find the bootstrap misclassification error for a bagged-tree model using 1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000 trees. Plot.

```{r}
#TO-DO
```

What is the percentage gain / loss in performance of the RF model vs bagged trees model?

```{r}
#TO-DO
```


Plot bootstrap misclassification error by number of trees for both RF and bagged trees using a long data frame.

```{r}
#TO-DO
```

Build RF models for 500 trees using different `mtry` values: 1, 2, ... the maximum (see above as maximum is defined by the specific RF algorithm implementation). Plot.

```{r}
#TO-DO
```

Plot bootstrap misclassification error by `mtry`.

```{r}
#TO-DO
```


```{r}
rm(list = ls())
```

Write a function `random_bagged_ols` which takes as its arguments `X` and `y` with further arguments `num_ols_models` defaulted to 100 and `mtry` defaulted to NULL which then gets set within the function to be 50% of available features. This argument builds an OLS on a bootstrap sample of the data and uses only `mtry < p` of the available features. The function then returns all the `lm` models as a list with size `num_ols_models`.

```{r}
#TO-DO

```


Load up the Boston Housing Data and separate into `X` and `y`.

```{r}
#TO-DO
```


Similar to lab 1, write a function that takes a matrix and punches holes (i.e. sets entries equal to `NA`) randomly with an argument `prob_missing`.

```{r}
#TO-DO
```

Create a matrix `Xmiss` which is `X` but has missingness with probability of 10%.

```{r}
#TO-DO
```

Use a random forest modeling procedure to iteratively fill in the `NA`'s by predicting each feature of X using every other feature of X. You need to start by filling in the holes to use RF. So fill them in with the average of the feature.

```{r}
Ximps = list()


t = 1
repeat {
  for (j in 1 : p){
    Ximps[[t]][, j] = randomForest(X = Ximps[[j - 1]] %>% select(-j), y = Ximps[[j - 1]][, j])
  }
  t = t + 1
  #stop condition if Ximps[[t]] - Ximps[[t - 1]] is close together
  if (stop){
    break
  }
}
```
