---
title: "Lab 4"
author: "Your Name Here"
output: pdf_document
date: "11:59PM Feb 29, 2020"
---

Load up the famous iris dataset. We are going to do a different prediction problem. Imagine the only input x is Species and you are trying to predict y which is Petal.Length. What would a reasonable, naive prediction be under all Species? Hint: it's what we did in class.

```{r}
#TO-DO
```

Prove that this is the OLS model by fitting an appropriate `lm` and then using the predict function to verify you get the same answers as you wrote previously. Show this by doing a linear regression with and without the intercept.

```{r}
#TO-DO
```

Use the `model.matrix` function to compute the matrix `X` for the regression with the intercept and without the intercept. What is different?

```{r}
#TO-DO
```





Find the hat matrix $H$.

```{r}
H = X %*% XtXinvX
H
```

Verify that this specific hat matrix is symmetric.

```{r}
expect_equal(H, t(H))

```


Using the `diag` function, find the trace of the hat matrix.

```{r}
diag(H)
sum(diag(H))
```







