---
title: "Lab 5"
author: "Your Name Here"
output: pdf_document
date: "11:59PM March 7, 2020"
---

Load the Boston housing data frame and create the vector $y$ (the median value) and matrix $X$ (all other features) from the data frame. Name the columns the same as Boston except for the first name it "(Intercept)".

```{r}
y = MASS::Boston$medv
X = MASS::Boston[ , 1: 13] 

```

Run the OLS linear model to get $b$, the vector of coefficients. Do not use `lm`.

```{r}
X = cbind(1,as.matrix(X))
b = solve(t(X) %*% X ) %*% t(X) %*% y

```

Find the hat matrix for this regression `H`. Verify its dimension is correct and verify its rank is correct.

```{r}
H = X %*% solve(t(X) %*% X ) %*% t(X)
dim(H)
pacman::p_load(Matrix)
rankMatrix(H)
```

Verify this is a projection matrix by verifying the two sufficient conditions. Use the `testthat` library's `expect_equal(matrix1, matrix2, tolerance = 1e-2)`.

```{r}
pacman::p_load(testthat)
expect_equal(H, t(H))
expect_equal(H, H %*% H)

```

Find the matrix that projects onto the space of residuals `Hcomp` and find its rank. Is this rank expected?

```{r}
Hcomp = diag(nrow(X)) - H
rankMatrix(H)
rankMatrix(Hcomp, tol = 1e-2)
```

Verify this is a projection matrix by verifying the two sufficient conditions. Use the `testthat` library.

```{r}
expect_equal(Hcomp, t(Hcomp))
expect_equal(Hcomp, Hcomp %*% Hcomp)
```

Use `diag` to find the trace of both `H` and `Hcomp`.

```{r}
sum(diag(H))
sum(diag(Hcomp))
```

Do you have a conjecture about the trace of an orthogonal projection matrix?
 
trace is equal to the rank

Find the eigendecomposition of both `H` and `Hcomp` as `eigenvals_H`, `eigenvecs_H`, `eigenvals_Hcomp`, `eigenvecs_Hcomp`. Verify these results are the correct dimensions.

```{r}
eigen_H = eigen(H)
eigen_Hcomp = eigen(Hcomp)

eigenvals_H = eigen_H$values
eigenvecs_H = eigen_H$vectors
eigenvals_Hcomp = eigen_Hcomp$values
eigenvecs_Hcomp = eigen_Hcomp$vectors

length(eigenvals_H)
dim(eigenvecs_H)
length(eigenvals_Hcomp)
dim(eigenvecs_Hcomp)
```

The eigendecomposition suffers from numerical error which is making them become imaginary. We can coerce imaginary numbers back to real by using the `Re` function. There is also lots of numerical error. Use the `Re` function to coerce to real and the `round` function to round all four objects to the nearest 10 digits.

```{r, warning = FALSE, message = FALSE}
eigenvals_H = round(as.numeric(eigenvals_H), 10)
eigenvecs_H = round(Re(eigenvecs_H), 10)
eigenvals_Hcomp = round(as.numeric(eigenvals_Hcomp), 10)
eigenvecs_Hcomp = round(Re(eigenvecs_Hcomp), 10)
```

Print out the eigenvalues of both `H` and `Hcomp`. Is this expected?

```{r}
eigenvals_H
eigenvecs_H
eigenvals_Hcomp
```

Find the length of all eigenvectors of `H` in one line. 

```{r}
apply(eigenvecs_H, MARGIN =2, FUN = function(v){
  sqrt(sum(v^2))
})
```

Is this expected? What is the convention for eigenvectors in R's `eigen` function?

Yes. The convention is length 1.

The first p+1 eigenvectors are the columns of $X$ but they are in arbitrary order. Find the column that represents the one-vector. 

```{r}
head(eigenvecs_H[, 3])
```

Why is it not exactly 506 1's?

Numeric error

Use the first p+1 eigenvectors as a model matrix and run the OLS model of medv on that model matrix. 


```{r}
mod1 = lm(y ~ X)
mod2 = lm(y ~ eigenvecs_H[, 1:14])
summary(mod1)
summary(mod2)
```

Is b about the same above (in arbitrary order)?

NO, the eigen vectors are scaled to be unit length

Calculate $\hat{y}$ using the hat matrix.

```{r}
y_hat= H %*% y
y_hat
```

Calculate $e$ two ways: (1) the difference of $y$ and $\hat{y}$ and (2) the projection onto the space of the residuals. Verify the two means of calculating the residuals provide the same results via `expect_equal`.

```{r}
e1 = y -y_hat
e2 = Hcomp %*% y
expect_equal(e1, e2)
```

Calculate $R^2$ using the angle relationship between the responses and their predictions.

```{r}

length_of_vec = function(v){sqrt(sum(v^2))}
y_avg_adj = y - mean(y)
y_yhat_adj = y_hat - mean(y)
(sum(y_avg_adj * y_yhat_adj) / (length_of_vec(y_avg_adj) * length_of_vec(y_yhat_adj))) ** 2

```

Find the cosine-squared of $y - \bar{y}$ and $\hat{y} - \bar{y}$ and verify it is the same as $R^2$.

```{r}
summary(mod1)$r.squared
```

Verify $\hat{y}$ and $e$ are orthogonal.

```{r}
sum(y_hat*e1)

```

Verify $\hat{y} - \bar{y}$ and $e$ are orthogonal.

```{r}
sum((y_hat -mean(y)) *e1)
```

Verify the sum of squares identity which we learned was due to the Pythagorean Theorem (applies since the projection is specifically orthogonal). You need to compute all three quantities first: SST, SSR and SSE.

```{r}
#TO-DO
```

Create a matrix that is $(p + 1) \times (p + 1)$ full of NA's. Label the columns the same columns as X. Do not label the rows. For the first row, find the OLS estimate of the $y$ regressed on the first column only and put that in the first entry. For the second row, find the OLS estimates of the $y$ regressed on the first and second columns of $X$ only and put them in the first and second entries. For the third row, find the OLS estimates of the $y$ regressed on the first, second and third columns of $X$ only and put them in the first, second and third entries, etc. For the last row, fill it with the full OLS estimates.

```{r}
#TO-DO
```

Examine this matrix. Why are the estimates changing from row to row as you add in more predictors?

#TO-DO

Clear the workspace and load the diamonds dataset in the package `ggplot2`.

```{r}
#TODO
```

Extract $y$, the price variable and `col`, the nominal variable "color" as vectors.

```{r}
#TO-DO
```

Convert the `col` vector to $X$ which contains an intercept and an appropriate number of dummies. Let the color G be the refernce category as it is the modal color. Name the columns of $X$ appropriately. The first should be "(Intercept)". Delete `col`.

```{r}
#TO-DO
```

Repeat the iterative exercise above we did for Boston here.

```{r}
#TO-DO
```

Why didn't the estimates change as we added more and more features?

#TO-DO

Model `price` with both `color` and `clarity` with and without an intercept and report the coefficients.

```{r}
#TO-DO
```

Which coefficients did not change between the models and why?

#TO-DO



Create a 2x2 matrix with the first column 1's and the next column iid normals. Find the absolute value of the angle (in degrees, not radians) between the two columns.

```{r}
#TO-DO
```

Repeat this exercise $Nsim = 1e5$ times and report the average absolute angle.

```{r}
#TO-DO
```

Create a 2xn matrix with the first column 1's and the next column iid normals. Find the absolute value of the angle (in degrees, not radians) between the two columns. For $n \in {10, 50, 100, 200, 500, 1000}$, report the average absolute angle over $Nsim = 1e5$ simulations.

```{r}
#TO-DO
```

What is this absolute angle converging to? Why does this make sense?

#TO-DO

Create a vector $y$ by simulating $n = 100$ standard iid normals. Create a matrix of size 100 x 2 and populate the first column by all ones (for the intercept) and the second column by 100 standard iid normals. Find the $R^2$ of an OLS regression of `y ~ X`. Use matrix algebra.

```{r}
#TO-DO
```

Write a for loop to each time bind a new column of 100 standard iid normals to the matrix $X$ and find the $R^2$ each time until the number of columns is 100. Create a vector to save all $R^2$'s. What happened??

```{r}
#TO-DO
```

Add one final column to $X$ to bring the number of columns to 101. Then try to compute $R^2$. What happens?

```{r}
#TO-DO
```

